---
title: "cytokine_analysis"
author: "Laurie Lyon"
date: "2025-07-08"
output: html_document
---

```{r}
#Replace with your FP to this folder
#setwd("~/CRS3123_omics_manuscript/analysis/4_fecal_immune/scripts")
```

```{r}
library(readxl)
library(eatTools) #for rbind_common
library(tidyverse)
library(broom)
library(magrittr)
source("../../art_lmer_functions.R")
source("../../diff_between_means_calc.R")
library(ggpubr)
#library(apppleplots)
library(lme4)
library(lmerTest)
library(lmtest)
```

```{r}
BCA_total_protein <- read_excel("../data/BCA_total_protein.xlsx")

cytokine_excel_fp <- "../data/cytokine_panel_1.xlsx"
proinflamm_cytokine_excel_fp <- "../data/proinflamm_cytokine_panel_1.xlsx"

cp_lf_results <- read_csv("../results/processed_cp_lf_data.csv")

metadata <- read_tsv(file = "../data/proc_seq_metadata.tsv")
clin_metadata <- read_tsv(file = "../data/clinical_metadata.txt")
subj_info <- read_tsv(file = "../data/subject_info.tsv")

ULOD <- read_excel("../data/ULOD_cytokines.xlsx")
plates <- read_excel("../data/cytokine_subj_plates.xlsx")
LLOD <- read_excel("../data/LLOD_cytokines.xlsx")
```

## reading in cytokine raw data excel sheet and making it into a code friendly tibble/dataframe :))
```{r}
## pulling sheet names off of excel workbook so we can read them all in 
cytokine_sheet_names <- excel_sheets(cytokine_excel_fp)

## empty tibble for outputs
cytokine_dfs <- tibble()

for (sheet_name in cytokine_sheet_names) {
  tmp_df <- read_excel(cytokine_excel_fp,
                       sheet = sheet_name, 
                       col_names = TRUE)
  ## use rbind to catch all outputs and smush them on top of each other (long format)
  cytokine_dfs <- rbind(cytokine_dfs, tmp_df)
}

## checking that all cytokine sheets were read in properly 
unique(cytokine_dfs$immune_product)


```
```{r}
## pulling sheet names off of excel workbook so we can read them all in 
proinflamm_cytokine_sheet_names <- excel_sheets(proinflamm_cytokine_excel_fp)

## empty tibble for outputs
proinflamm_cytokine_dfs <- tibble()

for (sheet_name in proinflamm_cytokine_sheet_names) {
  tmp_df <- read_excel(proinflamm_cytokine_excel_fp,
                       sheet = sheet_name, 
                       col_names = TRUE)
  ## use rbind to catch all outputs and smush them on top of each other (long format)
  proinflamm_cytokine_dfs <- rbind(proinflamm_cytokine_dfs, tmp_df)
}

## checking that all cytokine sheets were read in properly 
unique(proinflamm_cytokine_dfs$immune_product)
```

Merge with metadata

```{r}
(meta_trim <- metadata %>% 
  dplyr::select(sampleid, subject_id, visit_des, new_visit_label, new_visit_num) %>% 
  mutate(visit = gsub(" Cooled", "", visit_des),
         #one line if else: ifelse(if this, then this, else this)
         visit = ifelse(visit == "TOC/FUV1", "FUV1", visit))
  ) 

merged_cyto <- meta_trim %>% 
  left_join(cytokine_dfs, by = c("subject_id", "visit"))

merged_proinflamm <- meta_trim %>% 
  left_join(proinflamm_cytokine_dfs, by = c("subject_id", "visit"))

#  
all_cyto <- rbind(merged_cyto, merged_proinflamm)
all_cyto <- left_join(all_cyto, subj_info, by = c("subject_id"))

all_cyto <- all_cyto %>% 
  left_join(plates, by = c("subject_id")) %>% 
  left_join(LLOD, by = c("plate", "immune_product")) %>% 
  left_join(ULOD, by = c("subject_id", "visit", "immune_product")) %>% 
  left_join(BCA_total_protein, by = c("subject_id", "visit"))

```


```{r}
min(all_cyto$calc_concentration_mean, na.rm = TRUE)

#all_immune_markers$calc_concentration_mean <- type.convert(sub("NaN", 0, as.matrix(all_immune_markers$calc_concentration_mean))

#chose LOD/2
all_cytokines_filtered <- all_cyto %>% 
  filter(!is.na(calc_concentration_mean)) %>% 
  mutate(calc_concentration_mean = as.numeric(calc_concentration_mean),
    calc_concentration_mean = case_when(
    calc_concentration_mean == "NaN" ~ Min/2, #using LOD/2 for below LOD (curve minimum assigned value)
    Note.x == "Value is above the range"| Note.x == "Value is above the range." ~ Max, #default to max assignable value
    .default = calc_concentration_mean
    )) 
  # mutate(total_protein_ug_mL = ifelse(total_protein_ug_mL == "NA", NA, as.numeric(total_protein_ug_mL))) %>%
  # filter(!is.na(total_protein_ug_mL)) %>% 
  # #group_by(sampleid) %>% 
  # mutate(norm_conc_BCA = calc_concentration_mean/(total_protein_ug_mL * 1e6))

all_immune_markers_filtered <- rbind_common(all_cytokines_filtered, cp_lf_results)

class(all_immune_markers_filtered$calc_concentration_mean)

write_csv(all_immune_markers_filtered, "../results/merged_censored_immune_markers.csv")

```
## Going back now and looking at distributions of data 
- histograms/density plots to look at distributions
- look for normality skewness, bimodality or multiple peaks, spikes at LOD values 

```{r, fig.width=10, fig.height=5}

(data_distribution <- all_immune_markers_filtered %>% 
  ggplot(aes(x = calc_concentration_mean)) +
    geom_histogram(aes(y=..density..), bins=30, fill="lightblue", color="black") +
    geom_density(color="red", linewidth=1) +
    facet_wrap(~immune_product, scales="free") +
    theme_bw())

```
- these results show that the data is non-normal and has a strong right-tailed (positive) skew

### looking at censoring (LOD issues)
- see if a significant proportion of data are below or above assay limits of detection 
- Note from searching online: If censoring is substantial (>10–20%), consider methods that explicitly handle censored data, such as Tobit regression or imputation methods (LOD/2, etc.).
- for >20% censoring (10/20 cytokines) mixed-effects Tobit models can be used to model each cytokine longitudinally while accounting for censoring and within subjct correlations (look into censReg, survival::survreg, or glmmTMB packages)
- regression on order-statistics (ROS) can also be used (in NADA package) - fits log-normal or other distribution ot uncensored data then imputes plausible values for censored ones (I don't really get how though)
```{r}
#not sure why this is coming up as all 0's
LOD_percentages <- all_cyto %>% 
  group_by(immune_product) %>% 
  summarise(
    total_n = n(), 
    num_below_LOD = sum(calc_concentration_mean == Min/2, na.rm = TRUE),
    num_above_LOD = sum(calc_concentration_mean == Max, na.rm = TRUE), 
    percentage_below_LOD = (num_below_LOD / total_n) * 100, 
    percentage_above_LOD = (num_above_LOD / total_n) * 100 
  )
```

## Factoring values for lmer models
```{r}

all_immune_markers_factor <- all_immune_markers_filtered %>%
  convert_as_factor(subject_id, treatment_group, new_visit_num) %>%
  rename("calc_concentration" = "calc_concentration_mean")
```

## Trying lme4 for linear models with anova to compare models 

```{r}
immune_products <- unique(all_immune_markers_factor$immune_product)

model_list <- vector("list", length(immune_products))
anova_list <- vector("list", length(immune_products))

for (i in seq_along(immune_products)) {
  prod <- immune_products[i]
  
  tbl <- all_immune_markers_factor %>% 
    filter(immune_product == prod)
  
  model_2 <- lmer(
    calc_concentration ~ treatment_group * visit + (1|subject_id), 
    data = tbl, 
    REML = FALSE
  )
  
  anova_1 <- anova(model_2, type = "III")
  
  model_list[[i]] <- model_2
  anova_list[[i]] <- as.data.frame(anova_1) %>%
    rownames_to_column(var = "Term") %>%
    mutate(immune_product = prod)
}

# 4) Combine ANOVA tables into one data.frame
all_anova_2 <- bind_rows(anova_list)

all_summaries_2 <- purrr::map2_df(model_list, immune_products, function(mod, prod) {
  coef_df <- as.data.frame(summary(mod)$coefficients) %>%
    tibble::rownames_to_column("term") %>%
    rename(
      estimate  = `Estimate`,
      std.error = `Std. Error`,
      df        = `df`,
      statistic = `t value`,
      p.value   = `Pr(>|t|)`
    ) %>%
    mutate(immune_product = prod)
})

write.csv(all_summaries_2, "../data/cytokine_treatment_visit_lme4_model_summaries_rawdata.csv", row.names = FALSE)
```
- for all_summaries, intercept() will be the mean transformed concentration at the reference level (e.g. CRS_200mg at D1)
- i honestly don't know how to interpret these SOS 

```{r}
# significant_transformed_visit_treatment_ints <- all_summaries %>% 
#   filter(p.value <= 0.05)

significant_raw_visit_treatment_ints <- all_summaries_2 %>% 
  filter(p.value <= 0.05)
```

## Aligned Rank Transformed Linear Mixed Effect Models
- I want to repeat what we did in cp_lf_plotting by: 
  - vectorizing the variables we need
  - writing a for loop that will perform the art_lmer_strat function on all the cytokines
  - add a column for immune_product in the results and contrast outputs 
  - rbind all results and contrasts together (can distinguish based on above column)
  
```{r}

immune_products <- unique(all_immune_markers_factor$immune_product)

results_list <- vector("list", length(immune_products))
contrast_list <- vector("list", length(immune_products))

for (i in seq_along(immune_products)) {
  prod <- immune_products[i]
  
  tbl <- all_immune_markers_factor %>% 
    filter(immune_product == prod)
  
  art_out <- art_lmer_strat(
    input_table = tbl, 
    strat_variable = "new_visit_num", 
    model_formula_left = "calc_concentration",
    model_formula_right = "treatment_group",
    model_contrast_run_by = "treatment_group"
  )
  
  results_list[[i]] <- art_out$ARTResults %>% 
    mutate(immune_product = prod)
  contrast_list[[i]] <- art_out$ARTContrast %>% 
    mutate(immune_product = prod)
}

# bind them all together
visit_strat_results <- bind_rows(results_list)
visit_strat_contrast <- bind_rows(contrast_list)
```

would want to visualize contrast results as a heat map, but would want to calculate difference in the means to determine directionality 
- use Madi's apppleplot package
- need to put the contrast output into a specific format before running the heat map script 

```{r, fig.width=8, fig.height=3}
visit_strat_diff_means <- calc_diff_means(input_table = all_immune_markers_factor,
                first_group = "treatment_group",
                second_group = c("new_visit_num", "immune_product"),
                mean_value = "calc_concentration",
                post_hoc_test = visit_strat_contrast) %>% 
  filter(signif != 'ns')

visit_nums <- c('Screen',
                  'D1',
                  'D4',
                  'TOC',
                  'D17',
                  'D40',
                  'D70')
names(visit_nums) <- c(1, 2, 3, 4, 5, 6, 7)
 

(visit_strat_heat_plot <- stat_heat_plot(input_table = visit_strat_diff_means, 
                                        x_value = 'group1',
                                        y_value = 'group2',
                                        x_value_type = 'discrete',
                                        y_value_type = 'discrete',
                                        pairwise = TRUE,
                                        correlation = FALSE,
                                        tile_fill = 'stat_diff_means',
                                        text_value = 'signif',
                                        legend_name = 'Group 1 -\n Group 2',
                                        x_labels = c('CRS3123\n200mg',
                                                     'CRS3123\n400mg'),
                                        y_labels = c('CRS3123 400mg', 
                                                     'Vancomycin 125mg'),
                                        facet_by = 'new_visit_num',
                                        facet_labels = visit_nums,
                                        x_name = 'Group 1',
                                        y_name = 'Group 2',
                                        title_content = 'Pairwise Comparisons Between Treatment Groups') +
  facet_grid(rows = vars(immune_product), cols = vars(new_visit_num), labeller = labeller(.cols = visit_nums)) +
  labs(subtitle = "Visit"))
        
```
interpretation of AML-lmer results: 
- IL-10 is significantly higher at D1 in vanc vs. CRS 400mg (is this baseline, before or day of start of Abx?)
- IL-12p40 is higher in CRS 400mg than 200mg at D4 (subunit of IL-12 and IL-23?)
- IL-15 is higher in 400mg than 200mg at TOC and D40 (inducer of T and NK cells?)
- IL-1b is higher in vanc than CRS 200mg at D4 (fecal cp was correlated with IL-1b in the HIV fecal sIF paper)
- results were same whether I used transformed calc_concentration_means or not 

```{r}
ggsave("../plots/treatment_group_pairwise_heatmap.png", width = 8, height = 3)
```

ART-lmer where we stratify by treatment group and look at effect of day on calc_concentration_mean when accounting for random effects by subject
```{r}

immune_products <- unique(all_immune_markers_factor$immune_product)

results_list <- vector("list", length(immune_products))
contrast_list <- vector("list", length(immune_products))

for (i in seq_along(immune_products)) {
  prod <- immune_products[i]
  
  tbl <- all_immune_markers_factor %>% 
    filter(immune_product == prod)
  
  art_out <- art_lmer_strat(
    input_table = tbl, 
    strat_variable = "treatment_group", 
    model_formula_left = "calc_concentration",
    model_formula_right = "new_visit_num + (1|subject_id)",
    model_contrast_run_by = "new_visit_num"
  )
  
  results_list[[i]] <- art_out$ARTResults %>% 
    mutate(immune_product = prod)
  contrast_list[[i]] <- art_out$ARTContrast %>% 
    mutate(immune_product = prod)
}

# bind them all together
treatment_strat_raw_results <- bind_rows(results_list) %>% 
  rename(p.adj.signif = signif) 
treatment_strat_raw_contrast <- bind_rows(contrast_list) %>% 
  rename(p.adj.signif = signif) %>% 
  mutate(y.position = case_when(
    treatment_group == 'CRS3123_200mg_po_bid' ~ 1,
    treatment_group == 'CRS3123_400mg_po_bid' ~ 1,
    treatment_group == 'Vancomycin_125mg_po_qid' ~ 1), 
    ## pulls the 'new_visit_num' off of the values in these cols so it matches the plot below 
    group1 = gsub("new_visit_num*", "", group1),
    group2 = gsub("new_visit_num*", "", group2)) 

treatment_strat_significant_contrast <- treatment_strat_raw_contrast %>% 
  filter(p.adj.signif != 'ns')
```


```{r, fig.width = 10, fig.height = 20}
## renaming facet labels for treatment group and immune product
## for treatment group
treat_group_labs <- c('CRS3123 200mg',
                      'CRS3123 400mg',
                      'Vancomycin 125mg')
names(treat_group_labs) <- c('CRS3123_200mg_po_bid',
                             'CRS3123_400mg_po_bid',
                             'Vancomycin_125mg_po_qid')

## for immune product
immune_prod_labs <- c('Calprotectin',
                      'Lactotransferrin')
names(immune_prod_labs) <- c('calprotectin',
                             'lactotransferrin')

## to rename the x-axis labels (for visit)
## have to add breaks (aka tick marks) on the x axis so all visits can be labeled properly
x_axis_breaks <- c(1, 2, 3, 4, 5, 6, 7)
## the visit labels in order 
visit_x_labs <- c('Screen',
                  'D1',
                  'D4',
                  'TOC',
                  'D17',
                  'D40',
                  'D70')

cytokine_panel_1 <- all_immune_markers_factor %>% 
  filter(immune_product == unique(cytokine_dfs$immune_product))

proinflamm_panel_1 <- all_immune_markers_factor %>% 
  filter(immune_product == unique(proinflamm_cytokine_dfs$immune_product))

proinflamm_panel_1 %>% 
  #filter(immune_product == "") %>% 
  mutate(new_visit_num = as.numeric(new_visit_num)) %>% 
  ggplot(aes(x = new_visit_num, y = calc_concentration)) + 
  geom_boxplot(aes(group = new_visit_num), outlier.shape = NA) +
  geom_smooth(se = FALSE) +
  #geom_point(aes(fill = subject_id), size = 3, pch = 21) +
  geom_point(size = 2, alpha = 0.4) + ## non-colored point plot
  geom_line(aes(group = subject_id), alpha = 0.2) + 
  facet_grid(immune_product ~ treatment_group,
             scales = "free_y",
             labeller = labeller(treatment_group = treat_group_labs,
                                 immune_product = immune_prod_labs)) +
  scale_y_continuous(trans = "log10") +
  scale_x_continuous(breaks = x_axis_breaks, labels = visit_x_labs) +
  # stat_pvalue_manual(treatment_strat_contrast,
  #                    step.group.by = "treatment_group",
  #                    step.increase = 0.1,
  #                    hide.ns = TRUE,
  #                    label = 'p.adj.signif',
  #                    y.position = 1) +
  theme_bw(base_size = 14) +
  theme(strip.text.y.right = element_text(angle = 0)) +
  labs(x = 'Visit',
       y = 'Mean Concentration (pg/mL)',
       title = 'Immune Markers in Patients\' Fecal Water Over Time', 
       subtitle = 'Proinflammatory Panel 1')
```
```{r}
ggsave(filename = "../plots/proinflamm_cytokines_over_time_treatment.png", height = 20, width = 10)
```


```{r, fig.width = 12, fig.height = 15}
cytokine_panel_1 %>% 
  mutate(new_visit_num = as.numeric(new_visit_num)) %>%
  ggplot(aes(x = new_visit_num, y = calc_concentration)) + 
  geom_boxplot(aes(group = new_visit_num), outlier.shape = NA) +
  geom_smooth(se = FALSE) +
  #geom_point(aes(fill = subject_id), size = 3, pch = 21) +
  geom_point(size = 2, alpha = 0.4) + ## non-colored point plot
  geom_line(aes(group = subject_id), alpha = 0.2) + 
  facet_grid(immune_product ~ treatment_group,
             scales = "free_y",
             labeller = labeller(treatment_group = treat_group_labs,
                                 immune_product = immune_prod_labs)) +
  scale_y_continuous(trans = "log10") +
  #scale_x_discrete(labels = visit_x_labs) +
  scale_x_continuous(breaks = x_axis_breaks, labels = visit_x_labs) +
  # stat_pvalue_manual(treatment_strat_contrast,
  #                    step.group.by = "treatment_group",
  #                    step.increase = 0.1,
  #                    hide.ns = TRUE,
  #                    label = 'p.adj.signif',
  #                    y.position = 1) +
  theme_bw(base_size = 16) +
  theme(strip.text.y.right = element_text(angle = 0)) +
  labs(x = 'Visit',
       y = 'Mean Concentration (pg/mL)',
       title = 'Immune Markers in Patients\' Fecal Water Over Time', 
       subtitle = 'Cytokine Panel 1')
```

```{r}
ggsave("../plots/cytokines_over_time_plot.png", width = 10, height = 20)
```




