---
title: "cytokine_analysis"
author: "Laurie Lyon"
date: "2025-07-08"
output: html_document
---
Load in packages and functions
```{r}
library(tidyverse)
library(broom)
library(magrittr)
source("../../art_lmer_functions.R")
source("../../diff_between_means_calc.R")
library(ggpubr)
#library(apppleplots)
library(lme4)
```

Load in data (generated by immune_data_processing.Rmd)
```{r}
imm_df <- read_csv("../results/merged_censored_immune_markers.csv")
```

Factoring values for lmer models
And renaming calc_concentration_mean because Madi's diff_between_means_calc function will get mad
```{r}
imm_df <- imm_df %>%
  convert_as_factor(subject_id, treatment_group, new_visit_num) %>% 
  mutate(
    new_visit_label = factor(
      new_visit_label,
      levels = c("Screen","D1","D4","D10-15/TOC","D17","D40","D70")
    )
  ) %>% 
  rename("calc_concentration" = "calc_concentration_mean")
```
dfs for every immune_product
```{r}
il1b_df <- imm_df %>% 
  filter(immune_product == "IL-1b")

cp_df <- imm_df %>% 
  filter(immune_product == "calprotectin")

```

**Trial Stats with IL-1b**
# start with general ART model (notStrat)
- going to just do IL-1b as an example
```{r}
il1b_notstrat_mod <- art_lmer_notStrat(
  input_table = il1b_df, 
  formula_left = "calc_concentration", 
  formula_right = "treatment_group * new_visit_label + (1|subject_id)",
  contrast_run_by = "treatment_group"
)

il1b_notstrat_results <- il1b_notstrat_mod$ARTResults
il1b_notstrat_contrast <- il1b_notstrat_mod$ARTContrast
```
- results: IL1b is influenced by both treatment, visit, and interaction between the two
- contrast results are really long and it doesn't make sense to compare 200mg at D1 to say vanc on D4 -> stratify

# ART stratified by timepoint
- effect of treatment on IL-1b at each timepoint
```{r}
il1b_strat_mod <- art_lmer_strat(
  input_table = il1b_df, 
  strat_variable = "new_visit_label",
  model_formula_left = "calc_concentration",
  model_formula_right = "treatment_group", #since each subj only has 1 conc at each timepoint, removed random effect
  model_contrast_run_by = "treatment_group"
)

il1b_strat_results <- il1b_strat_mod$ARTResults
il1b_strat_contrast <- il1b_strat_mod$ARTContrast
```

**Visit strat on all immune product ~ treatment_group**
  
```{r}

immune_products <- unique(imm_df$immune_product)

results_list <- vector("list", length(immune_products))
contrast_list <- vector("list", length(immune_products))

for (i in seq_along(immune_products)) {
  prod <- immune_products[i]
  
  tbl <-imm_df %>% 
    filter(immune_product == prod)
  
  art_out <- art_lmer_strat(
    input_table = tbl, 
    strat_variable = "new_visit_label", 
    model_formula_left = "calc_concentration",
    model_formula_right = "treatment_group",
    model_contrast_run_by = "treatment_group"
  )
  
  results_list[[i]] <- art_out$ARTResults %>% 
    mutate(immune_product = prod)
  contrast_list[[i]] <- art_out$ARTContrast %>% 
    mutate(immune_product = prod)
}

# bind them all together
visit_strat_results <- bind_rows(results_list)
visit_strat_contrast <- bind_rows(contrast_list)
```

would want to visualize contrast results as a heat map, but would want to calculate difference in the means to determine directionality 
- use Madi's apppleplot package
- need to put the contrast output into a specific format before running the heat map script 

```{r, fig.width=8, fig.height=3}
visit_strat_diff_means <- calc_diff_means(input_table = imm_df,
                first_group = "treatment_group",
                second_group = c("new_visit_label", "immune_product"),
                mean_value = "calc_concentration",
                post_hoc_test = visit_strat_contrast) %>% 
  filter(signif != 'ns') %>% 
  #adding a log2fc column to get concentration differences on a similar scale
  mutate(log2fc = log2(group1_calc_concentration / group2_calc_concentration))

#since IL-1b and calprotectin are the only ones with differences between vanc and a CRS3123 group
#I want to make a df with only these for the main text figure, using all diff_means for supp. figure
visit_strat_ilib_cp <- visit_strat_diff_means %>% 
  filter(immune_product %in% c("calprotectin", "IL-1b"))

visit_nums <- c('Screen',
                  'D1',
                  'D4',
                  'TOC',
                  'D17',
                  'D40',
                  'D70')
names(visit_nums) <- c(1, 2, 3, 4, 5, 6, 7)
 

(visit_strat_heat_plot <- stat_heat_plot(input_table = visit_strat_diff_means, 
                                        x_value = 'group1',
                                        y_value = 'group2',
                                        x_value_type = 'discrete',
                                        y_value_type = 'discrete',
                                        pairwise = TRUE,
                                        correlation = FALSE,
                                        #tile_fill = 'stat_diff_means',
                                        tile_fill = 'log2fc',
                                        text_value = 'signif',
                                        legend_name = 'Log2FC\nof Group 1\nvs.Group 2',
                                        x_labels = c('CRS3123\n200mg',
                                                     'CRS3123\n400mg'),
                                        y_labels = c('CRS3123 400mg', 
                                                     'Vancomycin 125mg'),
                                        facet_by = 'new_visit_label',
                                        #facet_labels = 'new_visit_label',
                                        x_name = 'Group 1',
                                        y_name = 'Group 2',
                                        title_content = 'Pairwise Comparisons Between Treatment Groups') +
  facet_grid(rows = vars(immune_product), cols = vars(new_visit_label)) +
  labs(subtitle = "Visit")) +
  scale_fill_gradient2(low="blue", high = "red", mid = "white", midpoint = 0)
  #scale_fill_gradient(colors = rainbow(8))
```
interpretation of AML-lmer results: 
- generally
  - pos log2FC ---> Group 1 > Group 2
  - neg log2FC ---> Group 1 < Group 2
  - e.g. +2 = group 1 has 4x higher imm prod than group 2, -1 group 1 has 2x lower immune prod than group 2
- specific results, >0 = group 1 has more than group 2
  - calprotectin is signnificantly higher in vanc vs. 400mg at D4
  - IL-10 is significantly higher at D1 in vanc vs. CRS 400mg (is this baseline, before or day of start of Abx?)
  - IL-12p40 is higher in CRS 400mg than 200mg at D4 (subunit of IL-12 and IL-23?)
  - IL-15 is higher in 400mg than 200mg at TOC and D40 (inducer of T and NK cells?)
  - IL-1b is higher in vanc than CRS 200mg at D4 (fecal cp was correlated with IL-1b in the HIV fecal sIF paper)

```{r}
ggsave("../plots/treatment_group_pairwise_heatmap.png", width = 8, height = 3)
save(visit_strat_heat_plot, file = "../../../figures/ggplots/figS9b_imm_stats.rdat")
```

```{r, fig.width=6, fig.height=2.5}
(visit_strat_cp_il1b_heat_plot <- stat_heat_plot(input_table = visit_strat_ilib_cp, 
                                        x_value = 'group1',
                                        y_value = 'group2',
                                        x_value_type = 'discrete',
                                        y_value_type = 'discrete',
                                        pairwise = TRUE,
                                        correlation = FALSE,
                                        #tile_fill = 'stat_diff_means',
                                        tile_fill = 'log2fc',
                                        text_value = 'signif',
                                        legend_name = 'Log2FC\nof Group 1\nvs.Group 2',
                                        x_labels = c('CRS3123\n200mg',
                                                     'CRS3123\n400mg'),
                                        y_labels = c('Vancomycin 125mg', 
                                                     'Vancomycin 125mg'),
                                        facet_by = 'new_visit_label',
                                        #facet_labels = 'new_visit_label',
                                        x_name = 'Group 1',
                                        y_name = 'Group 2',
                                        title_content = 'Pairwise Comparisons Between Treatment Groups') +
  theme(legend.title = element_text(size = 10)) +
  facet_grid(rows = vars(immune_product), cols = vars(new_visit_label)) +
  labs(subtitle = "Visit", immune_products = immune_prod_labs)) +
  scale_fill_gradient2(low="blue", high = "red", mid = "white", midpoint = 0)
```

```{r}
ggsave("../plots/treatment_group_il1b_cp_pairwise_heatmap.png", width = 6, height = 2.5)
save(visit_strat_cp_il1b_heat_plot, file = "../../figures/ggplots/fig7_cp_il1b_stats.rdat")
```

interpretation of AML-lmer results: 
- generally
  - pos log2FC ---> Group 1 > Group 2
  - neg log2FC ---> Group 1 < Group 2
  - e.g. +2 = group 1 has 4x higher imm prod than group 2, -1 group 1 has 2x lower immune prod than group 2
- specific results, >0 = group 1 has more than group 2
  - calprotectin is signnificantly higher in vanc vs. 400mg at D4
  - IL-10 is significantly higher at D1 in vanc vs. CRS 400mg (is this baseline, before or day of start of Abx?)
  - IL-12p40 is higher in CRS 400mg than 200mg at D4 (subunit of IL-12 and IL-23?)
  - IL-15 is higher in 400mg than 200mg at TOC and D40 (inducer of T and NK cells?)
  - IL-1b is higher in vanc than CRS 200mg at D4 (fecal cp was correlated with IL-1b in the HIV fecal sIF paper)
- results were same whether I used box-cox transformed calc_concentration_means or not 

```{r}
ggsave("../plots/treatment_group_pairwise_heatmap.png", width = 8, height = 3)
```

*Treatment strat on all immune_product ~ visit*
```{r}
immune_products <- unique(imm_df$immune_product)

results_list <- vector("list", length(immune_products))
contrast_list <- vector("list", length(immune_products))

for (i in seq_along(immune_products)) {
  prod <- immune_products[i]
  
  tbl <- imm_df %>% 
    filter(immune_product == prod)
  
  art_out <- art_lmer_strat(
    input_table = tbl, 
    strat_variable = "treatment_group", 
    model_formula_left = "calc_concentration",
    model_formula_right = "new_visit_label + (1|subject_id)",
    model_contrast_run_by = "new_visit_label"
  )
  
  results_list[[i]] <- art_out$ARTResults %>% 
    mutate(immune_product = prod)
  contrast_list[[i]] <- art_out$ARTContrast %>% 
    mutate(immune_product = prod)
}

# bind them all together
treatment_strat_raw_results <- bind_rows(results_list) %>% 
  rename(p.adj.signif = signif) 
treatment_strat_raw_contrast <- bind_rows(contrast_list) %>% 
  rename(p.adj.signif = signif) %>% 
  mutate(y.position = case_when(
    treatment_group == 'CRS3123_200mg_po_bid' ~ 1,
    treatment_group == 'CRS3123_400mg_po_bid' ~ 1,
    treatment_group == 'Vancomycin_125mg_po_qid' ~ 1), 
    ## pulls the 'new_visit_num' off of the values in these cols so it matches the plot below 
    group1 = gsub("new_visit_label*", "", group1),
    group2 = gsub("new_visit_label*", "", group2)) 

treatment_strat_significant_contrast <- treatment_strat_raw_contrast %>% 
  filter(p.adj.signif != 'ns')
```


```{r, fig.height=6, fig.width=10}
il1b_df <- il1b_df %>%
  filter(!is.na(new_visit_label)) %>% 
  mutate(
    new_visit_label = factor(
      new_visit_label,
      levels = c("Screen", "D1", "D4", "D10-15/TOC", "D17", "D40", "D70")
    )
  )

il1b_contrasts <- treatment_strat_significant_contrast %>%
  filter(immune_product == "IL-1b") %>% 
  mutate(
    group1 = ifelse(group1 == "(D10-15/TOC)", "D10-15/TOC", group1),
    group2 = ifelse(group2 == "(D10-15/TOC)", "D10-15/TOC", group2)
  )

(il1b_plot <- il1b_df %>%
  mutate(
    log_conc = log10(calc_concentration)
  ) %>%
  ggplot(aes(x = new_visit_label, y = log_conc)) +
  geom_boxplot(aes(group = new_visit_label), outlier.shape = NA) +
  geom_point(size = 2, alpha = 0.4) +  # non-colored point plot
  geom_line(aes(group = subject_id), alpha = 0.2) +
  facet_grid(
    immune_product ~ treatment_group,
    scales = "free_y",
    labeller = labeller(
      treatment_group = treat_group_labs,
      immune_product  = immune_prod_labs
    )
  ) +
  scale_y_continuous() +
  scale_x_discrete(labels = visit_x_labs) +
  stat_pvalue_manual(
    il1b_contrasts,
    step.group.by = "treatment_group",  # step separately per treatment facet
    step.increase = 0.05,
    hide.ns       = TRUE,
    label         = "p.adj.signif",
    y.position = 5
  ) +
  theme_bw(base_size = 14) +
  theme(strip.text.y.right = element_text(angle = 0)) +
  labs(
    x     = "Visit",
    y     = "Log10 Concentration (pg/mL)",
    title = "IL-1b in Fecal Water Over Time"
  ))
```

```{r, fig.width=12, fig.height=8}
visit_levels <- c("Screen","D1","D4","D10-15/TOC","D17","D40","D70")

two_marker_df <- imm_df %>%
  filter(immune_product %in% c("IL-1b", "calprotectin")) %>%
  mutate(
    new_visit_label = factor(new_visit_label, levels = visit_levels),
    log_conc        = log10(calc_concentration)
  ) %>%
  filter(!is.na(new_visit_label)) %>%
  mutate(new_visit_label = droplevels(new_visit_label))


two_marker_contrasts <- treatment_strat_significant_contrast %>%
  filter(immune_product %in% c("IL-1b", "calprotectin")) %>%
  mutate(
    group1 = ifelse(group1 == "(D10-15/TOC)", "D10-15/TOC", group1),
    group2 = ifelse(group2 == "(D10-15/TOC)", "D10-15/TOC", group2),
    immune_product = trimws(immune_product)  # remove any stray spaces
  ) %>%
  group_by(immune_product, treatment_group) %>%
  mutate(
    y.position = case_when(
      immune_product == "IL-1b" ~ 4.5 + 0.25 * (row_number() - 1),
      grepl("calprotectin", immune_product, ignore.case = TRUE) ~ 8.5 + 0.2 * (row_number() - 1)
    )
  ) %>%
  ungroup()

(two_marker_plot <- two_marker_df %>%
  ggplot(aes(x = new_visit_label, y = log_conc)) +
  geom_boxplot(aes(group = new_visit_label), outlier.shape = NA) +
  geom_point(size = 2.3, alpha = 0.4) +
  geom_line(aes(group = subject_id), alpha = 0.2) +
  facet_grid(
    immune_product ~ treatment_group,
    scales  = "free_y",
    labeller = labeller(
      treatment_group = treat_group_labs,
      immune_product  = immune_prod_labs
    )
  ) +
  scale_y_continuous() +
  scale_x_discrete(labels = visit_x_labs) +
  stat_pvalue_manual(
    two_marker_contrasts,
    step.group.by = "treatment_group",  # separate stepping per treatment facet
    hide.ns       = TRUE,
    label         = "p.adj.signif"
    # uses y.position from two_marker_contrasts
  ) +
  theme_bw(base_size = 16) +
  theme(strip.text.y.right = element_text(angle = 0)) +
  labs(
    x     = "Visit",
    y     = "Log10 Concentration (pg/mL)",
    title = "Changes in IL-1b and Calprotectin over Time by Treatment"
  ))
```

```{r}
ggsave("../plots/il1b_cp_by_treat.png", two_marker_plot, width = 15, height = 8)
save(two_marker_plot, file = "../../../figures/ggplots/fig7a_cp_il1b.rdat")
```

