---
title: "sIF_pcoa"
author: "Laurie Lyon"
date: "2025-08-25"
output: html_document
---
Goal: create feature table and metadata to put into QIIME2 to create PCoA plots for fecal water sIF
- see Littlefield et al. 2022 Frontiers Imm. paper Figure 3 

```{r}
library(tidyverse)
library(readr)
library(vegan)
library(ape) #for pcoa
library(ggrepel)
theme_set(theme_bw(base_size = 12))
set.seed(1)
library(viridis)
```
## read in table
can use any .csv that has been censored for LLOD and ULOD 
(consider looking into Brandie's methods for imputing below LLOD to help with data distribution)

```{r}
sif_long <- read_csv("../results/merged_censored_immune_markers.csv")
```

## pivot long to wide
- starting with just all time points together, consider filtering by time point first later?
- I think there's one duplicate (calprotectin 101-1002.4), so adding values_fn to take mean of duplicates)

```{r}
sif_wide <- sif_long %>% 
  #filter(new_visit_label == "D10-15/TOC") %>% # could filter for single day
  dplyr::select(sampleid, immune_product, calc_concentration_mean) %>% 
  group_by(sampleid, immune_product) %>% 
  pivot_wider(names_from = immune_product, 
              values_from = calc_concentration_mean, 
              values_fn = mean)
```

## metadata
- could add other things but just keeping the basics here
- other clinical metadata could be helpful
- using the distinct dplyr function to make sure we only have one set of metadata per sample (needs the input list of columns to check)

```{r}
meta_cols <- c("sampleid","subject_id","treatment_group","new_visit_label","new_visit_num")
sample_metadata <- sif_long %>% 
  distinct(across(all_of(meta_cols))) %>% 
  mutate(abx_time = case_when(
    new_visit_num %in% c(1, 2) ~ "Screen, D1",
    new_visit_num %in% c(3) ~ "D4",
    new_visit_num %in% c(4) ~ "TOC",
    new_visit_num %in% c(5, 6, 7) ~ "D17, D40, D70", 
    .default = NA
  ) 
         )
```

## normalize
- divide each individual value by the mean value of that sIF across all samples
- across() applies same transformation to multiple columns (-sampleid to ignore that column)
  - -sampleid was causing an error so changed to where(is.numeric)
- purrr-function mean(.x, na.rm - TRUE) calculates mean of current column, ignoring NAs
- so .x / mean(.x, na.rm = TRUE) divides each value in the current column by that overall mean value (for unit testing, mean of each column should = 1)

## consider normalizing like relative abundance but for each cytokine

```{r}
sif_norm_median <- sif_wide %>% 
  ungroup() %>% 
  mutate(across(where(is.numeric), #perform only on columns with numeric values (not sampleid)
                ~ .x / median(.x, na.rm = TRUE))) %>% #divide value by column mean
  na.omit()
```

## write .tsv
```{r}
write_tsv(sif_norm_median, "../data/sif_norm_median_table.tsv")
write_tsv(sample_metadata, "../data/sif_sample_metadata.tsv")
```

## wrangling tables to input into canberra 
```{r}
#look for common ids between meta and sif concentrations
common_ids <- intersect(sif_norm_median$sampleid, sample_metadata$sampleid)

sif_filt_aligned <- sif_norm_median %>% 
  filter(sampleid %in% common_ids) %>% 
  arrange(match(sampleid, common_ids))

meta_filt_aligned <- sample_metadata %>% 
  filter(sampleid %in% common_ids) %>% 
  arrange(match(sampleid, common_ids))

#sanity check to make sure sampleids match
stopifnot(identical(sif_filt_aligned$sampleid, meta_filt_aligned$sampleid))

#something is going on with select in my environment so I have to specify dplyr
#need to remove the non-numeric column (sampleid) before converting to matrix
sif_nosampleid <- sif_filt_aligned %>% 
  dplyr::select(-sampleid)

#convert to matrix and put sampleid back as rownames 
sif_matrix <- as.matrix(sif_nosampleid)
rownames(sif_matrix) <- sif_filt_aligned$sampleid
```

```{r}
sif_dist <- vegdist(sif_matrix, method = "canberra")

sif_pcoa <- ape::pcoa(sif_dist)

pct <- round(100 * sif_pcoa$values$Relative_eig[1:3], 2)

# assemble sample coodinates with metadata
sif_pcoa_coords <- as_tibble(sif_pcoa$vectors[, 1:3], .name_repair = "minimal") %>%
  setNames(c("PC1","PC2","PC3")) %>%
  mutate(sampleid = rownames(sif_pcoa$vectors)) %>%
  left_join(meta_filt_aligned, by = "sampleid")

```

```{r}
#treatment groups
treatments <- sort(unique(sif_pcoa_coords$treatment_group))

# # ordination subspaces for envfit
# ord_1_2 <- as.data.frame(sif_pcoa$vectors[, 1:2])
# colnames(ord_1_2) <- c("Axis.1","Axis.2")
# rownames(ord_1_2) <- rownames(sif_matrix)
# 
# ord_2_3 <- as.data.frame(sif_pcoa$vectors[, 2:3])
# colnames(ord_2_3) <- c("Axis.2","Axis.3")
# rownames(ord_2_3) <- rownames(sif_matrix)

# % variance for axis labels
pct <- round(100 * sif_pcoa$values$Relative_eig[1:3], 2)
pc1_lab <- paste0("PC1 (", pct[1], "%)")
pc2_lab <- paste0("PC2 (", pct[2], "%)")
pc3_lab <- paste0("PC3 (", pct[3], "%)")

# #empty list for combining later
# envfit_list_1_2 <- list()  # per-treatment cytokine vectors for PC1-PC2
# envfit_list_2_3 <- list()  # per-treatment cytokine vectors for PC2-PC3
# 
# #run envfit separately on each group 
# for (trt in treatments) {
#   #get sample IDs per treatment
#   ids_trt <- sif_pcoa_coords %>% 
#     filter(treatment_group == trt) %>% 
#     pull(sampleid)
#   #subset ordination plane rows to only those sample IDs
#   ord_1_2_sub <- ord_1_2[ids_trt, , drop = FALSE]
#   ord_2_3_sub <- ord_2_3[ids_trt, , drop = FALSE]
#   #subset cytokine matrix rows for those samples
#   env_sub <- as.data.frame(sif_matrix[ids_trt, , drop = FALSE])
#   #run envfit for that treatment for each plane 
#   ef_1_2_trt <- vegan::envfit(
#     ord = ord_1_2_sub,
#     env = env_sub,
#     permutations = 999
#   )
#   ef_2_3_trt <- vegan::envfit(
#     ord = ord_2_3_sub,
#     env = env_sub,
#     permutations = 999
#   )
#   #extract vectors for PC1_2
#   vec_raw_1_2 <- as.data.frame(scores(ef_1_2_trt, display = "vectors"))
#   colnames(vec_raw_1_2)[1:2] <- c("Axis.1","Axis.2")
#   #make tibble
#   vec_tbl_1_2 <- vec_raw_1_2 %>%
#     rownames_to_column("Cytokine") %>%
#     as_tibble() %>%
#     mutate(
#       r2   = (ef_1_2_trt$vectors$r)^2,
#       pval = ef_1_2_trt$vectors$pvals,
#       padj = p.adjust(pval, method = "BH"),
#       treatment_group = trt
#     ) %>%
#     arrange(padj, desc(r2))
#   #select significant vectors 
#   vec_sig_1_2 <- vec_tbl_1_2 %>% 
#     filter(padj <= 0.05)
#   
#   #extract vectors for PC2_3
#   vec_raw_2_3 <- as.data.frame(scores(ef_2_3_trt, display = "vectors"))
#   colnames(vec_raw_2_3)[1:2] <- c("Axis.2","Axis.3")
#   #make tibble
#   vec_tbl_2_3 <- vec_raw_2_3 %>%
#     rownames_to_column("Cytokine") %>%
#     as_tibble() %>%
#     mutate(
#       r2   = (ef_2_3_trt$vectors$r)^2,
#       pval = ef_2_3_trt$vectors$pvals,
#       padj = p.adjust(pval, method = "BH"),
#       treatment_group = trt
#     ) %>%
#     arrange(padj, desc(r2))
#   #select significant vectors 
#   vec_sig_2_3 <- vec_tbl_2_3 %>% 
#     filter(padj <= 0.05)
#   
#   #store results 
#   envfit_list_1_2[[trt]] <- vec_sig_1_2
#   envfit_list_2_3[[trt]] <- vec_sig_2_3
# }
# 
# #combine all treatment-specific arrows into long dfs
# all_arrows_1_2 <- bind_rows(envfit_list_1_2)
# all_arrows_2_3 <- bind_rows(envfit_list_2_3)

#code that I used for single plot not splitting by treatment (used to make function above)
# Create a 3-axis ordination data frame for envfit
ord3 <- as.data.frame(sif_pcoa$vectors[, 1:3])
colnames(ord3) <- c("Axis.1","Axis.2","Axis.3")
rownames(ord3) <- rownames(sif_matrix)

# % variance for axis labels
pct <- round(100 * sif_pcoa$values$Relative_eig[1:3], 2)
pc1_lab <- paste0("PC1 (", pct[1], "%)")
pc2_lab <- paste0("PC2 (", pct[2], "%)")
pc3_lab <- paste0("PC3 (", pct[3], "%)")

#PC1_PC2
# Fit cytokines as vectors (correlations with axes) with envfit
ef1_2 <- vegan::envfit(ord = as.matrix(ord3),
                     env = as.data.frame(sif_matrix),
                     permutations = 999,
                     choices = 1:2)

# extract vector coordinates
vec_raw_1_2 <- as.data.frame(scores(ef1_2,
                                display = "vectors",
                                choices = 1:2))

# now we have vectors (correlations of each cytokine with the ordination axis) and pvals for significance
# r2 is how well each cytokine aligns with that axis
# permutation test determines signficance
vec_tbl_1_2 <- vec_raw_1_2 %>%
  rownames_to_column("Cytokine") %>%
  as_tibble() %>%
  mutate(
    r2   = (ef1_2$vectors$r)^2,
    pval = ef1_2$vectors$pvals,
    padj = p.adjust(pval, method = "BH")
  ) %>%
  arrange(padj, desc(r2))

##PC2_PC3
# Fit cytokines as vectors (correlations with axes) with envfit
ef2_3 <- vegan::envfit(ord = as.matrix(ord3),
                     env = as.data.frame(sif_matrix),
                     permutations = 9999,
                     choices = 2:3)

# extract vector coordinates
vec_raw_2_3 <- as.data.frame(scores(ef2_3,
                                display = "vectors",
                                choices = 1:2)) #1:2 because I already selected 2:3 in the last step

# now we have vectors (correlations of each cytokine with the ordination axis) and pvals for significance
# r2 is how well each cytokine aligns with that axis
# permutation test determines signficance
vec_tbl_2_3 <- vec_raw_2_3 %>%
  rownames_to_column("Cytokine") %>%
  as_tibble() %>%
  mutate(
    r2   = (ef2_3$vectors$r)^2,
    pval = ef2_3$vectors$pvals,
    padj = p.adjust(pval, method = "BH")
  ) %>%
  arrange(padj, desc(r2))

#pick significant cytokines to graph
vec_sig_1_2 <- vec_tbl_1_2 %>%
  filter(padj <= 0.05)

vec_sig_2_3 <- vec_tbl_2_3 %>%
  filter(padj <= 0.05)

top_vec_1_2 <- vec_sig_1_2 %>% 
  head(10)
  
top_vec_2_3 <- vec_sig_2_3 %>% 
  head(10)
```

**graphing**
```{r, fig.width=18, fig.height=5}
sif_pcoa_coords <- sif_pcoa_coords %>%
  arrange(subject_id, new_visit_num)

arrow_scale_1_2 <- 1.5 * min(
  diff(range(sif_pcoa_coords$PC1)),
  diff(range(sif_pcoa_coords$PC2))
)

arrows_1_2 <- top_vec_1_2 %>%
  transmute(
    Cytokine,
    x0 = 0,
    y0 = 0,
    x1 = Axis.1 * arrow_scale_1_2,
    y1 = Axis.2 * arrow_scale_1_2,
    r2,
    padj
  )
# plot_arrows_1_2 <- all_arrows_1_2 %>%
#   mutate(
#     x0 = 0,
#     y0 = 0,
#     x1 = Axis.1 * arrow_scale_1_2,
#     y1 = Axis.2 * arrow_scale_1_2
#   )

#more explicity define the color palette to match other figures
abx_levels <- c("Screen, D1", "D4", "TOC", "D17, D40, D70")

# generate the actual colors once, from viridis
abx_palette <- viridis(
  n = length(abx_levels),
  option = "C",
  begin = 0,
  end = 0.95
)
names(abx_palette) <- abx_levels

(p_pc1_pc2_facet <- ggplot(
  sif_pcoa_coords, 
  aes(x = PC1, y = PC2)) +
  # geom_path(
  #   data = sif_pcoa_coords,
  #   aes(x = PC1, y = PC2,
  #       group = subject_id,
  #       color = abx_time
  #       ),
  #   alpha = 0.4,
  #   linewidth = 0.6
  # ) +
    # focal samples for each panel (will be re-filtered per facet below)
  geom_point(
    aes(fill = abx_time),
    shape = 21,
    size = 3,
    alpha = 0.6
  ) +
  stat_ellipse(
    aes(color = abx_time),
    type = "t",
    level = 0.95,
    linewidth = 1, 
    fill = NA
  ) +
  scale_fill_manual(
    values = abx_palette,
    breaks = abx_levels,
    name = "Time Relative to Abx"
  ) +
  scale_color_manual(
    values = abx_palette,
    breaks = abx_levels,
    name = "Time Relative to Abx"
  ) +
  #cytokine arrows
  geom_segment(
    data = arrows_1_2,
    aes(x = x0, y = y0, xend = x1, yend = y1),
    arrow = arrow(length = unit(0.2, "cm")),
    linewidth = 0.3,
    color = "black"
  ) +
  # cytokine labels
  geom_text_repel(
    data = arrows_1_2,
    aes(x = x1, y = y1, label = Cytokine),
    size = 5,
    color = "black",
    max.overlaps = 200, 
    fontface = "bold", 
    direction = "both"
  ) +
  facet_wrap(~ treatment_group, 
             labeller = labeller(
              treatment_group = c(
                "CRS3123_200mg_po_bid" = "CRS3123 200mg",
                "CRS3123_400mg_po_bid" = "CRS3123 400mg",
                "Vancomycin_125mg_po_qid" = "Vancomycin 125mg"))) +
  labs(
    title = "Inflammatory Marker Profiles from Fecal Water",
    x = pc1_lab,
    y = pc2_lab,
    color = "Time Relative to Abx"
  ) +
  guides(
    fill  = guide_legend(override.aes = list(shape = 21, alpha = 0.8, color = "grey30")),
    color = guide_legend(override.aes = list(linewidth = 1)) 
    ) +
  theme_bw(base_size = 16))

arrow_scale_2_3 <- 0.9 * min(
  diff(range(sif_pcoa_coords$PC2)),
  diff(range(sif_pcoa_coords$PC3))
)

arrows_2_3 <- top_vec_2_3 %>%
  mutate(
    x0 = 0,
    y0 = 0,
    x1 = Axis.2 * arrow_scale_2_3,
    y1 = Axis.3 * arrow_scale_2_3
  )

(p_pc2_pc3_facet <- ggplot() +
  geom_point(
    data = sif_pcoa_coords,
    aes(x = PC2, y = PC3,
        color = abx_time),
    size = 2.3,
    alpha = 0.9
  ) +
  scale_shape_manual(values = c(21,24)) +
  stat_ellipse(
    data = sif_pcoa_coords,
    aes(x = PC2, y= PC3, color = abx_time, fill = abx_time),
    type = "t",
    level = 0.95,
    geom = "polygon",
    alpha = 0.1,
    linewidth = 1
  ) +
  # treatment-specific cytokine arrows
  geom_segment(
    data = arrows_2_3,
    aes(x = x0, y = y0, xend = x1, yend = y1),
    arrow = arrow(length = unit(0.22, "cm")),
    linewidth = 0.5,
    color = "black"
  ) +
  # cytokine labels
  geom_text_repel(
    data = arrows_2_3,
    aes(x = x1, y = y1, label = Cytokine),
    size = 4,
    color = "black",
    max.overlaps = 200, 
    fontface = "bold", 
    direction = "both"
  ) +
  facet_wrap(~ treatment_group, 
             labeller = labeller(
              treatment_group = c(
                "CRS3123_200mg_po_bid" = "CRS3123 200mg",
                "CRS3123_400mg_po_bid" = "CRS3123 400mg",
                "Vancomycin_125mg_po_qid" = "Vancomycin 125mg"))) +
  labs(
    title = "PCoA Canberra (PC2 vs PC3)",
    x = pc2_lab,
    y = pc3_lab,
    color = "Time Relative to Abx"
  ) +
  theme_bw(base_size = 16))
```

```{r}
ggsave("../plots/sIF_pcoa.pdf", p_pc1_pc2_facet, width = 18, height = 5.5)
ggsave("../plots/sIF_pcoa.png", p_pc1_pc2_facet, width = 18, height = 5.5)
save(p_pc1_pc2_facet, file = "../../../figures/fig7a_sif_pcoa.rdat")
```

## 9-5-25
adding on to this script to transform this table back to long and run LMMs
# wide back to long 
```{r}
# taking the transformed wide data frame back to long 
sif_norm_long <- sif_norm_median %>%
  pivot_longer(
    cols = -sampleid, #gather all columns except sampleid
    names_to = "immune_product", # new column for immune factors
    values_to = "normed_conc" # The new column for their values
  )

#merge with metadata again 
sif_norm_full <- sif_norm_long %>% 
  left_join(sample_metadata, by = "sampleid")
```

```{r}
write_tsv(sif_norm_full, "../data/sif_norm_long_w_metadata.tsv")
```

**adonis on dist matrix**
```{r}
#with all samples
meta_filt_aligned$new_visit_label <- factor(
  meta_filt_aligned$new_visit_label,
  levels = c("Screen","D1","D4","D10-15/TOC","D17","D40","D70")
)

sif_adonis <- adonis2(sif_dist ~ new_visit_label + treatment_group,
                      data = meta_filt_aligned, 
                      permutations = 999, 
                      strata = meta_filt_aligned$subject_id,  
                      by = "margin")
```

```{r}
#with only visits that happen once the antibiotic is given
post_abx_visits <- c("D4", "D10-15/TOC", "D17", "D40", "D70")

meta_post_abx <- meta_filt_aligned %>% 
  filter(new_visit_label %in% post_abx_visits)

#sample labels from the distance object
dist_labels <- attr(sif_dist, "Labels")

# Keep only samples that are in both the distance matrix and post-treatment metadata
meta_post_abx <- meta_post_abx %>%
  filter(sampleid %in% dist_labels) %>%
  arrange(match(sampleid, dist_labels))

# Indices of these samples in the distance matrix
keep_idx <- dist_labels %in% meta_post_abx$sampleid

# Subset the distance matrix
sif_dist_post <- as.dist(as.matrix(sif_dist)[keep_idx, keep_idx])

# quick sanity check
stopifnot(length(sif_dist_post) > 0)

sif_postabx_adonis <- adonis2(sif_dist_post ~ new_visit_label + treatment_group,
                      data = meta_post_abx, 
                      permutations = 999, 
                      strata = meta_post_abx$subject_id,  
                      by = "margin")

sif_postabx_adonis
```

Pairwise PERMANOVA
- wrangle - get labels, reorder metadata
- did this for the data with all timepoints and everything is significant (I think because of baseline Screen+D1 differences)
- trying with only the post-abx timepoints

```{r}
#sample labels from the distance object
dist_labels <- attr(sif_dist, "Labels")

#reorder metadata to match sif_dist order if it's not
meta_filt_aligned2 <- meta_filt_aligned %>%
  filter(sampleid %in% dist_labels) %>%
  arrange(match(sampleid, dist_labels))

meta_filt_aligned2$treatment_group <- factor(meta_filt_aligned2$treatment_group)
meta_filt_aligned2$subject_id      <- factor(meta_filt_aligned2$subject_id)

#check for treatment names
levels(meta_filt_aligned2$treatment_group)

```

CRS200 vs. 400 pairwise
```{r}
t1 <- "CRS3123_200mg_po_bid"
t2 <- "CRS3123_400mg_po_bid"

sub_meta <- meta_filt_aligned2 %>%
  filter(treatment_group %in% c(t1, t2))

# indices into the distance matrix using dist_labels
keep_ids <- sub_meta$sampleid
idx      <- dist_labels %in% keep_ids

sub_dist <- as.dist(as.matrix(sif_dist)[idx, idx])

res_200_vs_400 <- adonis2(
  sub_dist ~ new_visit_label + treatment_group,
  data        = sub_meta,
  permutations = 999,
  strata       = sub_meta$subject_id, 
  by = "margin"
)

res_200_vs_400
```
```{r}
t1 <- "CRS3123_200mg_po_bid"
t3 <- "Vancomycin_125mg_po_qid"

sub_meta <- meta_filt_aligned2 %>%
  filter(treatment_group %in% c(t1, t3))

keep_ids <- sub_meta$sampleid
idx      <- dist_labels %in% keep_ids

sub_dist <- as.dist(as.matrix(sif_dist)[idx, idx])

res_200_vs_vanc <- adonis2(
  sub_dist ~ new_visit_label + treatment_group,
  data        = sub_meta,
  permutations = 999,
  strata       = sub_meta$subject_id, 
  by = "margin"
)

res_200_vs_vanc
```

```{r}
t2 <- "CRS3123_400mg_po_bid"
t3 <- "Vancomycin_125mg_po_qid"

sub_meta <- meta_filt_aligned2 %>%
  filter(treatment_group %in% c(t2, t3))

keep_ids <- sub_meta$sampleid
idx      <- dist_labels %in% keep_ids

sub_dist <- as.dist(as.matrix(sif_dist)[idx, idx])

res_400_vs_vanc <- adonis2(
  sub_dist ~ new_visit_label + treatment_group,
  data        = sub_meta,
  permutations = 999,
  strata       = sub_meta$subject_id, 
  by = "margin"
)

res_400_vs_vanc
```

only want to look at post abx (not screen and D1)
function for the pairwise written by ChatGPT 5.1 11-18-25
```{r}
post_visits <- c("D4", "D10-15/TOC", "D17", "D40", "D70")

meta_post_abx <- meta_filt_aligned %>%
  filter(new_visit_label %in% post_visits)

meta_post_abx$treatment_group <- factor(meta_post_abx$treatment_group)
meta_post_abx$subject_id      <- factor(meta_post_abx$subject_id)

meta_post_abx <- meta_post_abx %>%
  filter(sampleid %in% dist_labels) %>%
  arrange(match(sampleid, dist_labels))

# Indices of these samples in the distance matrix
keep_idx <- dist_labels %in% meta_post_abx$sampleid

# Subset the distance matrix
sif_dist_post <- as.dist(as.matrix(sif_dist)[keep_idx, keep_idx])

# quick sanity check
stopifnot(length(sif_dist_post) > 0)
```

overall Adonis post-abx
```{r}
adonis_post <- adonis2(
  sif_dist_post ~ new_visit_label + treatment_group,
  data        = meta_post_abx,
  permutations = 999,
  strata       = meta_post_abx$subject_id,
  by           = "terms"   # or "margin" if you prefer marginal tests
)

adonis_post
```


```{r}
pairwise_post <- function(g1, g2, meta_post, sif_dist_post) {

  # get sample IDs in the pair
  sub_meta <- meta_post %>% 
    dplyr::filter(treatment_group %in% c(g1, g2))

  # get labels from the POST distance matrix
  post_labels <- attr(sif_dist_post, "Labels")

  # find indices for samples belonging to g1 or g2
  idx <- post_labels %in% sub_meta$sampleid

  # subset the distance matrix
  sub_dist <- as.dist(as.matrix(sif_dist_post)[idx, idx])

  # now run the PERMANOVA adjusting for visit
  res <- adonis2(
    sub_dist ~ new_visit_label + treatment_group,
    data        = sub_meta,
    permutations = 999,
    strata       = sub_meta$subject_id,
    by           = "terms"   # tests treatment controlling for visit
  )

  return(res)
}
```

```{r}
res_200_vs_400_post <- pairwise_post(t1, t2, meta_post_abx, sif_dist_post)
res_200_vs_400_post

res_200_vs_vanc_post <- pairwise_post(t1, t2, meta_post_abx, sif_dist_post)
res_200_vs_vanc_post

res_400_vs_vanc_post <- pairwise_post(t1, t2, meta_post_abx, sif_dist_post)
res_400_vs_vanc_post
```
-results show that even only looking at the profiles after abx (D4 and later), all treatments are significantly different 
