---
title: "bile_acid_analysis"
output: html_document
date: "2025-08-18"
author: "Madi Apgar"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
library(magrittr)
library(vroom)
library(apppleplots)
```

**Functions**
```{r}
source("../../art_lmer_functions.R")
source("../../diff_between_means_calc.R")
source("../../test_lmer_vars_function.R")

## function that replaces the need for case_when/if_else statements to assign column values (i.e. bile acids) to
## larger categories (primary vs secondary)
## takes a named list (list_names) of character vectors (list_values)
bin_bile_acids <- function(bin_list,
                           list_names,
                           list_values,
                           input_table,
                           group_by_cols,
                           summed_col){
  all_grouped_cols <- append(group_by_cols, list_names)
  
  ## enframe converts named lists/vectors to two col dataframes 
  summed_df <- enframe(bin_list, name = list_names, value = list_values) %>% 
    unnest(all_of(list_values)) %>% 
    right_join(input_table, by = list_values) %>% 
    group_by(across(all_of(all_grouped_cols))) %>% 
    summarise(summed_conc = sum(.data[[summed_col]])) %>% 
    filter(!is.na(.data[[list_names]]))
  
  return(summed_df)
}

## small function that calculates the ratio between two values in a column (takes it from long to wide)
calculate_ratio <- function(input_table,
                            ratio_col,
                            ratio_values,
                            numerator,
                            denominator){
  ratio_df <- input_table %>% 
    spread(key = ratio_col, value = ratio_values) %>% 
    mutate(ratio = (.data[[numerator]]/.data[[denominator]]))
  
  return(ratio_df)
}

## pretty big function that takes a named list input with all the wanted selections for running art lmer on the data
## need to specify:
## - an input dataframe
## - cols converted to factors (for art bc it breaks if not)
## - which to variables you want the tests stratified by 
## - cols for left and right sides of the formula 
## - which right hand side variable you want the contrasts run by
lazy_girl_stats <- function(stats_list){
  stats_res <- tibble()
  stats_contrast <- tibble()
  
  for (stats in unlist(names(stats_list))){
    
    factored_df <- stats_list[[stats]][['df']] %>% 
      ## this looks a lot scarier than it is, just converting the left hand side col to log10
      mutate(log_conc = log10(.data[[stats_list[[stats]][['model_formula_left']]]]))
    
    ## converting all strat/right hand side vars to factors
    all_factor_cols <- stats_list[[stats]][['factor']]
    factored_df[all_factor_cols] <- lapply(factored_df[all_factor_cols], as.factor)
    
    if (hasName(stats_list[[stats]], 'strat_variable2') & 
        hasName(stats_list[[stats]], 'strat_variable')){
      art_lmer <- art_lmer_doubleStrat(input_table = factored_df,
                                       strat_variable1 = stats_list[[stats]][['strat_variable']],
                                       strat_variable2 = stats_list[[stats]][['strat_variable2']],
                                       model_formula_left = 'log_conc',
                                       model_formula_right =
                                         stats_list[[stats]][['model_formula_right']],
                                       model_contrast_run_by =
                                         stats_list[[stats]][['model_run_contrast_by']])
      
      art_res <- art_lmer$ARTResults %>% 
        mutate(stats_run = stats) %>% 
        rename(strat_var = stats)
      
      art_contrast <- art_lmer$ARTContrast %>% 
        mutate(stats_run = stats) %>% 
        rename(strat_var = stats) 
    } else if (hasName(stats_list[[stats]], 'strat_variable') & 
               hasName(stats_list[[stats]], 'strat_variable_name')) {
      art_lmer <- art_lmer_strat(input_table = factored_df,
                               strat_variable = stats_list[[stats]][['strat_variable']],
                               model_formula_left = 'log_conc',
                               model_formula_right =
                                         stats_list[[stats]][['model_formula_right']],
                               model_contrast_run_by =
                                         stats_list[[stats]][['model_run_contrast_by']])
  
      art_res <- art_lmer$ARTResults %>% 
        mutate(stats_run = stats,
               strat_var = stats_list[[stats]][['strat_variable_name']])
      art_contrast <- art_lmer$ARTContrast %>% 
        mutate(stats_run = stats,
               strat_var = stats_list[[stats]][['strat_variable_name']]) 
    } else {
      print("Please specify at least one variable to stratify the statistics by!")
    }
    
    stats_res <- rbind(stats_res, art_res)
    stats_contrast <- rbind(stats_contrast, art_contrast)
  }
  ## creating list of outputs 
  my_list <- list('Contrasts' = stats_contrast,
                  'Results' = stats_res)
  return(my_list)
}

## small function to assign a value to a key in a dataframe column (i.e. numbers to actual categorical variables)
values_to_keys <- function(input_table,
                           key_col_name,
                           value_col_name,
                           key_value_list,
                           col_order_list){
  
  input_table[value_col_name] <- unlist(lapply(unlist(input_table[key_col_name]), 
                                        FUN = function(x){key_value_list[[x]]}))
  
  output_table <- input_table %>%
    select(all_of(col_order_list), everything())
  
  return(output_table)
}
```

**File paths**
```{r}
bile_acid_fp <- '../data/bile_acid_results.txt'
visit_key_fp <- '../data/visit_name_key.tsv'
meta_fp <- '../../../data_submission/1_16S_rRNA_qiita/CRS3123_fecal_samples/crestone_qiita_sample_information.tsv'

## global lists for renaming plot labels
visit_x_labs <- c('1' = 'Screen',
                  '2' = 'D1',
                  '3' = 'D4',
                  '4' = 'TOC',
                  '5' = 'D17', 
                  '6' = 'D40', 
                  '7' = 'D70') 

treat_group_labs <- c('CRS3123 200mg',
                      'CRS3123 400mg',
                      'Vancomycin 125mg')
names(treat_group_labs) <- c('CRS3123_200mg_po_bid',
                             'CRS3123_400mg_po_bid',
                             'Vancomycin_125mg_po_bid')

state_labs <- c('conjugated' = 'Conjugated',
                'deconjugated' = 'Deconjugated',
                'dehydroxylated' = 'Dehydroxylated')

class_labs <- c('primary' = 'Primary',
                'secondary' = 'Secondary')

cdiff_effect_labs <- c('promotes' = 'Promotes',
                       'inhibits' = 'Inhibits')
```

**Reading in files**
```{r}
bile_acid_df <- read_tsv(bile_acid_fp)
visit_key_df <- read_tsv(visit_key_fp)
meta_df <- read_tsv(meta_fp) %>% 
  select(anonymized_name, host_subject_id, visit, visit_number, treatment_group) %>%
  rename(sampleid = anonymized_name,
         subject_id = host_subject_id,
         new_visit_label = visit,
         new_visit_num = visit_number) %>%
  ## add visit name key to combine w bile acid data
  left_join(visit_key_df, by = "sampleid")
```

**Data wrangled**
getting bile acid data ready to be combined with the metadata
```{r}
unwanted_cols <- c('SampleType',
                   'UniqueSampleID',
                   'BatchNum',
                   'IntraBatchOrder',
                   'Rep',
                   'ProjectCode',
                   'SampleGroup',
                   'RequisitionNum',
                   'ACMUniqueID')

(proc_bileAcid_df <- bile_acid_df %>% 
  select(!all_of(unwanted_cols)) %>% 
  mutate(visit = paste(visit, 'Cooled')) %>% 
  gather(contains('Acid'), key = "bile_acid", value = "concentration") %>% 
  rename(visit_name_key = visit) %>% 
  filter(!is.na(subject_id)))
```

bile acid classifications (for binning):
- conjugated (created and secreted by the liver, have an amino acid attached)
  - TaurochenodeoxycholicAcid_(TCDCA)
  - GlycoursodeoxycholicAcid_(GUDCA)
  - GlycochenodeoxycholicAcid_(GCDCA)
  - TaurocholicAcid_(TCA)
  - GlycocholicAcid_(GCA)
- deconjugated (amino acid pulled off by microbes)
  - CholicAcid_(CA)
  - ChenodeoxycholicAcid_(CDCA)
  - UrsodeoxycholicAcid_(UDCA)
- dehydroxylated (-OH group pulled off by microbes, secondary bile acids)
  - DeoxycholicAcid_(DCA)
  - LithocholicAcid_(LCA)
  
- primary bile acids
  - TaurochenodeoxycholicAcid_(TCDCA)
  - GlycoursodeoxycholicAcid_(GUDCA)
  - GlycochenodeoxycholicAcid_(GCDCA)
  - TaurocholicAcid_(TCA)
  - GlycocholicAcid_(GCA)
  - CholicAcid_(CA)
  - ChenodeoxycholicAcid_(CDCA)
- secondary bile acids
  - UrsodeoxycholicAcid_(UDCA)
  - DeoxycholicAcid_(DCA)
  - LithocholicAcid_(LCA)
  
- c diff growth promoters
  - CholicAcid_(CA)
  - TaurocholicAcid_(TCA)
- c diff growth inhibitors
  - ChenodeoxycholicAcid_(CDCA) -> yes, this is a primary bile acid but it inhibits spore germination
  - UrsodeoxycholicAcid_(UDCA)
  - DeoxycholicAcid_(DCA)
  - LithocholicAcid_(LCA)
  
bile acid binning lists
```{r}
## bile acids im filtering out bc the concentration was basically zero across all treatment groups 
unwanted_bas <- c('BetaMuricholicAcid_(bMCA)',
                  'GlycodeoxycholicAcid_(GDCA)',
                  'GlycolithocholicAcid_(GLCA)',
                  'TaurodeoxycholicAcid_(TDCA)',
                  'TaurolithocholicAcid_(TLCA)',
                  'TauroursodeoxycholicAcid_(TUDCA)')

## based on bile acid state/what reactions have been performed by enzymes
conjugated_bas <- c('TaurochenodeoxycholicAcid_(TCDCA)',
                    'GlycoursodeoxycholicAcid_(GUDCA)',
                    'GlycochenodeoxycholicAcid_(GCDCA)',
                    'TaurocholicAcid_(TCA)',
                    'GlycocholicAcid_(GCA)')

deconjugated_bas <- c('CholicAcid_(CA)',
                      'ChenodeoxycholicAcid_(CDCA)',
                      'UrsodeoxycholicAcid_(UDCA)')

dehydroxylated_bas <- c('DeoxycholicAcid_(DCA)',
                        'LithocholicAcid_(LCA)')

state_list <- list('conjugated' = conjugated_bas,
                   'deconjugated' = deconjugated_bas,
                   'dehydroxylated' = dehydroxylated_bas)

## based on whether bile acids are primary or secondary
primary_bas <- c('TaurochenodeoxycholicAcid_(TCDCA)',
                 'GlycoursodeoxycholicAcid_(GUDCA)',
                 'GlycochenodeoxycholicAcid_(GCDCA)',
                 'TaurocholicAcid_(TCA)',
                 'GlycocholicAcid_(GCA)',
                 'CholicAcid_(CA)',
                 'ChenodeoxycholicAcid_(CDCA)')

secondary_bas <- c('DeoxycholicAcid_(DCA)',
                   'LithocholicAcid_(LCA)',
                   'UrsodeoxycholicAcid_(UDCA)')

class_list <- list('primary' = primary_bas,
                   'secondary' = secondary_bas)

## based on whether bile acids promote or inhibit cdiff growth
promoter_bas <- c('TaurocholicAcid_(TCA)',
                  'CholicAcid_(CA)')

inhibitor_bas <- c('DeoxycholicAcid_(DCA)',
                   'LithocholicAcid_(LCA)',
                   'UrsodeoxycholicAcid_(UDCA)',
                   'ChenodeoxycholicAcid_(CDCA)')

cdiff_effect_list <- list('promotes' = promoter_bas,
                          'inhibits' = inhibitor_bas)
```
  
then join with metadata!
```{r}
bileAcid_meta_df <- proc_bileAcid_df %>% 
  left_join(meta_df, by = c("subject_id", "visit_name_key")) %>% 
  filter(!is.na(concentration)) %>% 
  mutate(bile_acid = gsub(bile_acid, pattern = "_([^_]+)$", replacement = ""),
         bile_acid = gsub(bile_acid, pattern = "-", replacement = ""),
         log_ready_conc = concentration + (min(concentration[concentration != 0])/2))

## pulling out low concentration bile acids and days 
filt_bileAcid_meta_df <- bileAcid_meta_df %>% 
  filter(!(bile_acid %in% unwanted_bas),
         new_visit_num != 6)
```

using a for loop to do all my data wrangling since its very similar for the three bile acid bins I'm creating
- i have a named list inside a named list to provide additional flexibility for various function inputs without creating
separate lists for each input (more concise)
- also calculate ratios between primary/secondary and promoter/inhibitor bile acids for the qualifying bins 
```{r}
sum_bins_cols <- c('sampleid', 'subject_id', 'new_visit_num', 'treatment_group')

bile_acid_bins <- list('state' = list('list' = state_list, 
                                      'df' = filt_bileAcid_meta_df,
                                      'grouped_cols' = sum_bins_cols,
                                      'sum_col_name' = 'log_ready_conc'),
                       'class' = list('list' = class_list, 
                                      'df' = filt_bileAcid_meta_df,
                                      'grouped_cols' = sum_bins_cols,
                                      'sum_col_name' = 'log_ready_conc',
                                      'numerator' = 'primary', 
                                      'denominator' = 'secondary'),
                       'cdiffEffect' = list('list' = cdiff_effect_list,
                                            'df' = filt_bileAcid_meta_df,
                                            'grouped_cols' = sum_bins_cols,
                                            'sum_col_name' = 'log_ready_conc',
                                            'numerator' = 'promotes', 
                                            'denominator' = 'inhibits'))

for (bin in unlist(names(bile_acid_bins))){
  new_col_name <- paste0('bile_acid_', bin)
  df_var_name <- paste0('summed_bileAcid_', bin, '_df')
  
  tmp_df <- bin_bile_acids(bin_list = bile_acid_bins[[bin]][['list']],
                           list_names = new_col_name,
                           list_values = 'bile_acid',
                           input_table = bile_acid_bins[[bin]][['df']],
                           group_by_cols = bile_acid_bins[[bin]][['grouped_cols']],
                           summed_col = bile_acid_bins[[bin]][['sum_col_name']])
  
  assign(df_var_name,
         tmp_df)
  
  if (bin == 'class' | bin == 'cdiffEffect'){
    ratio_df_var_name <- paste0('ratio_bileAcid_', bin, '_df')
    
    tmp_ratio_df <- calculate_ratio(input_table = tmp_df,
                                    ratio_col = new_col_name,
                                    ratio_values = 'summed_conc',
                                    numerator = bile_acid_bins[[bin]][['numerator']],
                                    denominator = bile_acid_bins[[bin]][['denominator']])
    assign(ratio_df_var_name,
           tmp_ratio_df)
  }
  
}
```

**Stats**
stratified by bile acid and visit between treatment groups 
- once again, no significant differences in bile acid concentrations between any of the treatment groups lmao

for loop to run stats for everything together and then concatenate into same table 
```{r}
## testing concentrations of bile acids between treatment groups 
treatGroup_stats_list <- list('bile_acid' = list('df' = filt_bileAcid_meta_df,
                                                'factor' = 
                                                  c('subject_id', 'new_visit_num', 'treatment_group', 'bile_acid'),
                                                'strat_variable' = 'bile_acid',
                                                'strat_variable2' = 'new_visit_num',
                                                'model_formula_left' = 'log_ready_conc',
                                                'model_formula_right' = 'treatment_group',
                                                'model_run_contrast_by' = 'treatment_group'),
                       'bile_acid_state' = list('df' = summed_bileAcid_state_df,
                                                'factor' = 
                                                  c('subject_id', 'new_visit_num', 'treatment_group', 'bile_acid_state'),
                                                'strat_variable' = 'bile_acid_state',
                                                'strat_variable2' = 'new_visit_num',
                                                'model_formula_left' = 'summed_conc',
                                                'model_formula_right' = 'treatment_group',
                                                'model_run_contrast_by' = 'treatment_group'),
                       'bile_acid_class' = list('df' = summed_bileAcid_class_df,
                                                'factor' = 
                                                  c('subject_id', 'new_visit_num', 'treatment_group', 'bile_acid_class'),
                                                'strat_variable' = 'bile_acid_class',
                                                'strat_variable2' = 'new_visit_num',
                                                'model_formula_left' = 'summed_conc',
                                                'model_formula_right' = 'treatment_group',
                                                'model_run_contrast_by' = 'treatment_group'),
                       'bile_acid_cdiffEffect' = list('df' = summed_bileAcid_cdiffEffect_df,
                                                      'factor' = 
                                                        c('subject_id', 'new_visit_num', 'treatment_group', 
                                                          'bile_acid_cdiffEffect'),
                                                      'strat_variable' = 'bile_acid_cdiffEffect',
                                                      'strat_variable2' = 'new_visit_num',
                                                      'model_formula_left' = 'summed_conc',
                                                      'model_formula_right' = 'treatment_group',
                                                      'model_run_contrast_by' = 'treatment_group'),
                       'ratio_bileAcid_class' = list('df' = ratio_bileAcid_class_df,
                                                'factor' = 
                                                  c('subject_id', 'new_visit_num', 'treatment_group'),
                                                'strat_variable' = 'new_visit_num',
                                                'strat_variable_name' = 'primary:secondary_ratio',
                                                'model_formula_left' = 'ratio',
                                                'model_formula_right' = 'treatment_group',
                                                'model_run_contrast_by' = 'treatment_group'),
                         'ratio_bileAcid_cdiffEffect' = list('df' =
                                                                ratio_bileAcid_cdiffEffect_df,
                                                      'factor' = 
                                                        c('subject_id', 'new_visit_num', 'treatment_group'),
                                                      'strat_variable' = 'new_visit_num',
                                                      'strat_variable_name' = 'promoter:inhibitor_ratio',
                                                      'model_formula_left' = 'ratio',
                                                      'model_formula_right' = 'treatment_group',
                                                      'model_run_contrast_by' = 'treatment_group'))

treatGroup_stats <- lazy_girl_stats(stats_list = treatGroup_stats_list)

treatGroup_stats_res <- treatGroup_stats$Results
treatGroup_stats_contrast <- treatGroup_stats$Contrasts

## only pulling significant results so I can visually parse them 
signifOnly_treatGroup_stats_contrast <- treatGroup_stats_contrast %>% 
  filter(signif != 'ns')
```

for stats between treatment groups
```{r}
numOnly_visit_key_value <- c('1' = 'Screen',
                             '2' = 'D1',
                             '3' = 'D4',
                             '4' = 'D10-15/TOC',
                             '5' = 'D17',
                             '6' = 'D40')

treatGroup_col_order <- c('strat_var',
                          'visit_label',
                          'new_visit_num')

signifOnly_treatGroup_stats_contrast_wValues <- values_to_keys(input_table = 
                                                                  signifOnly_treatGroup_stats_contrast,
                                                                key_col_name = 'new_visit_num',
                                                                value_col_name = 'visit_label',
                                                                key_value_list = numOnly_visit_key_value,
                                                                col_order_list = treatGroup_col_order)
```

stratified by treatment group and bile acid between visits
- literally no significant differences in the concentration of DCA between time points so idk what to think of that 
```{r}
## testing concentrations of bile acids within a treatment group between time points 
visit_stats_list <- list('bile_acid' = list('df' = filt_bileAcid_meta_df,
                                            'factor' = 
                                              c('subject_id', 'new_visit_num', 'treatment_group', 'bile_acid'),
                                            'strat_variable' = 'treatment_group',
                                            'strat_variable2' = 'bile_acid',
                                            'model_formula_left' = 'log_ready_conc',
                                            'model_formula_right' = 'new_visit_num + (1|subject_id)',
                                            'model_run_contrast_by' = 'new_visit_num'),
                       'bile_acid_state' = list('df' = summed_bileAcid_state_df,
                                                'factor' = 
                                                  c('subject_id', 'new_visit_num', 'treatment_group',
                                                    'bile_acid_state'),
                                                'strat_variable' = 'treatment_group',
                                                'strat_variable2' = 'bile_acid_state',
                                                'model_formula_left' = 'summed_conc',
                                                'model_formula_right' = 'new_visit_num + (1|subject_id)',
                                                'model_run_contrast_by' = 'new_visit_num'),
                       'bile_acid_class' = list('df' = summed_bileAcid_class_df,
                                                'factor' = 
                                                  c('subject_id', 'new_visit_num', 'treatment_group',
                                                    'bile_acid_class'),
                                                'strat_variable' = 'treatment_group',
                                                'strat_variable2' = 'bile_acid_class',
                                                'model_formula_left' = 'summed_conc',
                                                'model_formula_right' = 'new_visit_num + (1|subject_id)',
                                                'model_run_contrast_by' = 'new_visit_num'),
                       'bile_acid_cdiffEffect' = list('df' = summed_bileAcid_cdiffEffect_df,
                                                      'factor' = 
                                                        c('subject_id', 'new_visit_num', 'treatment_group',
                                                          'bile_acid_cdiffEffect'),
                                                      'strat_variable' = 'treatment_group',
                                                      'strat_variable2' = 'bile_acid_cdiffEffect',
                                                      'model_formula_left' = 'summed_conc',
                                                      'model_formula_right' = 'new_visit_num + (1|subject_id)',
                                                      'model_run_contrast_by' = 'new_visit_num'),
                       'ratio_bileAcid_class' = list('df' = ratio_bileAcid_class_df,
                                                'factor' = 
                                                  c('subject_id', 'new_visit_num', 'treatment_group'),
                                                'strat_variable' = 'treatment_group',
                                                'strat_variable_name' = 'primary:secondary_ratio',
                                                'model_formula_left' = 'ratio',
                                                'model_formula_right' = 'new_visit_num + (1|subject_id)',
                                                'model_run_contrast_by' = 'new_visit_num'),
                         'ratio_bileAcid_cdiffEffect' = list('df' =
                                                                ratio_bileAcid_cdiffEffect_df,
                                                      'factor' = 
                                                        c('subject_id', 'new_visit_num', 'treatment_group'),
                                                      'strat_variable' = 'treatment_group',
                                                      'strat_variable_name' = 'promoter:inhibitor_ratio',
                                                      'model_formula_left' = 'ratio',
                                                      'model_formula_right' = 'new_visit_num + (1|subject_id)',
                                                      'model_run_contrast_by' = 'new_visit_num'))

visit_stats <- lazy_girl_stats(stats_list = visit_stats_list)

visit_stats_res <- visit_stats$Results
visit_stats_contrast <- visit_stats$Contrasts

## only pulling significant results so I can visually parse them 
signifOnly_visit_stats_contrast <- visit_stats_contrast %>% 
  filter(signif != 'ns')
```

for stats between visits
```{r}
visit_key_value <- c('new_visit_num1' = 'Screen',
                     'new_visit_num2' = 'D1',
                     'new_visit_num3' = 'D4',
                     'new_visit_num4' = 'D10-15/TOC',
                     'new_visit_num5' = 'D17',
                     'new_visit_num6' = 'D40')

group1_visit_col_order <- c('treatment_group', 
                            'strat_var',
                            'group1_label',
                            'group1',
                            'group2')

group2_visit_col_order <- c('treatment_group', 
                            'strat_var',
                            'group1_label',
                            'group2_label',
                            'group1',
                            'group2')

signifOnly_visit_stats_contrast_wGroup1 <- values_to_keys(input_table = 
                                                                  signifOnly_visit_stats_contrast,
                                                                key_col_name = 'group1',
                                                                value_col_name = 'group1_label',
                                                                key_value_list = visit_key_value,
                                                                col_order_list = group1_visit_col_order)

signifOnly_visit_stats_contrast_wValues <- values_to_keys(input_table = 
                                                            signifOnly_visit_stats_contrast_wGroup1,
                                                                key_col_name = 'group2',
                                                                value_col_name = 'group2_label',
                                                                key_value_list = visit_key_value,
                                                                col_order_list = group2_visit_col_order)

```

linear mixed effects model with all variables 
- this doesn't show a ton since trends differ between bile acids and treatment groups over time and don't follow a linear curve
```{r}
state_nums <- c('conjugated' = 1,
                'deconjugated' = 2,
                'dehydroxylated' = 3)

treat_group_nums <- c("CRS3123_200mg_po_bid" = 1,
                      "CRS3123_400mg_po_bid" = 2,
                      "Vancomycin_125mg_po_bid" = 3)

visit_bins <- c('1' = 'before',
                '2' = 'before',
                '3' = 'during',
                '4' = 'during',
                '5' = 'after')

lm_summed_bileAcid_state_df <- summed_bileAcid_state_df %>% 
  mutate(visit_bins = visit_bins[new_visit_num],
         ba_state_num = as.numeric(state_nums[bile_acid_state]),
         treat_group_nums = as.numeric(treat_group_nums[treatment_group])) %>% 
  filter(bile_acid_state == "conjugated")

## refactoring visit bins so that adonis will compare other time points to toc 
lm_summed_bileAcid_state_df$visit_bins <- factor(lm_summed_bileAcid_state_df$visit_bins, levels = c("during",
                                                                        "before",
                                                                        # "during",
                                                                        "after"))

mixed_modeling(predictors = c('treat_group_nums', 'new_visit_num', 'visit_bins'),
               response_vars = c('summed_conc'),
               random_effect = '(1|subject_id)',
               data = lm_summed_bileAcid_state_df,
               reml = FALSE)
```

**Overall plot!**
making a giant plot of all bile acid concentrations
- can probably filter out:
  - beta muricholic acid
  - glycodeoxycholic acid
  - glycolithocholic acid 
  - taurodeoxycholic acid
  - taurolithocholic acid
  - tauroursodeoxycholic acid
```{r, fig.width=17, fig.height=17}
(bile_acidConc_plot <- apppleplots::facet_twice_plots(input_table = filt_bileAcid_meta_df,
                               x_value = 'new_visit_num',
                               y_value = 'log_ready_conc',
                               x_value_type = 'discrete',
                               y_transform = TRUE,
                               x_labels = visit_x_labs,
                               box_group_by = 'new_visit_num',
                               line_group_by = 'subject_id',
                               smooth_line = NULL,
                               point_alpha = 0,
                               facet_rows = 'bile_acid',
                               facet_cols = 'treatment_group',
                               row_labs = NULL,
                               col_labs = treat_group_labs,
                               x_name = 'Study Day',
                               y_name = 'Concentration (ug/g) (log10)',
                               title_content = 'Bile Acid Concentration Over Time') +
  scale_x_continuous(breaks = c(1, 2, 3, 4, 5, 6, 7),
                     labels = visit_x_labs) +
  # scale_y_continuous(trans = 'log10',
  #                    labels = scales::comma) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.4, size = 2.5) +
  theme_bw(base_size = 20) +
  theme(strip.text.y.right = element_text(angle = 0)))
```

**Individual plots**
per bile acid
- idk why dca is absolutely wiped out in the 200mg group but thats not great 
```{r, fig.width=13, fig.height=5}
## named list to rename the bile acids for the plot title 
bileAcid_name_list <- c("TaurocholicAcid_(TCA)" = "Taurocholic Acid (TCA)",
                        "TaurochenodeoxycholicAcid_(TCDCA)" = "Taurochenodeoxycholic Acid (TCDCA)",
                        "GlycocholicAcid_(GCA)" = "Glycocholic Acid (GCA)",
                        "GlycoursodeoxycholicAcid_(GUDCA)" = "Glycoursodeoxycholic Acid (GUDCA)",
                        "CholicAcid_(CA)" = "Cholic Acid (CA)",
                        "GlycochenodeoxycholicAcid_(GCDCA)" = "Glycochenodeoxycholic Acid (GCDCA)",
                        "UrsodeoxycholicAcid_(UDCA)" = "Ursodeoxycholic Acid (UDCA)",
                        "ChenodeoxycholicAcid_(CDCA)" = "Chenodeoxycholic Acid (CDCA)",
                        "DeoxycholicAcid_(DCA)" = "Deoxycholic Acid (DCA)",
                        "LithocholicAcid_(LCA)" = "Lithocholic Acid (LCA)")


for (wanted_bile_acid in unique(unlist(filt_bileAcid_meta_df['bile_acid']))) {
  ## filter table to only include wanted genus
  bile_acid_filt_df <- filt_bileAcid_meta_df %>% 
    filter(bile_acid == wanted_bile_acid)
  
  ## adding stats to plots 
  tmp_stats <- visit_stats_contrast %>% 
    rename(p.adj.signif = signif,
           bile_acid = strat_var) %>% 
    mutate(across(all_of(c('group1', 'group2')), ~gsub(., pattern = "new_visit_num*", replacement = "")),
           y.position = as.numeric(3),
           group1 = as.numeric(group1),
           group2 = as.numeric(group2)) %>% 
    filter(bile_acid == wanted_bile_acid)
  
  ## use genus to put together variable name and plot title so I can save it to that 
  variable_name <- paste(tolower(wanted_bile_acid), "only_plot", sep = "_")
  pdf_plot_file_name <- paste0("../plots/", tolower(wanted_bile_acid), "_only_plot.pdf")
  png_plot_file_name <- paste0("../plots/pngs/", tolower(wanted_bile_acid), "_only_plot.png")
  plot_title <- paste(bileAcid_name_list[[wanted_bile_acid]], "Concentration Over Time", sep = " ")
  
  int_plot <- apppleplots::facet_twice_plots(input_table = bile_acid_filt_df,
                               x_value = 'new_visit_num',
                               y_value = 'log_ready_conc',
                               x_value_type = 'discrete',
                               y_transform = TRUE,
                               x_labels = visit_x_labs,
                               box_group_by = 'new_visit_num',
                               line_group_by = 'subject_id',
                               smooth_line = NULL,
                               point_alpha = 0,
                               facet_rows = NULL,
                               facet_cols = 'treatment_group',
                               row_labs = NULL,
                               col_labs = treat_group_labs,
                               x_name = 'Study Day',
                               y_name = 'Concentration (ug/g) (log10)',
                               title_content = plot_title) +
        scale_x_continuous(breaks = c(1, 2, 3, 4, 5, 6, 7),
                           labels = visit_x_labs) +
        geom_jitter(width = 0.1, height = 0, alpha = 0.4, size = 2.5) +
        theme_bw(base_size = 20) +
        theme(strip.text.y.right = element_text(angle = 0)) +
        stat_pvalue_manual(tmp_stats,
                           step.group.by = c('treatment_group'),
                           step.increase = 0.1,
                           hide.ns = TRUE,
                           label = 'p.adj.signif')
  
  print(int_plot)
  
  ## assign plots to a variable so I can continue to look at them 
  assign(variable_name,
         int_plot)
  
  ## save plots to the plots subdirectory via ggsave (AS A PDF)
  ggsave(filename = pdf_plot_file_name,
         plot = int_plot,
         width = 13,
         height = 5)
  
  ## save plots to the plots/pngs/ subdirectory via ggsave (AS A PNG)
  ggsave(filename = png_plot_file_name,
         plot = int_plot,
         width = 13,
         height = 5)
}
```

**Bile acid bin plots**
```{r, fig.width=13, fig.height=9}
bin_plot_list <- list('state' = list('df' = summed_bileAcid_state_df,
                                     'filter_stats_val' = 'bile_acid_state',
                                     'y_pos_val' = 3.5,
                                     'title' = 'Bile Acid Pool Over Time by Degree of Transformation',
                                     'facet_labs' = state_labs,
                                     'height' = 9),
                      'class' = list('df' = summed_bileAcid_class_df,
                                     'filter_stats_val' = 'bile_acid_class',
                                     'y_pos_val' = 3.2,
                                     'title' = 'Bile Acid Pool Over Time by Type',
                                     'facet_labs' = class_labs,
                                     'height' = 7),
                      'cdiffEffect' = list('df' = summed_bileAcid_cdiffEffect_df,
                                           'filter_stats_val' = 'bile_acid_cdiffEffect',
                                           'y_pos_val' = 4,
                                           'title' = 'Bile Acid Pool Over Time by Effect on C. difficile Growth',
                                           'facet_labs' = cdiff_effect_labs,
                                           'height' = 7))

for (bin in unlist(names(bin_plot_list))){
  plot_name <- paste0('bile_acid_', bin, '_plot')
  facetRow_col_name <- paste0('bile_acid_', bin)
  pdf_plot_file_name <- paste0("../plots/bile_acid_", bin, "_plot.pdf")
  png_plot_file_name <- paste0("../plots/pngs/bile_acid_", bin, "_plot.png")
  
  ## adding stats to plots 
  tmp_stats <- visit_stats_contrast %>% 
    rename(p.adj.signif = signif,
           "{facetRow_col_name}" := strat_var) %>% 
    mutate(across(all_of(c('group1', 'group2')), ~gsub(., pattern = "new_visit_num*", replacement = "")),
           y.position = as.numeric(bin_plot_list[[bin]][['y_pos_val']]),
           group1 = as.numeric(group1),
           group2 = as.numeric(group2)) %>% 
    filter(stats_run == bin_plot_list[[bin]][['filter_stats_val']])

  tmp_plot <- apppleplots::facet_twice_plots(input_table = bin_plot_list[[bin]][['df']],
                               x_value = 'new_visit_num',
                               y_value = 'summed_conc',
                               x_value_type = 'discrete',
                               y_transform = TRUE,
                               x_labels = visit_x_labs,
                               box_group_by = 'new_visit_num',
                               line_group_by = 'subject_id',
                               smooth_line = NULL,
                               point_alpha = 0,
                               facet_rows = facetRow_col_name,
                               facet_cols = 'treatment_group',
                               row_labs = bin_plot_list[[bin]][['facet_labs']],
                               col_labs = treat_group_labs,
                               x_name = 'Study Day',
                               y_name = 'Concentration (ug/g) (log10)',
                               title_content = bin_plot_list[[bin]][['title']]) +
  scale_x_continuous(breaks = c(1, 2, 3, 4, 5, 6, 7),
                     labels = visit_x_labs) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.4, size = 2.5) +
  theme_bw(base_size = 20) +
  theme(strip.text.y.right = element_text(angle = 0)) +
  stat_pvalue_manual(tmp_stats,
                     step.group.by = c('treatment_group', facetRow_col_name),
                     step.increase = 0.1,
                     hide.ns = TRUE,
                     label = 'p.adj.signif')
    
    
  print(tmp_plot)
  
  assign(plot_name,
         tmp_plot)
  
  ## save plots to the plots subdirectory via ggsave (AS A PDF)
  ggsave(filename = pdf_plot_file_name,
         plot = tmp_plot,
         width = 13,
         height = bin_plot_list[[bin]][['height']])
  
  ## save plots to the plots/pngs/ subdirectory via ggsave (AS A PNG)
  ggsave(filename = png_plot_file_name,
         plot = tmp_plot,
         width = 13,
         height = bin_plot_list[[bin]][['height']])
}
```

**Bile acid ratio plots**
```{r, fig.width=13, fig.height=5}
ratio_plot_list <- list('primary:secondary' = list('df' = ratio_bileAcid_class_df,
                                                   'title' = 'Ratio of primary:secondary Bile Acids Over Time'),
                        'promotes:inhibits' = list('df' = ratio_bileAcid_cdiffEffect_df,
                                                   'title' = 'Ratio of promoter:inhibitor Bile Acids Over Time'))

for (ratio in unlist(names(ratio_plot_list))){
  plot_name <- paste0('bile_acid_', str_replace(ratio, "[:]", "_"), '_plot')
  pdf_plot_file_name <- paste0("../plots/", plot_name, ".pdf")
  png_plot_file_name <- paste0("../plots/pngs/", plot_name, ".png")
  plot_title <- paste0('Bile Acid ', ratio, ' Ratio Over Time')

  tmp_plot <- apppleplots::facet_twice_plots(input_table = ratio_plot_list[[ratio]][['df']],
                               x_value = 'new_visit_num',
                               y_value = 'ratio',
                               x_value_type = 'discrete',
                               y_transform = TRUE,
                               x_labels = visit_x_labs,
                               box_group_by = 'new_visit_num',
                               line_group_by = 'subject_id',
                               smooth_line = NULL,
                               point_alpha = 0,
                               facet_rows = NULL,
                               facet_cols = 'treatment_group',
                               row_labs = NULL,
                               col_labs = treat_group_labs,
                               x_name = 'Study Day',
                               y_name = 'Ratio (log10)',
                               title_content = ratio_plot_list[[ratio]][['title']]) +
  scale_x_continuous(breaks = c(1, 2, 3, 4, 5, 6, 7),
                     labels = visit_x_labs) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.4, size = 2.5) +
  theme_bw(base_size = 20) +
  theme(strip.text.y.right = element_text(angle = 0))
  
  print(tmp_plot)
  
  assign(plot_name,
         tmp_plot)
  
  ## save plots to the plots subdirectory via ggsave (AS A PDF)
  ggsave(filename = pdf_plot_file_name,
         plot = tmp_plot,
         width = 13,
         height = 5)
  
  ## save plots to the plots/pngs/ subdirectory via ggsave (AS A PNG)
  ggsave(filename = png_plot_file_name,
         plot = tmp_plot,
         width = 13,
         height = 5)
  
}
```

**Saving my outputs**
```{r}
save(list = c('bile_acidConc_plot',
                'bile_acid_cdiffEffect_plot',
                'bile_acid_class_plot',
                'bile_acid_state_plot',
                'bile_acid_primary_secondary_plot',
                'bile_acid_promotes_inhibits_plot',
                'lithocholicacid_(lca)_only_plot',
                'deoxycholicacid_(dca)_only_plot'),
       file = '../../../manuscript_figures/ggplots/bile_acid_plots.rdat')
```

```{r}
## overall bile acid plot (as a pdf)
ggsave('../plots/overall_bileAcid_plot.pdf',
       plot = bile_acidConc_plot,
       width = 17,
       height = 17)

## as a png
ggsave('../plots/pngs/overall_bileAcid_plot.png',
       plot = bile_acidConc_plot,
       width = 17,
       height = 17)

## stats 
write_tsv(signifOnly_visit_stats_contrast_wValues,
          file = '../stats/bileAcid_betweenVisits_postHoc.tsv')
write_tsv(signifOnly_treatGroup_stats_contrast_wValues,
          file = '../stats/bileAcid_betweenTreatGroups_postHoc.tsv')
```
