---
title: "scfa_analysis"
output: html_document
date: "2025-08-06"
author: "Madi Apgar"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
library(magrittr)
library(vroom)
library(apppleplots)
library(cgwtools)
library(ggpubr)
```

**Functions**
```{r}
source("../../art_lmer_functions.R")
source("../../diff_between_means_calc.R")

## small function to assign a value to a key in a dataframe column (i.e. numbers to actual categorical variables)
values_to_keys <- function(input_table,
                           key_col_name,
                           value_col_name,
                           key_value_list,
                           col_order_list){
  
  input_table[value_col_name] <- unlist(lapply(unlist(input_table[key_col_name]), 
                                        FUN = function(x){key_value_list[[x]]}))
  
  output_table <- input_table %>%
    select(all_of(col_order_list), everything())
  
  return(output_table)
}
```

**File paths**
```{r}
scfa_fp <- '../data/scfa_results.txt'
visit_key_fp <- '../data/visit_name_key.tsv'
meta_fp <- '../../../data_submission/1_16S_rRNA_qiita/CRS3123_fecal_samples/crestone_qiita_sample_information.tsv'

## global lists for renaming plot labels
visit_x_labs <- c('Screen',
                  'D1',
                  'D4',
                  'TOC',
                  'D17', 
                  'D40', 
                  'D70') 

treat_group_labs <- c('CRS3123 200mg',
                      'CRS3123 400mg',
                      'Vancomycin 125mg')
names(treat_group_labs) <- c('CRS3123_200mg_po_bid',
                             'CRS3123_400mg_po_bid',
                             'Vancomycin_125mg_po_bid')
```

**Reading in files**
```{r}
scfa_df <- read_tsv(scfa_fp)
visit_key_df <- read_tsv(visit_key_fp)
meta_df <- read_tsv(meta_fp) %>% 
  select(anonymized_name, host_subject_id, visit, visit_number, treatment_group) %>%
  rename(sampleid = anonymized_name,
         subject_id = host_subject_id,
         new_visit_label = visit,
         new_visit_num = visit_number) %>%
  ## add visit name key to combine w scfa data
  left_join(visit_key_df, by = "sampleid")
```

**Data wrangled**
first getting scfa df ready to be merged w the metadata so I can get sampleids
- need to double check that concentrations are already normalized based on dilution factor!!
```{r}
unwanted_cols <- c('UniqueID',
                   'BatchNum',
                   'RepNum',
                   'RecquisitionNumber',
                   'ACMUniqueID',
                   'ISStandard')

(proc_scfa_df <- scfa_df %>% 
  select(!unwanted_cols) %>% 
  mutate(Rerun = ifelse(is.na(Rerun), 0, 1),
         visit = paste(visit, "Cooled")) %>% 
  gather(contains("Acid"), key = "scfa", value = "concentration") %>% 
  rename(visit_name_key = visit) %>% 
  filter(!is.na(subject_id)))
```

then joining with metadata to get correct visits
- might filter out D40 since there are only a handful of samples from that day and it could be throwing the stats off
```{r}
(scfa_meta_df <- proc_scfa_df %>% 
  left_join(meta_df, by = c("subject_id", "visit_name_key")) %>% 
  filter(!is.na(concentration),
         new_visit_num != 6) %>% 
  mutate(log_ready_conc = (concentration + (min(concentration[concentration != 0])/2)),
         scfa = gsub(scfa, pattern = "-", replacement = "_")))
```
**Stats**
stratified by scfa and visit between treatment groups 
- wow there really isn't any significant differences at all...
```{r}
factored_scfa_meta_df <- scfa_meta_df %>% 
  convert_as_factor(subject_id, scfa, new_visit_num, treatment_group) %>% 
  mutate(log_conc = log10(log_ready_conc))

treatGroup_art_lmer_doubleStrat <- art_lmer_doubleStrat(input_table = factored_scfa_meta_df,
                                                       strat_variable1 = 'scfa',
                                                       strat_variable2 = 'new_visit_num',
                                                       model_formula_left = 'concentration',
                                                       model_formula_right = 'treatment_group',
                                                       model_contrast_run_by = 'treatment_group')

(treatGroup_doubleStrat_res <- treatGroup_art_lmer_doubleStrat$ARTResults)
(treatGroup_doubleStrat_contrast <- treatGroup_art_lmer_doubleStrat$ARTContrast)
```

stratified by treatment group and scfa between visits
- there are some differences in how hard the concentrations of scfas are hit by the antibiotics..need to investigate more though 
```{r}
visit_art_lmer_doubleStrat <- art_lmer_doubleStrat(input_table = factored_scfa_meta_df,
                                                       strat_variable1 = 'treatment_group',
                                                       strat_variable2 = 'scfa',
                                                       model_formula_left = 'log_conc',
                                                       model_formula_right = 'new_visit_num + (1|subject_id)',
                                                       model_contrast_run_by = 'new_visit_num')

(visit_doubleStrat_res <- visit_art_lmer_doubleStrat$ARTResults)
(visit_doubleStrat_contrast <- visit_art_lmer_doubleStrat$ARTContrast)

## only pulling significant results so I can visually parse them 
rename_stat_cols <- c('group1',
                      'group2')

signifOnly_visit_doubleStrat_contrast <- visit_doubleStrat_contrast %>% 
  filter(signif != 'ns') %>% 
  mutate(across(all_of(rename_stat_cols), ~gsub(pattern = "new_visit_num", replacement = "", .)))
```

stratified by treatment group and visit between scfas
- had to filter out d40 because there weren't enough data points to compare (may just filter it out overall)
```{r}
noD40_scfa_meta_df <- factored_scfa_meta_df %>% 
  filter(new_visit_num != "6")

scfaConc_art_lmer_doubleStrat <- art_lmer_doubleStrat(input_table = noD40_scfa_meta_df,
                                                       strat_variable1 = 'treatment_group',
                                                       strat_variable2 = 'new_visit_num',
                                                       model_formula_left = 'concentration',
                                                       model_formula_right = 'scfa',
                                                       model_contrast_run_by = 'scfa')

(scfaConc_doubleStrat_res <- scfaConc_art_lmer_doubleStrat$ARTResults)
(scfaConc_doubleStrat_contrast <- scfaConc_art_lmer_doubleStrat$ARTContrast)
```

**Plot!**
making an overall plot of scfa results over time
- idk how I feel about this since a lot of the trends were erased by log10 but there are also wildly different concentrations 
- using my plots function bc i'm lazy
```{r, fig.width=15, fig.height=15}
(scfaConc_plot <- apppleplots::facet_twice_plots(input_table = scfa_meta_df,
                               x_value = 'new_visit_num',
                               y_value = 'log_ready_conc',
                               x_value_type = 'discrete',
                               y_transform = TRUE,
                               x_labels = visit_x_labs,
                               box_group_by = 'new_visit_num',
                               line_group_by = 'subject_id',
                               point_alpha = 0,
                               facet_rows = 'scfa',
                               facet_cols = 'treatment_group',
                               row_labs = NULL,
                               col_labs = treat_group_labs,
                               x_name = 'Study Day',
                               y_name = 'Concentration (mL/g) (log10)',
                               title_content = 'SCFA Concentration Over Time') +
  scale_x_continuous(breaks = c(1, 2, 3, 4, 5, 6, 7),
                     labels = visit_x_labs) +
  # scale_y_continuous(trans = 'log10',
  #                    labels = scales::comma) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.4, size = 2.5) +
  theme_bw(base_size = 20) +
  theme(strip.text.y.right = element_text(angle = 0)))
```

**mini plot for paper**
only has acetate, butyrate, and propionate
```{r, fig.width=13, fig.height=10}
## making a mini dataframe 
mini_scfas <- c('AceticAcid',
                'n_ButanoicAcid',
                'PropanoicAcid')
mini_scfa_labs <- c('AceticAcid' = 'Acetate',
                    'n_ButanoicAcid' = 'Butyrate',
                    'PropanoicAcid' = 'Propionate')

mini_scfa_meta_df <- scfa_meta_df %>% 
  filter(scfa %in% mini_scfas)

## stats for plot
mini_scfa_stats <- visit_doubleStrat_contrast %>% 
    rename(p.adj.signif = signif) %>% 
    filter(scfa %in% mini_scfas) %>% 
    mutate(across(all_of(c('group1', 'group2')), ~gsub(., pattern = "new_visit_num*", replacement = "")),
           y.position = case_when(
             scfa == "AceticAcid" ~ 3.7,
             scfa == "n_ButanoicAcid" ~ 3.2,
             scfa == "PropanoicAcid" ~ 3.3
           ),
           y.position = as.numeric(y.position),
           group1 = as.numeric(group1),
           group2 = as.numeric(group2))

## plot  
(mini_scfaConc_plot <- apppleplots::facet_twice_plots(input_table = mini_scfa_meta_df,
                               x_value = 'new_visit_num',
                               y_value = 'log_ready_conc',
                               x_value_type = 'discrete',
                               y_transform = TRUE,
                               x_labels = visit_x_labs,
                               box_group_by = 'new_visit_num',
                               line_group_by = 'subject_id',
                               point_alpha = 0,
                               facet_rows = 'scfa',
                               facet_cols = 'treatment_group',
                               row_labs = mini_scfa_labs,
                               col_labs = treat_group_labs,
                               x_name = 'Study Day',
                               y_name = 'Concentration (mL/g) (log10)',
                               title_content = 'SCFA Concentration Over Time') +
  scale_x_continuous(breaks = c(1, 2, 3, 4, 5, 6, 7),
                     labels = visit_x_labs) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.4, size = 2.5) +
  theme_bw(base_size = 20) +
  facet_grid(rows = vars(scfa),
             cols = vars(treatment_group),
             scales = "free_y",
             labeller = labeller(scfa = mini_scfa_labs,
                                 treatment_group = treat_group_labs)) +
  theme(strip.text.y.right = element_text(angle = 0)) +
  stat_pvalue_manual(mini_scfa_stats,
                     step.group.by = c('treatment_group', 'scfa'),
                     step.increase = 0.1,
                     hide.ns = TRUE,
                     label = 'p.adj.signif'))
```


**individual plots**
- this for loop will save the plots as a .pdf via ggsave 
```{r, fig.height=6, fig.width=13}
## named list to rename the scfas for the plot title 
scfa_name_list <- c('2_methyl_ButanoicAcid' = '2-methyl-Butanoic Acid',
                    '2_methyl_propanoicAcid' = '2-methyl-Propanoic Acid',
                    'AceticAcid' = 'Acetic Acid',
                    'IsopentanoicAcid' = 'Isopentanoic Acid',
                    'n_ButanoicAcid' = 'n-Butanoic Acid',
                    'PentanoicAcid' = 'Pentanoic Acid',
                    'PropanoicAcid' = 'Propanoic Acid')

scfa_y_pos_list <- c('2_methyl_ButanoicAcid' = '2',
                    '2_methyl_propanoicAcid' = '2.5',
                    'AceticAcid' = '3.5',
                    'IsopentanoicAcid' = '1',
                    'n_ButanoicAcid' = '3.1',
                    'PentanoicAcid' = '2.2',
                    'PropanoicAcid' = '3.2')

for (wanted_scfa in unique(unlist(scfa_meta_df['scfa']))) {
  ## filter table to only include wanted genus
  scfa_filt_df <- scfa_meta_df %>% 
    filter(scfa == wanted_scfa)
  
  ## adding stats to plots 
  tmp_stats <- visit_doubleStrat_contrast %>% 
    rename(p.adj.signif = signif) %>% 
    mutate(across(all_of(c('group1', 'group2')), ~gsub(., pattern = "new_visit_num*", replacement = "")),
           y.position = as.numeric(scfa_y_pos_list[[wanted_scfa]]),
           group1 = as.numeric(group1),
           group2 = as.numeric(group2)) %>% 
    filter(scfa == wanted_scfa)
  
  ## use genus to put together variable name and plot title so I can save it to that 
  variable_name <- paste(tolower(wanted_scfa), "only_plot", sep = "_")
  pdf_plot_file_name <- paste0("../plots/", tolower(wanted_scfa), "_only_plot.pdf")
  png_plot_file_name <- paste0("../plots/pngs/", tolower(wanted_scfa), "_only_plot.png")
  plot_title <- paste(scfa_name_list[[wanted_scfa]], "Concentration Over Time", sep = " ")
  
  int_plot <- apppleplots::facet_twice_plots(input_table = scfa_filt_df,
                               x_value = 'new_visit_num',
                               y_value = 'log_ready_conc',
                               x_value_type = 'discrete',
                               y_transform = TRUE,
                               x_labels = visit_x_labs,
                               box_group_by = 'new_visit_num',
                               line_group_by = 'subject_id',
                               point_alpha = 0,
                               facet_rows = NULL,
                               facet_cols = 'treatment_group',
                               row_labs = NULL,
                               col_labs = treat_group_labs,
                               x_name = 'Study Day',
                               y_name = 'Concentration (mL/g) (log10)',
                               title_content = plot_title) +
        scale_x_continuous(breaks = c(1, 2, 3, 4, 5, 6, 7),
                           labels = visit_x_labs) +
        geom_jitter(width = 0.1, height = 0, alpha = 0.4, size = 2.5) +
        theme_bw(base_size = 20) +
        theme(strip.text.y.right = element_text(angle = 0)) +
        stat_pvalue_manual(tmp_stats,
                           step.group.by = c('treatment_group'),
                           step.increase = 0.1,
                           hide.ns = TRUE,
                           label = 'p.adj.signif')
  
  print(int_plot)
  
  ## assign plots to a variable so I can continue to look at them 
  assign(variable_name,
         int_plot)
  
  ## save plots to the plots subdirectory via ggsave (AS A PDF)
  ggsave(filename = pdf_plot_file_name,
         plot = int_plot,
         width = 13,
         height = 5)
  
  ## save plots to the plots subdirectory via ggsave (AS A PNG)
  ## this is bc powerpoint and google slides are stupid
  ggsave(filename = png_plot_file_name,
         plot = int_plot,
         width = 13,
         height = 5)
  
}
```

**Saving my outputs**
```{r}
save(list = c('aceticacid_only_plot',
              'propanoicacid_only_plot',
              'n_butanoicacid_only_plot',
              'mini_scfaConc_plot'),
       file = '../../../manuscript_figures/ggplots/scfa_plots.rdat')
```

saving the pre-wrangled scfa file for correlation comparisons
```{r}
write_tsv(mini_scfa_meta_df,
          file = '../data/proc_scfa_results.tsv')
```

```{r}
## as a pdf
ggsave('../plots/overall_scfa_plot.pdf',
       plot = scfaConc_plot,
       width = 15,
       height = 15)

## as a png
ggsave('../plots/pngs/overall_scfa_plot.png',
       plot = scfaConc_plot,
       width = 15,
       height = 15)

## stats
write_tsv(treatGroup_doubleStrat_contrast,
          file = '../stats/scfa_visit_bt_treatGroup.tsv')
write_tsv(signifOnly_visit_doubleStrat_contrast,
          file = '../stats/treatGroup_scfa_bt_visit.tsv')
```
