---
title: "generate_tax_biplot"
output: html_document
date: "2025-10-01"
author: "Madi Apgar"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(tidytext)
library(dplyr)
library(magrittr)
library(vroom)
library(vegan)
library(phyloseq)
library(ape)
library(viridis)
library(ggrepel)
library(ggh4x)
library(ggpubr)

set.seed(5)
```

**Functions**
```{r}
## break this up into two functions?? idk
calculate_my_biplot_vals <- function(otu_table,
                                     tax_table,
                                     nwk_tree,
                                     weighted_dist = FALSE,
                                     top_tax_n,
                                     lines_longer_by,
                                     metadata_table,
                                     sample_col_name){
  ## converting taxonomy to phyloseq object
  ## has to be a matrix or phyloseq loses its mind
  bug_tax <- tax_table(as.matrix(tax_table))
  
  ## converting otu table to phyloseq object
  ## taxa are rows, samples are cols 
  ## tax abundances need to be transformed to relative abundances 
  ## also making sure that the same sampleids are included in the otu_table as the metadata
  filt_proc_meta <- filter(metadata_table, metadata_table[[sample_col_name]] %in% colnames(otu_table))
  filt_otu_table <- select(otu_table, filt_proc_meta[[sample_col_name]])
  bug_otu <- otu_table(filt_otu_table, taxa_are_rows = TRUE)
  
  ## creating phyloseq object of tax, otus, and tree to calculate unifrac distances
  bug_phyloseq <- phyloseq(bug_otu, nwk_tree, bug_tax)
  
  ## calculating unifrac distances 
  bug_dist <- phyloseq::UniFrac(bug_phyloseq, weighted = weighted_dist)
  bug_dist_mat <- as.matrix(bug_dist)
  
  ## calculating pcoa of distances 
  bug_pcoa <- ape::pcoa(bug_dist_mat)
  
  ## pulling  proportion of variance explained for each axis (only positive relative eigenvalues)
  ## only positive eigenvalues have pcoa vector calculations 
  eigenvals <- bug_pcoa$values$Relative_eig
  pos_eigenvals <- eigenvals[eigenvals > 0]
  
  ## putting together axis labels for plots
  proportion_exp_df <- as.data.frame(pos_eigenvals) %>% 
    rownames_to_column(var = "axis") %>% 
    rename(proportion_explained = pos_eigenvals) %>% 
    mutate(axis_label = paste0('PC', axis, ' - ', round(proportion_explained*100, 2), '%'))
  
  ## saving axis labels to a list so can easily be referenced 
  axis_labs <- proportion_exp_df$axis_label
  
  ## pulling out pcoa axis values and setting otu table sampleids as rows 
  bug_pcoa_vals <- bug_pcoa$vectors
  transposed_bug_otu <- as.matrix(t(filt_otu_table))
  
  ## calculating covariance between pcoa values and otu table counts 
  covariance <- cov(bug_pcoa_vals, transposed_bug_otu)
  
  ## adding taxonomy back up with covariances
  tax_key_df <- as.data.frame(tax_table)
  tax_key_df['tax_id'] <- rownames(tax_key_df)
  
  ## calculating biplot lines between the first 4 pc axes wanted (1 to 2, 2 to 3, etc)
  biplot_tax_dist <- covariance %>% 
    as_tibble(rownames = "axis") %>% 
    gather(-axis, key = "tax_id", value = "covariance") %>% 
    spread(key = "axis", value = "covariance") %>% 
    mutate(dist_origin1to2 = sqrt(Axis.1^2 + Axis.2^2),
           dist_origin2to3 = sqrt(Axis.2^2 + Axis.3^2),
           dist_origin3to4 = sqrt(Axis.3^2 + Axis.4^2))
  
  ## pulling the distances for the top n tax given by the user for all axes in one table 
  ## so function only has top be run once
  top_tax_biplot_lines <- tibble()
  dist_col_list <- c('dist_origin1to2', 'dist_origin2to3', 'dist_origin3to4')
  for (dist_calc in dist_col_list){
    biplot_lines <- biplot_tax_dist %>% 
      top_n(top_tax_n, .data[[dist_calc]]) %>% 
      select(tax_id, Axis.1, Axis.2, Axis.3, Axis.4) %>% 
      mutate(top_tax_axis = paste(dist_calc),
             across(contains('Axis.'), ~(. *0.4))) %>% 
      left_join(tax_key_df, by = "tax_id")
    
    top_tax_biplot_lines <- rbind(top_tax_biplot_lines, biplot_lines)
  }
  
  ## joining pcoa values with metadata for plotting 
  biplot_points <- bug_pcoa_vals %>% 
    as_tibble(rownames = sample_col_name) %>% 
    select(all_of(sample_col_name), Axis.1, Axis.2, Axis.3, Axis.4) %>% 
    left_join(filt_proc_meta, by = sample_col_name)
  
  ## list of outputs 
  my_list <- list(PCvar = axis_labs,
                  DFPoints = biplot_points,
                  DFLines = top_tax_biplot_lines,
                  DistMat = bug_dist_mat)
  return(my_list)
}


## function for making the actual plot
make_my_biplot <- function(x_axis,
                           y_axis,
                           points_df,
                           points_color,
                           points_id, 
                           color_palette,
                           color_name,
                           color_labels,
                           facet_by,
                           facet_labels,
                           facet_scales = "fixed",
                           facet_rows = 1,
                           lines_df,
                           lines_label,
                           lines_label_size,
                           ellipse = TRUE,
                           x_name,
                           y_name,
                           title_content){
         plot <- ggplot() +
                  geom_jitter(data = points_df,
                             aes(x = .data[[x_axis]], y = .data[[y_axis]], 
                                 fill = .data[[points_color]]),
                             size = 5,
                             alpha = 0.8,
                             pch = 21) +
                 geom_line(data = points_df,
                           aes(x = .data[[x_axis]], y = .data[[y_axis]],
                               group = .data[[points_id]]),
                           alpha = 0.3) +
                 scale_fill_manual(values = color_palette,
                                    name = color_name,
                                    labels = color_labels) +
                  facet_wrap(~.data[[facet_by]],
                             labeller = labeller(.cols = facet_labels),
                             scales = facet_scales,
                             nrow = facet_rows) +
                  theme_bw(base_size = 20) +
                  geom_segment(data = lines_df,
                               aes(x = 0, xend = .data[[x_axis]], y = 0, yend = .data[[y_axis]]),
                               arrow = arrow(length = unit(0.3, "cm"))) +
                  geom_text_repel(data = lines_df,
                                  aes(x = .data[[x_axis]], y = .data[[y_axis]], label = .data[[lines_label]]),
                                  color = 'black',
                                  position = 'jitter',
                                  size = lines_label_size,
                                  fontface = 'bold') +
                  labs(x = x_name,
                       y = y_name,
                       title = title_content)
         
         ## only add ellipses to the plot if input is true 
         if (isTRUE(ellipse)){
           plot <- plot +
                     stat_ellipse(data = points_df,
                                  geom = "polygon",
                                  alpha = 0.15,
                                  aes(x = .data[[x_axis]], y = .data[[y_axis]], fill = .data[[points_color]]))
         }
         
  return(plot)
}
```

**File paths**
```{r}
## publicly available online through metaphlan resources
metaphlan_tree_fp <- '../references/mpa_vJun23_CHOCOPhlAnSGB_202307.nwk'
## not provided due to file size
bugs_list_fp <- '../data/all_bugs_list.tsv'
## use the sample metadata from qiita
meta_fp <- '../../../data_submission/1_16S_rRNA_qiita/CRS3123_fecal_samples/crestone_qiita_sample_information.tsv'

## global variables for plot labels
treat_group_labs <- c('CRS3123_200mg_po_bid' = 'CRS3123 200mg',
                      'CRS3123_400mg_po_bid' = 'CRS3123 400mg',
                      'Vancomycin_125mg_po_qid' = 'Vancomycin 125mg')

## taking out subject samples that had a screen but not D1 sample (123-1001)
take_out_samples <- c('123-1001.1', '123-1001.4')

## putting together custom color palette based on viridis
# viridis_pal(option = "C", begin = 0, end = 1)(3)
my_viridis_c_palette <- c("#0D0887FF", "#CC4678FF")
```

**Reading in files**
- subject 123-1001 has screen and TOC samples, need to filter out since everyone else has D1 (not sure how this happened?)
```{r}
bugs_list <- read_tsv(bugs_list_fp) %>% 
  filter(!(sampleid %in% take_out_samples))
tree <- read_tree(metaphlan_tree_fp)
meta <- read_tsv(meta_fp)
```

**Data wrangled**
making bugs list an otu table
```{r}
## go down to speciesLevel_genome_bin - all relative abundances add up to 100 per sampleid
## which is what we want (I think)
wide_tax <- bugs_list %>% 
  filter(!is.na(speciesLevel_genome_bin)) %>% 
  select(!c('relative_abundance', 'additional_species')) %>% 
  mutate(test = gsub(paste0("|", NCBI_tax_id), pattern = "\\|\\|+$", replacement = "|"),
         tax_id = str_extract(test, "(?<=\\|)[^|]+(?=\\|[^|]*$)")) %>% 
  select(tax_id, everything()) %>% 
  select(!c('NCBI_tax_id', 'test', 'sampleid', 'clade_name', 'speciesLevel_genome_bin')) %>% 
  distinct(tax_id, .keep_all = TRUE) %>% 
  column_to_rownames(var = "tax_id")
```

```{r}
sample_list <- unique(bugs_list$sampleid)

## may have issues with duplicates since its broken down at every taxonomic level 
## filled all NAs with 0, idk if that's right either
wide_bug_otu <- bugs_list %>% 
  filter(!is.na(speciesLevel_genome_bin)) %>% 
  select(clade_name, NCBI_tax_id, sampleid, relative_abundance) %>% 
  spread(key = 'sampleid', value = 'relative_abundance', fill = 0) %>% 
  group_by(NCBI_tax_id) %>% 
  summarise(across(all_of(sample_list), sum)) %>% 
  mutate(test = gsub(paste0("|", NCBI_tax_id), pattern = "\\|\\|+$", replacement = "|"),
         tax_id = str_extract(test, "(?<=\\|)[^|]+(?=\\|[^|]*$)")) %>% 
  select(tax_id, everything()) %>% 
  select(!c('NCBI_tax_id', 'test')) %>% 
  column_to_rownames(var = "tax_id")

# taxa_names(taxonomy)
```

metadata processing
```{r}
proc_meta <- meta %>%
  select(anonymized_name, host_subject_id, visit, visit_number, treatment_group) %>% 
  rename(sampleid = anonymized_name) %>% 
  ## matching sampleids with the shotgun data since not all samples were included 
  filter(sampleid %in% colnames(wide_bug_otu))
```

creating releveled metadata for adonis testing 
```{r}
fact_meta <- proc_meta

## releveling to compare crs to vanc 
fact_meta$treatment_group <- factor(fact_meta$treatment_group, levels = c(
  "Vancomycin_125mg_po_qid",
  "CRS3123_200mg_po_bid",
  "CRS3123_400mg_po_bid"))

## comparing everything to D1 
fact_meta$visit_number <- factor(fact_meta$visit_number, levels = c(
  "2",
  "4"))
```

**Calculating UniFrac distances**
unweighted - stratified by treatment group 
```{r}
uu_adonis_strat_treatGroup <- tibble()
uu_lines_strat_treatGroup <- tibble()
uu_points_strat_treatGroup <- tibble()

for (treatment in unique(unlist(fact_meta$treatment_group))){
  mini_meta <- filter(fact_meta, treatment_group == treatment)
  
  uu_treatment_biplot <- calculate_my_biplot_vals(otu_table = wide_bug_otu,
                                                  tax_table = wide_tax,
                                                  nwk_tree = tree,
                                                  weighted_dist = FALSE,
                                                  top_tax_n = 10,
                                                  metadata_table = mini_meta,
                                                  sample_col_name = 'sampleid')
  ## running adonis test
  tmp_adonis_res <- adonis2(as.dist(uu_treatment_biplot$DistMat) ~ visit_number,
                            data = mini_meta, 
                            permutations = 999,
                            strata = mini_meta$host_subject_id,
                            by = "terms")
  
  uu_adonis_res <- tidy(tmp_adonis_res) %>% 
    mutate(treatment_group = paste(treatment))
  
  uu_treatment_lines <- uu_treatment_biplot$DFLines %>% 
    mutate(treatment_group = paste(treatment))
  uu_treatment_points <- uu_treatment_biplot$DFPoints %>% 
    mutate(treatment_group = paste(treatment))
  
  uu_adonis_strat_treatGroup <- rbind(uu_adonis_strat_treatGroup, uu_adonis_res)
  uu_lines_strat_treatGroup <- rbind(uu_lines_strat_treatGroup, uu_treatment_lines)
  uu_points_strat_treatGroup <- rbind(uu_points_strat_treatGroup, uu_treatment_points)
}

uu_adonis_strat_treatGroup$dist <- 'unweighted_unifrac'
```

unweighted - all treatment groups together to get overall variance explained by an axis 
```{r}
uu_biplot <- calculate_my_biplot_vals(otu_table = wide_bug_otu,
                                      tax_table = wide_tax,
                                      nwk_tree = tree,
                                      weighted_dist = FALSE,
                                      top_tax_n = 10,
                                      metadata_table = proc_meta,
                                      sample_col_name = 'sampleid')
uu_axis_labs <- uu_biplot$PCvar
uu_biplot_lines <- uu_biplot$DFLines
uu_biplot_points <- uu_biplot$DFPoints
uu_biplot_dist_mat <- uu_biplot$DistMat
```

unweighted - overall adonis test 
```{r}
uu_biplot_adonis_overall <- adonis2(as.dist(uu_biplot_dist_mat) ~ visit_number * treatment_group,
                                    data = fact_meta, 
                                    permutations = 999,
                                    strata = fact_meta$host_subject_id,
                                    by = "terms")

(uu_biplot_adonis_overall_df <- tidy(uu_biplot_adonis_overall) %>% 
  mutate(dist = 'unweighted_unifrac'))
```

unweighted unifrac biplots
```{r, fig.width=25, fig.height=8}
(shotgun_uu_biplot_1to2 <- make_my_biplot(x_axis = 'Axis.1',
                                 y_axis = 'Axis.2',
                                 points_df = uu_points_strat_treatGroup,
                                 points_color = 'visit',
                                 points_id = 'host_subject_id',
                                 color_palette = my_viridis_c_palette,
                                 color_name = 'Visit',
                                 color_labels = c('D1', 'TOC'),
                                 facet_by = 'treatment_group',
                                 facet_labels = treat_group_labs,
                                 facet_scales = "free",
                                 facet_rows = 1,
                                 lines_df = filter(uu_lines_strat_treatGroup, top_tax_axis == 'dist_origin1to2'),
                                 lines_label = 'species',
                                 lines_label_size = 5.5,
                                 ellipse = TRUE,
                                 x_name = uu_axis_labs[1],
                                 y_name = uu_axis_labs[2],
                                 title_content = 'PC1-PC2: Unweighted UniFrac Biplot'))

(shotgun_uu_biplot_2to3 <- make_my_biplot(x_axis = 'Axis.2',
                                 y_axis = 'Axis.3',
                                 points_df = uu_points_strat_treatGroup,
                                 points_color = 'visit',
                                 points_id = 'host_subject_id',
                                 color_palette = my_viridis_c_palette,
                                 color_name = 'Visit',
                                 color_labels = c('D1', 'TOC'),
                                 facet_by = 'treatment_group',
                                 facet_labels = treat_group_labs,
                                 facet_scales = "free",
                                 facet_rows = 1,
                                 lines_df = filter(uu_lines_strat_treatGroup, top_tax_axis == 'dist_origin2to3'),
                                 lines_label = 'species',
                                 lines_label_size = 5.5,
                                 ellipse = TRUE,
                                 x_name = uu_axis_labs[2],
                                 y_name = uu_axis_labs[3],
                                 title_content = 'PC2-PC3: Unweighted UniFrac Biplot'))
```

weighted - stratified by treatment group 
```{r}
wu_adonis_strat_treatGroup <- tibble()
wu_lines_strat_treatGroup <- tibble()
wu_points_strat_treatGroup <- tibble()

for (treatment in unique(unlist(fact_meta$treatment_group))){
  mini_meta <- filter(fact_meta, treatment_group == treatment)
  
  wu_treatment_biplot <- calculate_my_biplot_vals(otu_table = wide_bug_otu,
                                                  tax_table = wide_tax,
                                                  nwk_tree = tree,
                                                  weighted_dist = TRUE,
                                                  top_tax_n = 10,
                                                  metadata_table = mini_meta,
                                                  sample_col_name = 'sampleid')
  ## running adonis test
  tmp_adonis_res <- adonis2(as.dist(wu_treatment_biplot$DistMat) ~ visit_number,
                            data = mini_meta, 
                            permutations = 999,
                            strata = mini_meta$host_subject_id,
                            by = "terms")
  
  wu_adonis_res <- tidy(tmp_adonis_res) %>% 
    mutate(treatment_group = paste(treatment))
  
  wu_treatment_lines <- wu_treatment_biplot$DFLines %>% 
    mutate(treatment_group = paste(treatment))
  wu_treatment_points <- wu_treatment_biplot$DFPoints %>% 
    mutate(treatment_group = paste(treatment))
  
  wu_adonis_strat_treatGroup <- rbind(wu_adonis_strat_treatGroup, wu_adonis_res)
  wu_lines_strat_treatGroup <- rbind(wu_lines_strat_treatGroup, wu_treatment_lines)
  wu_points_strat_treatGroup <- rbind(wu_points_strat_treatGroup, wu_treatment_points)
}

wu_adonis_strat_treatGroup$dist <- 'weighted_unifrac'
```

weighted - all treatment groups together to get overall variance explained by an axis 
```{r}
wu_biplot <- calculate_my_biplot_vals(otu_table = wide_bug_otu,
                                      tax_table = wide_tax,
                                      nwk_tree = tree,
                                      weighted_dist = TRUE, ## change to TRUE for weighted unifrac calc
                                      top_tax_n = 10,
                                      metadata_table = proc_meta,
                                      sample_col_name = 'sampleid')
wu_axis_labs <- wu_biplot$PCvar
wu_biplot_lines <- wu_biplot$DFLines
wu_biplot_points <- wu_biplot$DFPoints
wu_biplot_dist_mat <- wu_biplot$DistMat
```

weighted - overall adonis test 
```{r}
## shouldnt need to relevel metadata since ive already done that above 
wu_biplot_adonis_overall <- adonis2(as.dist(wu_biplot_dist_mat) ~ visit_number * treatment_group,
                                    data = fact_meta, 
                                    permutations = 999,
                                    strata = fact_meta$host_subject_id,
                                    by = "terms")

(wu_biplot_adonis_overall_df <- tidy(wu_biplot_adonis_overall) %>% 
  mutate(dist = 'weighted_unifrac'))
```

weighted unifrac biplots
```{r, fig.width=25, fig.height=8}
(shotgun_wu_biplot_1to2 <- make_my_biplot(x_axis = 'Axis.1',
                                  y_axis = 'Axis.2',
                                  points_df = wu_points_strat_treatGroup,
                                  points_color = 'visit',
                                  points_id = 'host_subject_id',
                                  color_palette = my_viridis_c_palette,
                                  color_name = 'Visit',
                                  color_labels = c('D1', 'TOC'),
                                  facet_by = 'treatment_group',
                                  facet_labels = treat_group_labs,
                                  facet_scales = "free",
                                  facet_rows = 1,
                                  lines_df = filter(wu_lines_strat_treatGroup, top_tax_axis == 'dist_origin1to2'),
                                  lines_label = 'species',
                                  lines_label_size = 5.5,
                                  ellipse = TRUE,
                                  x_name = wu_axis_labs[1],
                                  y_name = wu_axis_labs[2],
                                  title_content = 'PC1-PC2: Weighted UniFrac Biplot'))

(shotgun_wu_biplot_2to3 <- make_my_biplot(x_axis = 'Axis.2',
                                  y_axis = 'Axis.3',
                                  points_df = wu_points_strat_treatGroup,
                                  points_color = 'visit',
                                  points_id = 'host_subject_id',
                                  color_palette = my_viridis_c_palette,
                                  color_name = 'Visit',
                                  color_labels = c('D1', 'TOC'),
                                  facet_by = 'treatment_group',
                                  facet_labels = treat_group_labs,
                                  facet_scales = "free",
                                  facet_rows = 1,
                                  lines_df = filter(wu_lines_strat_treatGroup, top_tax_axis == 'dist_origin2to3'),
                                  lines_label = 'species',
                                  lines_label_size = 5.5,
                                  ellipse = TRUE,
                                  x_name = wu_axis_labs[2],
                                  y_name = wu_axis_labs[3],
                                  title_content = 'PC2-PC3: Weighted UniFrac Biplot'))
```

**calculating alpha rarefaction curve**
- need to take it out of relative abundance to just raw counts for vegan::rarecurve()
```{r}
## making sampleids rows and taxa cols
test <- as.data.frame(t(wide_bug_otu)) 
## calculating total for each sampleid (taking it out of rel abun step 1)
test$RowTotal <- apply(test, MARGIN = 1, FUN = sum, na.rm = TRUE)
## saving those values
multiplier_col <- test$RowTotal
## taking out total abun col
test <- test[, !(names(test) == 'RowTotal')]
## multiplying all rel abunds for a sample by the total abun (taking it out of rel abun step 2)
res_test <- apply(test, 1, "*", multiplier_col)
## rounding to nearest integer
round_res_test <- round(res_test)
## multiplication step transposes df so transposing it back so sampleids are rows
proc_res_test <- t(round_res_test)

## calculate alpha rarefaction curve to determine of sequencing depth covers microbial diversity of sample
vegan::rarecurve(proc_res_test, label = T)
```


**Saving my outputs**

**adonis2 permanvoa stat results**
```{r}
## overall adonis (takes into account time and treatment group with crs compared to vanc and toc compared to d1)
allDist_overall_adonis <- rbind(uu_biplot_adonis_overall_df,
                                wu_biplot_adonis_overall_df)

write_tsv(allDist_overall_adonis,
          file = '../stats/overall_unifrac_biplot_adonis.tsv')

## per treatment group adonis (takes into account time, comparing toc to d1 per treatment group)
allDist_strat_treatGroup_adonis <- rbind(uu_adonis_strat_treatGroup,
                                         wu_adonis_strat_treatGroup)

write_tsv(allDist_strat_treatGroup_adonis,
          file = '../stats/by_treatGroup_unifrac_biplot_adonis.tsv')
```

```{r}
save(list = c('shotgun_uu_biplot_1to2',
                          'shotgun_uu_biplot_2to3',
                          'shotgun_wu_biplot_1to2',
                          'shotgun_wu_biplot_2to3'),
                 file = '../../../manuscript_figures/ggplots/shotgun_biplots.rdat')
```

```{r}
ggsave('../plots/uu_biplot_1to2.pdf',
       plot = shotgun_uu_biplot_1to2,
       width = 25,
       height = 8)
ggsave('../plots/pngs/uu_biplot_1to2.png',
       plot = shotgun_uu_biplot_1to2,
       width = 25,
       height = 8)

ggsave('../plots/uu_biplot_2to3.pdf',
       plot = shotgun_uu_biplot_2to3,
       width = 25,
       height = 8)
ggsave('../plots/pngs/uu_biplot_2to3.png',
       plot = shotgun_uu_biplot_2to3,
       width = 25,
       height = 8)

ggsave('../plots/wu_biplot_1to2.pdf',
       plot = shotgun_wu_biplot_1to2,
       width = 25,
       height = 8)
ggsave('../plots/pngs/wu_biplot_1to2.png',
       plot = shotgun_wu_biplot_1to2,
       width = 25,
       height = 8)

ggsave('../plots/wu_biplot_2to3.pdf',
       plot = shotgun_wu_biplot_2to3,
       width = 25,
       height = 8)
ggsave('../plots/pngs/wu_biplot_2to3.png',
       plot = shotgun_wu_biplot_2to3,
       width = 25,
       height = 8)
```

