---
title: "beta_div"
output: html_document
date: "2025-02-18"
author: "Madi Apgar"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(magrittr)
library(broom)
library(qiime2R)
library(apppleplots)
library(vegan)
library(cgwtools)

set.seed(5)
```

## **Functions**
```{r}
###### plot functions ######
## 1 
## unweighted/weighted unifrac pcoa result, faith's pd, and shannon entropy file prep 
## going to attempt to return multiple outputs so I can just have one function for file prep
biom_table_prep <- function(unweighted_fp,
                            weighted_fp,
                            faith_fp,
                            metadata_file){
  ## unweighted pcoa
  unweighted <- read_qza(unweighted_fp)$data
  unweighted_var <- unweighted$ProportionExplained
  unweighted_pcoa <- unweighted$Vectors ##used for pcoa plot
  names(unweighted_pcoa)[names(unweighted_pcoa) == 'SampleID'] <- 'sampleid'
  
  ## weighted pcoa
  weighted <- read_qza(weighted_fp)$data
  weighted_var <- weighted$ProportionExplained
  weighted_pcoa <- weighted$Vectors
  names(weighted_pcoa)[names(weighted_pcoa) == 'SampleID'] <- 'sampleid'
  
  ## faith's 
  faith <- read_tsv(faith_fp)
  names(faith)[names(faith) == '#SampleID'] <- 'sampleid'
  
  ## unweighted biom 
  unweighted_pcoa %>% 
    left_join(metadata_file, by = 'sampleid') %>% 
    left_join(faith, by = 'sampleid') -> unweighted_biom
  
  ## weighted biom
  weighted_pcoa %>% 
    left_join(metadata_file, by = 'sampleid') %>% 
    left_join(faith, by = 'sampleid') -> weighted_biom
  
  ## creating a list to return multiple outputs 
  my_list <- list(UnweightedVar = unweighted_var, 
                  WeightedVar = weighted_var,
                  UnweightedBiom = unweighted_biom,
                  WeightedBiom = weighted_biom)
  return(my_list)
}

## 2
## this function will pull out the percent variations from a specified column so you can add it to your pcoa plots 
pcoa_ax_lab <- function(unifrac_var, col_name){
  uni_lab <- as.character(round(unifrac_var[col_name] * 100, 2))
  uni_lab <- paste0(col_name, ' - ', uni_lab, '%')
  return(uni_lab)
}

###### stats functions ######
## 1
## for distance matrix processing
## for beta diversity statistical analysis 
dist_matrix_prep <- function(metadata_file,
                             dist_matrix_fp){
  ## metadata
  metadata <- metadata_file
  ## distance matrix
  dist <- read_tsv(dist_matrix_fp)
  names(dist)[names(dist) == '...1'] <- 'sampleid'
  dist %>% 
    gather(-sampleid, key = sample_col, value = dist) %>% 
    filter(sampleid %in% metadata$sampleid) %>% 
    filter(sample_col %in% metadata$sampleid) %>% 
    spread(sample_col, dist) -> dist_long
  dist_long %>% 
    select(-sampleid) -> dist_proc
  metadata %>% 
    arrange(sampleid) -> metadata
  metadata %>% 
    filter(sampleid %in% dist_long$sampleid) -> filt_meta
  dist_proc <- as.matrix(dist_proc)
  row.names(dist_proc) <- colnames(dist_proc)
  filt_meta <- filt_meta[order(filt_meta$sampleid),]
  ## list of outputs
  my_list <- list(Metadata = filt_meta,
                  DistanceMatrix = dist_proc)
  return(my_list)
}


## 2
## beta diversity adonis2 testing function
adonis_test <- function(dist_matrix,
                        metadata_file,
                        formula_left,
                        formula_right){
  dist_matrix <- as.dist(dist_matrix)
  funky_formula <- paste(formula_left, formula_right, sep = "~")
  adonis_results <- adonis2(as.formula(funky_formula),
                            data = metadata_file,
                            permutations = 999, 
                            parallel = 4,
                            strata = metadata_file$subject_id,
                            by = "terms")
  adonis_results <- tidy(adonis_results)
  adonis_results['signif'] <- symnum(adonis_results$p.value,
                                     cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
                                     symbols = c("****", "***", "**", "*", "+", "ns"),
                                     abbr.colnames = FALSE,
                                     na = "")
  return(adonis_results)
}

## 3
## to run an adonis test by day relative to infection
adonis_per_day <- function(metadata_file,
                           wanted_col,
                           col_filter,
                           dist_fp,
                           formula_left,
                           formula_right){
  metadata_file %>% 
    filter(.data[[wanted_col]] == col_filter) -> col_filt_metadata
  dist_files <- dist_matrix_prep(col_filt_metadata,
                                 dist_fp)
  dist <- dist_files$DistanceMatrix
  filt_meta <- dist_files$Metadata
  adonis_results <- adonis_test(dist,
                                filt_meta,
                                formula_left,
                                formula_right)
  adonis_results[wanted_col] <- col_filter
  return(adonis_results)
}

## 4 
## function that will run a for loop to stratify the adonis test by a certain variable
adonis_for_loop <- function(metadata_file,
                            dist_matrix_fp,
                            strat_column,
                            formula_left,
                            formula_right){
  output <- list()
  for (i in unique(unlist(metadata_file[strat_column]))) {
    tmp_output <- adonis_per_day(metadata_file,
                                 strat_column,
                                 i,
                                 dist_matrix_fp,
                                 formula_left,
                                 formula_right)
    tmp_output[strat_column] <- i
    # when saving something as a named list in R, 
    # we need to use the double brackets
    # output[[as.character(i)]] <- tmp_output
    
    # if we're appending a higher dimensional object (df, matrix, vector,) 
    # to a list, we need to wrap it in a list() within the append function.
    output <- append(output, list(tmp_output))
  }
  output <- bind_rows(output)
  return(output)
}

```

## **File paths**
```{r}
uu_fp <- '../data/unweighted_unifrac_pcoa_results.qza'
wu_fp <- '../data/weighted_unifrac_pcoa_results.qza'
uu_dist_fp <- '../data/uw_dist_matrix.tsv'
wu_dist_fp <- '../data/w_dist_matrix.tsv'
faith_pd_fp <- '../data/hc_faith_pd.tsv'
comb_seq_meta_fp <- '../data/healthyControl_comb_metadata.tsv'
crestone_meta_fp <- '../../../../data_submission/1_16S_rRNA_qiita/CRS3123_fecal_samples/crestone_qiita_sample_information.tsv'

## global plot labels 
treat_group_labs <- c('CRS3123 200mg',
                      'CRS3123 400mg',
                      'Vancomycin 125mg')
names(treat_group_labs) <- c('CRS3123_200mg_po_bid',
                             'CRS3123_400mg_po_bid',
                             'Vancomycin_125mg_po_qid')

visit_legend_labs <- c('HC',
                       'Screen',
                       'D1',
                       'D4',
                       'TOC',
                       'D17', 
                       'D40', 
                       'D70') 

visit_bins <- c('1' = 'before',
                '2' = 'before',
                '3' = 'during',
                '4' = 'toc',
                '5' = 'after',
                '6' = 'after',
                '7' = 'after')

facet_visit_labs <- visit_legend_labs
names(facet_visit_labs) <- c(0, 1, 2, 3, 4, 5, 6, 7)
```

## **metadata: data wrangled**
putting together crestone metadata files so I can have visit data 
```{r}
subject_info <- read_tsv(subject_info_fp)
crestone_meta <- read_tsv(crestone_meta_fp) %>% 
  select(anonymized_name, host_subject_id, visit, visit_number, treatment_group) %>%
  rename(sampleid = anonymized_name,
         subject_id = host_subject_id,
         new_visit_label = visit,
         new_visit_num = visit_number) %>%
  na.omit() %>% 
  mutate(project = paste('crestone'),
         visit_bins = visit_bins[new_visit_num])

## refactoring visit bins so that adonis will compare other time points to toc 
crestone_meta$visit_bins <- factor(crestone_meta$visit_bins, levels = c("toc",
                                                                        "before",
                                                                        "during",
                                                                        "after"))
```

wrangling healthy control meta to match up with crestone metadata for plots 
```{r}
hc_meta <- read_tsv(comb_seq_meta_fp) %>% 
  rename(sampleid = SampleID,
         subject_id = Subject_ID)

proc_hc_meta <- hc_meta %>% 
  filter(project == "healthy_control_AMC") %>% 
  mutate(treat_group1 = paste('CRS3123_200mg_po_bid'),
         treat_group2 = paste('CRS3123_400mg_po_bid'),
         treat_group3 = paste('Vancomycin_125mg_po_qid')) %>% 
  gather(!c('sampleid', 'subject_id', 'project'), key = 'madi_label', value = 'treatment_group') %>% 
  select(-madi_label) %>% 
  mutate(new_visit_num = paste(0),
         new_visit_label = paste('HC'))
```


combining crestone and healthy control metadata 
```{r}
hc_crestone_meta <- rbind(crestone_meta, 
                          proc_hc_meta)
```

## **pcoa files: data wrangled**
distance matrices
```{r}
wu_dist_matrix <- read_tsv(wu_dist_fp) %>% 
  rename('sampleid' = '...1') %>% 
  filter(sampleid %in% crestone_meta$sampleid) %>% 
  column_to_rownames(var = 'sampleid') %>% 
  select(all_of(crestone_meta$sampleid))

uu_dist_matrix <- read_tsv(uu_dist_fp) %>% 
  rename('sampleid' = '...1') %>% 
  filter(sampleid %in% crestone_meta$sampleid) %>% 
  column_to_rownames(var = 'sampleid') %>% 
  select(all_of(crestone_meta$sampleid))
```


actual pcoa values 
```{r}
## for plots 
pcoa_files <- biom_table_prep(unweighted_fp = uu_fp,
                              weighted_fp = wu_fp,
                              faith_fp = faith_pd_fp,
                              metadata_file = hc_crestone_meta)

unweighted_df <- pcoa_files$UnweightedBiom %>% 
  na.omit()
uw_var <- pcoa_files$UnweightedVar

weighted_df <- pcoa_files$WeightedBiom %>% 
  na.omit()
wu_var <- pcoa_files$WeightedVar
```


## **Unweighted UniFrac PCoA plot**
```{r, fig.width=14, fig.height=5}
uw_uni_xlab <- pcoa_ax_lab(uw_var, 'PC1')
uw_uni_ylab <- pcoa_ax_lab(uw_var, 'PC2')

## plot all together
(unweighted_pcoa_plot <- apppleplots::pcoa_plots(input_table = unweighted_df,
                                                x_value = 'PC1',
                                                y_value = 'PC2',
                                                facet_once = TRUE,
                                                facet_twice = FALSE,
                                                point_fill = 'new_visit_num',
                                                point_alpha = 0.7,
                                                point_size = 3,
                                                fill_type = 'discrete',
                                                palette = 'Spectral',
                                                facet_rows = NULL,
                                                facet_cols = 'treatment_group',
                                                row_labs = NULL,
                                                col_labs = treat_group_labs,
                                                legend_name = "Visit",
                                                x_name = uw_uni_xlab,
                                                y_name = uw_uni_ylab,
                                                title_content = 'Unweighted UniFrac PCoA') +
    scale_fill_brewer(palette = 'Spectral',
                      labels = visit_legend_labs) +
    theme_bw(base_size = 20) +
    theme(strip.text.y.right = element_text(angle = 0)))

## plot faceted by visit/HC
(facetVisit_unweighted_pcoa_plot <- apppleplots::pcoa_plots(input_table = unweighted_df,
                                                x_value = 'PC1',
                                                y_value = 'PC2',
                                                facet_once = FALSE,
                                                facet_twice = TRUE,
                                                point_fill = 'new_visit_num',
                                                point_alpha = 0.7,
                                                point_size = 3,
                                                fill_type = 'discrete',
                                                palette = 'Spectral',
                                                facet_rows = 'new_visit_num',
                                                facet_cols = 'treatment_group',
                                                row_labs = facet_visit_labs,
                                                col_labs = treat_group_labs,
                                                legend_name = "Visit",
                                                x_name = uw_uni_xlab,
                                                y_name = uw_uni_ylab,
                                                title_content = 'Unweighted UniFrac PCoA') +
    scale_fill_brewer(palette = 'Spectral',
                      labels = visit_legend_labs) +
    theme_bw(base_size = 20) +
    theme(strip.text.y.right = element_text(angle = 0)))
```

**Unweighted UniFrac Adonis Test**
bin to before, during, toc, and after and compare to toc 
```{r}
## testing distances between visits within a treatment group 
(uu_adonis_byTreatment <- adonis_for_loop(metadata_file = crestone_meta,
                                         dist_matrix_fp = uu_dist_fp,
                                         strat_column = 'treatment_group',
                                         formula_left = 'dist_matrix',
                                         formula_right = 'visit_bins') %>% 
   mutate(dist = "unweighted_unifrac"))
```


## **Weighted UniFrac PCoA Plot**
```{r, fig.width=14, fig.height=5}
wu_uni_xlab <- pcoa_ax_lab(wu_var, 'PC1')
wu_uni_ylab <- pcoa_ax_lab(wu_var, 'PC2')

## plot all together
(weighted_pcoa_plot <- apppleplots::pcoa_plots(input_table = weighted_df,
                                                x_value = 'PC1',
                                                y_value = 'PC2',
                                                facet_once = TRUE,
                                                facet_twice = FALSE,
                                                point_fill = 'new_visit_num',
                                                point_alpha = 0.7,
                                                point_size = 3,
                                                fill_type = 'discrete',
                                                palette = 'Spectral',
                                                facet_rows = NULL,
                                                facet_cols = 'treatment_group',
                                                row_labs = NULL,
                                                col_labs = treat_group_labs,
                                                legend_name = "Visit",
                                                x_name = wu_uni_xlab,
                                                y_name = wu_uni_ylab,
                                                title_content = 'Weighted UniFrac PCoA') +
    scale_fill_brewer(palette = 'Spectral',
                      labels = visit_legend_labs) +
    theme_bw(base_size = 20) +
    theme(strip.text.y.right = element_text(angle = 0)))

## plot faceted by visit/HC
(facetVisit_weighted_pcoa_plot <- apppleplots::pcoa_plots(input_table = weighted_df,
                                                x_value = 'PC1',
                                                y_value = 'PC2',
                                                facet_once = FALSE,
                                                facet_twice = TRUE,
                                                point_fill = 'new_visit_num',
                                                point_alpha = 0.7,
                                                point_size = 3,
                                                fill_type = 'discrete',
                                                palette = 'Spectral',
                                                facet_rows = 'new_visit_num',
                                                facet_cols = 'treatment_group',
                                                row_labs = facet_visit_labs,
                                                col_labs = treat_group_labs,
                                                legend_name = "Visit",
                                                x_name = wu_uni_xlab,
                                                y_name = wu_uni_ylab,
                                                title_content = 'Weighted UniFrac PCoA') +
    scale_fill_brewer(palette = 'Spectral',
                      labels = visit_legend_labs) +
    theme_bw(base_size = 20) +
    theme(strip.text.y.right = element_text(angle = 0)))
```

**Weighted UniFrac Adonis Test**
```{r}
## testing distances between visits within a treatment group
(wu_adonis_byTreatment <- adonis_for_loop(metadata_file = crestone_meta,
                                         dist_matrix_fp = wu_dist_fp,
                                         strat_column = 'treatment_group',
                                         formula_left = 'dist_matrix',
                                         formula_right = 'visit_bins') %>% 
   mutate(dist = "weighted_unifrac"))
```

- treatment group first in formula = tests effects of time (visits - do they differ?) while controlling for different treatment groups
- visits first = tests effects of treatment group (do they differ?) while controlling for time 
  - this explains why you get significance when treatment group is first even though the sum of squares doesn't change!
- filter to only include D4 and TOC visits? 
- higher F statistic = greater variation between groups than within samples of one group (greater differences between treat groups than samples within a treatment group)
- when looking at results sequentially (by = "terms"), significance in a certain term means that it explained significant amounts of the variation in the dataset - if its any term after the first in the formula, it explains a significant portion of the remaining variation that wasn't explained by the terms before it 
```{r}
## significant p-values mean that centroids of groups differ significantly 
## treatment group by itself is v not significant (p-value = 1)

crestone_meta$treatment_group <- factor(crestone_meta$treatment_group, levels = c(
  "Vancomycin_125mg_po_qid",
  "CRS3123_200mg_po_bid",
  "CRS3123_400mg_po_bid"))

test <- adonis2(as.dist(wu_dist_matrix) ~ as.numeric(new_visit_num) * treatment_group,
        data = crestone_meta, 
        permutations = 999,
        strata = crestone_meta$subject_id,
        by = "terms")

tidy(test)
```

**putting together adonis results for weighted/unweighted unifrac**
```{r}
allAdonis_res_by_treatGroup <- rbind(uu_adonis_byTreatment,
                                     wu_adonis_byTreatment)
```


## **Saving my Outputs**
```{r}
save(list = c('weighted_pcoa_plot',
              'unweighted_pcoa_plot'),
       file = '../../../../manuscript_figures/ggplots/pcoa_plots.rdat')
```

```{r}
write_tsv(allAdonis_res_by_treatGroup,
          file = "../stats/16s_pcoa_adonis_res_by_treatGroup.tsv")
```

```{r}
## as a pdf
ggsave('../plots/unweighted_unifrac_pcoa.pdf',
       plot = unweighted_pcoa_plot,
       width = 14,
       height = 5)
ggsave('../plots/weighted_unifrac_pcoa.pdf',
       plot = weighted_pcoa_plot,
       width = 14,
       height = 5)
ggsave('../plots/facetVisit_unweighted_unifrac_pcoa.pdf',
       plot = facetVisit_unweighted_pcoa_plot,
       width = 17,
       height = 15)
ggsave('../plots/facetVisit_weighted_unifrac_pcoa.pdf',
       plot = facetVisit_weighted_pcoa_plot,
       width = 17,
       height = 15)

## png version
ggsave('../plots/pngs/unweighted_unifrac_pcoa.png',
       plot = unweighted_pcoa_plot,
       width = 14,
       height = 5)
ggsave('../plots/pngs/weighted_unifrac_pcoa.png',
       plot = weighted_pcoa_plot,
       width = 14,
       height = 5)
ggsave('../plots/pngs/facetVisit_unweighted_unifrac_pcoa.png',
       plot = facetVisit_unweighted_pcoa_plot,
       width = 17,
       height = 15)
ggsave('../plots/pngs/facetVisit_weighted_unifrac_pcoa.png',
       plot = facetVisit_weighted_pcoa_plot,
       width = 17,
       height = 15)
```

