---
title: "dysbiosis_score"
output: html_document
date: "2025-03-31"
author: "Madi Apgar"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(magrittr)
library(broom)
library(qiime2R)
library(apppleplots)
library(vegan)
library(ggpubr)
library(rio)
```

## **Functions**
`source()` pulls in functions placed in other R scripts and saves them to your global environment!
- here i'm pulling in the statistical analysis functions:
  - `test_normality()` - does basic R tests for normality on the data
  - `art_lmer_notStrat()` - runs aligned rank transform lmer on the data in an overall model (not stratified)
  - `art_lmer_strat()` - runs aligned rank transform lmer on the data in a variable stratified model
  - `art_lmer_doubleStrat()` - runs aligned rank transform lmer on the data in a model stratified by two variables
```{r}
source("../../../art_lmer_functions.R")
source("../../../test_lmer_vars_function.R")
source("../../../diff_between_means_calc.R")

## for distance matrix processing
## for beta diversity statistical analysis 
dist_matrix_prep <- function(metadata_file,
                             dist_matrix_fp){
  ## distance matrix
  dist <- rio::import(dist_matrix_fp)
  names(dist)[names(dist) == 'V1'] <- 'sampleid'
  
  dist %>% 
    gather(-sampleid, key = comp_sample_col, value = dist) %>% 
    filter(sampleid %in% metadata_file$sampleid) %>% 
    filter(comp_sample_col %in% metadata_file$sampleid) %>% 
    filter(str_detect(comp_sample_col, 'HC')) %>% 
           # !str_detect(sampleid, 'HC')) %>% 
    group_by(sampleid) %>% 
    summarise(hc_av_dist = mean(dist)) %>% 
    left_join(metadata_file, by = 'sampleid') -> dist_proc

  return(dist_proc)
}

## small function to assign a value to a key in a dataframe column (i.e. numbers to actual categorical variables)
values_to_keys <- function(input_table,
                           key_col_name,
                           value_col_name,
                           key_value_list,
                           col_order_list){
  
  input_table[value_col_name] <- unlist(lapply(unlist(input_table[key_col_name]), 
                                        FUN = function(x){key_value_list[[x]]}))
  
  output_table <- input_table %>%
    select(all_of(col_order_list), everything())
  
  return(output_table)
}
```

## **File paths**
```{r}
uu_dist_fp <- '../data/uw_dist_matrix.tsv'
wu_dist_fp <- '../data/w_dist_matrix.tsv'
comb_seq_meta_fp <- '../data/healthyControl_comb_metadata.tsv'
crestone_meta_fp <- '../../../../data_submission/1_16S_rRNA_qiita/CRS3123_fecal_samples/crestone_qiita_sample_information.tsv'

## global lists for renaming plot labels
visit_x_labs <- c('0' = 'HC',
                  '1' = 'Screen',
                  '2' = 'D1',
                  '3' = 'D4',
                  '4' = 'TOC',
                  '5' = 'D17', 
                  '6' = 'D40', 
                  '7' = 'D70') 

treat_group_labs <- c('CRS3123 200mg',
                      'CRS3123 400mg',
                      'Vancomycin 125mg')
names(treat_group_labs) <- c('CRS3123_200mg_po_bid',
                             'CRS3123_400mg_po_bid',
                             'Vancomycin_125mg_po_qid')

visit_key_value <- c('1' = 'Screen',
                     '2' = 'D1',
                     '3' = 'D4',
                     '4' = 'D10-15/TOC',
                     '5' = 'D17',
                     '6' = 'D40',
                     '7' = 'D70',
                     '0' = 'HC')
visit_col_order <- c('new_visit_num', 
                     'new_visit_label')

visit_bins <- c('1' = 'before',
                '2' = 'before',
                '3' = 'during',
                '4' = 'toc',
                '5' = 'after',
                '6' = 'after',
                '7' = 'after')
```

## **Metadata: data wrangled**
putting together crestone metadata files so I can have visit data 
```{r}
crestone_meta <- read_tsv(crestone_meta_fp) %>% 
  select(anonymized_name, host_subject_id, visit, visit_number, treatment_group) %>%
  rename(sampleid = anonymized_name,
         subject_id = host_subject_id,
         new_visit_label = visit,
         new_visit_num = visit_number) %>%
  mutate(project = paste('crestone'))
```

wrangling healthy control meta to match up with crestone metadata for plots 
```{r}
hc_meta <- read_tsv(comb_seq_meta_fp) %>% 
  rename(sampleid = SampleID,
         subject_id = Subject_ID)

proc_hc_meta <- hc_meta %>% 
  filter(project == "healthy_control_AMC") %>% 
  mutate(treat_group1 = paste('CRS3123_200mg_po_bid'),
         treat_group2 = paste('CRS3123_400mg_po_bid'),
         treat_group3 = paste('Vancomycin_125mg_po_qid')) %>% 
  gather(!c('sampleid', 'subject_id', 'project'), key = 'madi_label', value = 'treatment_group') %>% 
  select(-madi_label) %>% 
  mutate(new_visit_num = paste(0),
         new_visit_label = paste('HC'))
```

combining crestone and healthy control metadata 
```{r}
hc_crestone_meta <- rbind(crestone_meta, 
                          proc_hc_meta)
```

## **Distance matrix: data wrangled**
```{r}
uu_dist_table <- dist_matrix_prep(hc_crestone_meta,
                                  uu_dist_fp)

wu_dist_table <- dist_matrix_prep(hc_crestone_meta,
                                  wu_dist_fp)

uu_dist_table
wu_dist_table
```

## **Stats**
**Between visits per treatment group**
unweighted unifrac distances 
```{r}
factored_uu_dist_table <- uu_dist_table %>% 
  select(subject_id, new_visit_num, treatment_group, hc_av_dist) %>% 
  na.omit() %>% 
  convert_as_factor(subject_id, new_visit_num, treatment_group)

uu_art_lmer_strat <- art_lmer_strat(input_table = factored_uu_dist_table,
                                    strat_variable = 'treatment_group',
                                    model_formula_left = 'hc_av_dist',
                                    model_formula_right = 'new_visit_num + (1|subject_id)',
                                    model_contrast_run_by = 'new_visit_num')

uu_art_lmerStrat_res <- uu_art_lmer_strat$ARTResults
uu_art_lmerStrat_contrast <- uu_art_lmer_strat$ARTContrast %>% 
  rename(p.adj.signif = signif) %>% 
  mutate(y.position = case_when(
          treatment_group == 'CRS3123_200mg_po_bid' ~ 0.9,
          treatment_group == 'CRS3123_400mg_po_bid' ~ 0.9,
          treatment_group == 'Vancomycin_125mg_po_qid' ~ 0.9
         ),
         ## pulling the column name off the group 1/2 comparisons so I can put the p value signif on the plot
         group1 = gsub("new_visit_num*", "", group1),
         group2 = gsub("new_visit_num*", "", group2)) 
```

weighted unifrac distances 
```{r}
factored_wu_dist_table <- wu_dist_table %>% 
  select(subject_id, new_visit_num, treatment_group, hc_av_dist) %>% 
  na.omit() %>% 
  convert_as_factor(subject_id, new_visit_num, treatment_group)

wu_art_lmer_strat <- art_lmer_strat(input_table = factored_wu_dist_table,
                                    strat_variable = 'treatment_group',
                                    model_formula_left = 'hc_av_dist',
                                    model_formula_right = 'new_visit_num + (1|subject_id)',
                                    model_contrast_run_by = 'new_visit_num')

wu_art_lmerStrat_res <- wu_art_lmer_strat$ARTResults
wu_art_lmerStrat_contrast <- wu_art_lmer_strat$ARTContrast %>% 
  rename(p.adj.signif = signif) %>% 
  mutate(y.position = case_when(
          treatment_group == 'CRS3123_200mg_po_bid' ~ 2,
          treatment_group == 'CRS3123_400mg_po_bid' ~ 2,
          treatment_group == 'Vancomycin_125mg_po_qid' ~ 2
         ),
         ## pulling the column name off the group 1/2 comparisons so I can put the p value signif on the plot
         group1 = gsub("new_visit_num*", "", group1),
         group2 = gsub("new_visit_num*", "", group2)) 
```

**Between treatment group for each visit**
unweighted unifrac distances 
```{r}
uu_lmer_treatGroup_strat <- art_lmer_strat(input_table = factored_uu_dist_table,
                                    strat_variable = 'new_visit_num',
                                    model_formula_left = 'hc_av_dist',
                                    model_formula_right = 'treatment_group',
                                    model_contrast_run_by = 'treatment_group')

(uu_lmer_treatGroup_strat_res <- uu_lmer_treatGroup_strat$ARTResults)
uu_lmer_treatGroup_strat_contrast <- uu_lmer_treatGroup_strat$ARTContrast 

(proc_uu_lmer_treatGroup_strat_contrast <- values_to_keys(input_table = uu_lmer_treatGroup_strat_contrast,
                                                         key_col_name = 'new_visit_num',
                                                         value_col_name = 'new_visit_label',
                                                         key_value_list = visit_key_value,
                                                         col_order_list = visit_col_order))
```

weighted unifrac distances 
```{r}
wu_lmer_treatGroup_strat <- art_lmer_strat(input_table = factored_wu_dist_table,
                                    strat_variable = 'new_visit_num',
                                    model_formula_left = 'hc_av_dist',
                                    model_formula_right = 'treatment_group',
                                    model_contrast_run_by = 'treatment_group')

(wu_lmer_treatGroup_strat_res <- wu_lmer_treatGroup_strat$ARTResults)
wu_lmer_treatGroup_strat_contrast <- wu_lmer_treatGroup_strat$ARTContrast 

(proc_wu_lmer_treatGroup_strat_contrast <- values_to_keys(input_table = wu_lmer_treatGroup_strat_contrast,
                                                         key_col_name = 'new_visit_num',
                                                         value_col_name = 'new_visit_label',
                                                         key_value_list = visit_key_value,
                                                         col_order_list = visit_col_order) %>% 
  select(new_visit_label, new_visit_num, group1, group2, p.value, signif))
```

**Overall linear mixed effects model**
unweighted unifrac 
```{r}
filt_factored_uu_dist_table <- factored_uu_dist_table %>% 
  filter(new_visit_num != 0) %>% 
  mutate(treat_group_num = case_when(
    treatment_group == "CRS3123_200mg_po_bid" ~ 1,
    treatment_group == "CRS3123_400mg_po_bid" ~ 2,
    treatment_group == "Vancomycin_125mg_po_qid" ~ 3,
  ),
  new_visit_num = as.character(new_visit_num),
  visit_bins = visit_bins[new_visit_num],
  visit_bins = factor(visit_bins, levels = c("toc",
                                             "before",
                                             "during",
                                             "after")),
  new_visit_num = as.numeric(new_visit_num))

mixed_modeling(predictors = c('treat_group_num', 'new_visit_num', 'visit_bins'),
               response_vars = c('hc_av_dist'),
               random_effect = '(1|subject_id)',
               data = filt_factored_uu_dist_table,
               reml = FALSE)
```
 weighted unifrac 
```{r}
filt_factored_wu_dist_table <- factored_wu_dist_table %>% 
  filter(new_visit_num != 0) %>% 
  mutate(treat_group_num = case_when(
    treatment_group == "CRS3123_200mg_po_bid" ~ 1,
    treatment_group == "CRS3123_400mg_po_bid" ~ 2,
    treatment_group == "Vancomycin_125mg_po_qid" ~ 3,
  ),
  new_visit_num = as.character(new_visit_num),
  visit_bins = visit_bins[new_visit_num],
  visit_bins = factor(visit_bins, levels = c("toc",
                                             "before",
                                             "during",
                                             "after")),
  new_visit_num = as.numeric(new_visit_num))

mixed_modeling(predictors = c('treat_group_num', 'new_visit_num', 'visit_bins'),
               response_vars = c('hc_av_dist'),
               random_effect = '(1|subject_id)',
               data = filt_factored_wu_dist_table,
               reml = FALSE)
```

## **Attempting some kind of visualization**
unweighted unifrac distances
```{r, fig.width=17, fig.height=7}
(uu_dist_plot <- uu_dist_table %>% 
   na.omit() %>% 
  ggplot(aes(x = new_visit_num, y = hc_av_dist)) +
  geom_boxplot(aes(group = new_visit_num, fill = new_visit_num), outlier.shape = NA) +
  geom_jitter(aes(fill = new_visit_num), width = 0.1, height = 0, alpha = 0.7, size = 2.5, pch = 21) +
  scale_fill_brewer(palette = 'Spectral') +
  theme_bw(base_size = 20) +
  theme(legend.position = 'none') +
  facet_wrap(~treatment_group,
             labeller = labeller(treatment_group = treat_group_labs)) +
  stat_pvalue_manual(uu_art_lmerStrat_contrast,
                     step.group.by = 'treatment_group',
                     step.increase = 0.1,
                     hide.ns = TRUE,
                     label = 'p.adj.signif') +
  scale_x_discrete(labels = visit_x_labs) +
  labs(x = 'Visit',
       y = 'Average Distance',
       title = 'Unweighted UniFrac Distance from Healthy Controls'))
```

weighted unifrac distances 
```{r, fig.width=17, fig.height=7}
(wu_dist_plot <- wu_dist_table %>% 
   na.omit() %>% 
  ggplot(aes(x = new_visit_num, y = hc_av_dist)) +
  geom_boxplot(aes(group = new_visit_num, fill = new_visit_num), outlier.shape = NA) +
  geom_jitter(aes(fill = new_visit_num), width = 0.1, height = 0, alpha = 0.7, size = 2.5, pch = 21) +
  scale_fill_brewer(palette = 'Spectral') +
  theme_bw(base_size = 20) +
  theme(legend.position = 'none') +
  facet_wrap(~treatment_group,
             labeller = labeller(treatment_group = treat_group_labs)) +
  stat_pvalue_manual(wu_art_lmerStrat_contrast,
                     step.group.by = 'treatment_group',
                     step.increase = 0.1,
                     hide.ns = TRUE,
                     label = 'p.adj.signif') +
  scale_x_discrete(labels = visit_x_labs) +
  labs(x = 'Visit',
       y = 'Average Distance',
       title = 'Weighted UniFrac Distance from Healthy Controls'))
```

## **Stat heat plots**
unweighted unifrac has no significant differences so not plotting!

weighted unifrac only 
```{r, fig.width=15, fig.height=5}
## calculating difference bt means to give directionality
wu_treatGroup_strat_diffs <- calc_diff_means(input_table = factored_wu_dist_table,
                                             first_group = 'treatment_group',
                                             second_group = 'new_visit_num',
                                             mean_value = 'hc_av_dist',
                                             post_hoc_test = proc_wu_lmer_treatGroup_strat_contrast) %>% 
  filter(new_visit_num != 0)

(wu_treatGroup_strat_heatPlot <- apppleplots::stat_heat_plot(input_table = wu_treatGroup_strat_diffs,
                            x_value = 'new_visit_num',
                            y_value = 'group2',
                            x_value_type = 'discrete',
                            x_labels = visit_x_labs,
                            y_value_type = 'discrete',
                            y_labels = treat_group_labs,
                            tile_fill = 'stat_diff_means',
                            text_value = 'signif',
                            legend_name = "Group 1 -\nGroup 2",
                            facet_by = 'group1',
                            facet_labels = treat_group_labs,
                            x_name = 'Visit',
                            y_name = 'Group 2',
                            title_content = 'Weighted UniFrac Average Distance from HCs Differences',
                            pairwise = TRUE,
                            low_color = "#0D0887FF",
                            high_color = "#CC4678FF") +
  theme_bw(base_size = 20) +
  theme(strip.text.y.right = element_text(angle = 0),
        plot.subtitle = element_text(hjust = 0.5)))
```

## **Saving my outputs**
plots
```{r}
save(list = c('uu_dist_plot',
              'wu_dist_plot',
              'wu_treatGroup_strat_heatPlot'),
       file = '../../../../manuscript_figures/ggplots/dysbiosis_score_plots.rdat')
```

```{r}
## as a pdf
ggsave('../plots/uu_hc_avDist_plot.pdf',
       plot = uu_dist_plot,
       width = 17,
       height = 7)
ggsave('../plots/wu_hc_avDist_plot.pdf',
       plot = wu_dist_plot,
       width = 17,
       height = 7)


## as a png
ggsave('../plots/pngs/uu_hc_avDist_plot.png',
       plot = uu_dist_plot,
       width = 17,
       height = 7)
ggsave('../plots/pngs/wu_hc_avDist_plot.png',
       plot = wu_dist_plot,
       width = 17,
       height = 7)
```

stats
```{r}
write_tsv(proc_wu_lmer_treatGroup_strat_contrast,
          '../stats/wu_avDist_treatGroup_strat_postHoc.tsv')
write_tsv(wu_treatGroup_strat_diffs,
          '../stats/diffs_wu_avDist_treatGroup_strat_postHoc.tsv')
```


