---
title: "figures"
output: html_document
date: "2025-10-07"
author: "Madi Apgar"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(cowplot)
library(tidyverse)
library(viridis)
library(ggplot2)
library(ggrepel)
library(ggh4x)
library(ggpubr)
library(tidytext)
```

**Reading in plots from .rdat**
```{r}
## loads in anything 
figure_list <- Sys.glob('ggplots/*.rdat')

for(file in unique(unlist(figure_list))){
  load(file)
}
```

# **Main Figures**

## **Figure 1**
probs won't be in here, will be put together via inkscape 
plots:
- participant demographics 
- experimental timeline 

## **Figure 2**
16S microbiome data (microbiome data in general)
plots:
- faiths pd w stats
- weighted unifrac pcoa
- weighted unifrac dysbiosis index

**editing plots for figure**
```{r}
# viridis_pal(option = "C", begin = 0, end = 1)(3)

my_viridis_c_pal <- c("#0D0887FF",
                      "#5D01A6FF",
                      "#9C179EFF",
                      "#CC4678FF",
                      "#ED7953FF",
                      "#FDB32FFF",
                      "#F0F921FF",
                      "#999999")

my_viridis_c_pal_rev <- c("#999999",
                          "#0D0887FF",
                          "#5D01A6FF",
                          "#9C179EFF",
                          "#CC4678FF",
                          "#ED7953FF",
                          "#FDB32FFF",
                          "#F0F921FF")
```

```{r}
## changing transparency of boxplots
wu_dist_plot@layers$geom_boxplot$aes_params[["alpha"]] <- 0.7

## weighted dysbiosis index plot color palette and label changes
wu_dist_plot_v2 <- wu_dist_plot +
                     scale_fill_manual(values = my_viridis_c_pal_rev) +
                     labs(title = "Beta Diversity: Sample Average Distance from Healthy Controls",
                          subtitle = "Weighted UniFrac")
```

**putting figure together**
possibly put the faith's pd deltas 
```{r, fig.width=17, fig.height=14}
(fig2 <- plot_grid(faith_plot_wStats,
                   wu_dist_plot_v2,
                   nrow = 2,
                   labels = c('a)', 'b)'),
                   align = c('hv'),
                   axis = c('tblr'),
                   label_size = 22))
```

## **Figure 3**
16S taxonomic information and biplots
plots: only show PC1/2?
- 16S taxa biplot (weighted?)

**editing plots for figure**
```{r}
## changing plot titles and legend positions 
fig3a_plot_v2 <- fig3a_plot +
                   labs(title = "16S Relative Abundance: Pathogenic Taxa of Interest")

fig3b_plot_v2 <- fig3b_plot +
                   labs(title = "16S Relative Abundance: Enterococcus Compared to VRE Culture Status") +
                   theme(legend.position = "top",
                         plot.subtitle = element_blank())

fig3c_plot_v2 <- fig3c_plot +
                    labs(title = '16S Relative Abundance: Mutualistic Taxa of Interest')
```

**putting figure together**
```{r, fig.width=16, fig.height=21}
(fig3 <- plot_grid(fig3a_plot_v2,
                   fig3b_plot_v2,
                   fig3c_plot_v2,
                   labels = c('a)', 'b)', 'c)'),
                   label_size = 22,
                   ncol = 1,
                   rel_heights = c(0.7, 0.55, 0.7)))
```

## **Figure 4**
functional potential/shotgun meta figure
plots:
- gsea between treatment groups - send 200/400 comparison to supplemental bc idk what its adding here
- secondary bile acid pathway ko breakdown 
- secondary bile acid pathway taxa contributions

**editing plots for figure**
```{r}
## changing color palettes and plot labels 
bile_acid_path_wTax_v2 <- bile_acid_path_wTax +
                               labs(title = 'TOC: Taxonomic Contribution to Secondary Bile Acid Biosynthesis Pathway Genes') +
                               theme_bw(base_size = 20) +
                               theme(legend.position = "none",
                                     plot.subtitle = element_blank(),
                                     strip.text.y.right = element_text(angle = 0),
                                     axis.text.x = element_blank(),
                                     axis.ticks.x = element_blank(),
                                     axis.title.x = element_blank())

toc_gsea_plot_v2 <- toc_gsea_plot +
  labs(title = 'TOC: Treatment Group Pathway Representation Differences')
```

**putting figure together**
```{r, fig.width=27, fig.height=20}
fig4_top_row <- plot_grid(toc_gsea_plot_v2,
                          labels = c('a)'),
                          label_size = 22,
                          nrow = 1)

fig4_bottom_row <- plot_grid(bile_acid_path_plot,
                             bile_acid_path_wTax_v2,
                             labels = c('b)', 'c)'),
                             label_size = 22,
                             nrow = 1,
                             rel_widths = c(1, 1))

(fig4 <- plot_grid(fig4_top_row,
                   fig4_bottom_row,
                   nrow = 2,
                   rel_heights = c(0.5, 1)))
```

## **Figure 5**
targeted metabolomics data (SCFAs and bile acids)
plots:
- scfa concentration plots (acetate, butyrate, propionate) - move this to supplemental!!
- bile acid plots - degree of transformation, primary v secondary, LCA plot
- i think i should definitely put the stats on these plots

**editing plots for figure**
```{r}
## changing x-axis labels 
lithocholicacid_only_plot_v2 <- `lithocholicacid_(lca)_only_plot` +
                                    labs(x = 'Visit')

bile_acid_state_plot_v2 <- bile_acid_state_plot +
                              labs(x = 'Visit')
```

**putting figure together**
```{r, fig.width=15, fig.height=14}
(fig5 <- plot_grid(lithocholicacid_only_plot_v2,
                   bile_acid_state_plot_v2,
                   labels = c('a)', 'b)'),
                   label_size = 22,
                   rel_heights = c(0.6, 1),
                   ncol = 1) +
          theme(plot.background = element_rect(fill = "white", color = "white")))

```

## **Figure 6**
untargeted metabolomics data 
plots:
- lipid fraction pca
- aqueous fraction pca 

**editing plots for figure**
```{r}
lipid_euclidean_plot_v2 <- lipid_euclidean_plot +
                              labs(title = "Lipid Metabolome Changes",
                                   subtitle = "Euclidean Distances")


aqueous_euclidean_plot_v2 <- aqueous_euclidean_plot +
                                labs(title = "Aqueous Metabolome Changes",
                                     subtitle = "Euclidean Distances")
```

**putting figure together**
```{r, fig.width=15, fig.height=14}
(fig6 <- plot_grid(aqueous_euclidean_plot_v2,
                   lipid_euclidean_plot_v2,
                   nrow = 2, 
                   labels = c('a)', 'b)'),
                   label_size = 22))
```

## **Figure 7** 
host immune profiling figure (fecal water assay)
plots:
- concentration of significant immune markers over time 
- stats comparing treatment group concentrations at D4

**editing plots for figure**
```{r}
two_marker_plot_v2 <- two_marker_plot +
                        theme_bw(base_size = 20) +
                        theme(strip.text.y.right = element_text(angle = 0)) +
                        labs(y = 'Concentration (pg/mL) (log10)')

visit_strat_cp_il1b_heat_plot_v2 <- visit_strat_cp_il1b_heat_plot +
                               theme_bw(base_size = 20) +
                               theme(strip.text.y.right = element_text(angle = 0),
                                     plot.subtitle = element_text(hjust = 0.5)) +
                               scale_fill_gradient2(low = "#0D0887FF",
                                                    high = "#CC4678FF",
                                                    name = 'Log2FC\nof Group 1\nvs. Group 2')
```

**putting figure together**
```{r, fig.width=17, fig.height=17}
fig7_bottom_row <- plot_grid(visit_strat_cp_il1b_heat_plot_v2,
                             NULL,
                             nrow = 1,
                             rel_widths = c(1, 0.6))


(fig7 <- plot_grid(two_marker_plot_v2,
                  fig7_bottom_row,
                  nrow = 2,
                  rel_heights = c(1, 0.5),
                  labels = c('a)', ''),
                  label_size = 22) +
    theme(plot.background = element_rect(fill = "white", color = "white")))
```

# **Supplemental figure(s)**

## **Figure S5**
- used to be figure S2

**editing plots for figure**
```{r}
## faiths pd deltas plots
d1_toc_deltas_plot_v2 <- d1_toc_deltas_plot +
                              labs(title = "Faith's PD Differences within Subject",
                                   subtitle = "TOC-D1\nDiff > 0: higher at TOC | Diff < 0: higher at D1",
                                   y = "Difference") +
                              theme(axis.title.x = element_blank())

d40_toc_deltas_plot_v2 <- d40_toc_deltas_plot +
                              labs(subtitle = "D40-TOC\nDiff > 0: higher at D40 | Diff < 0: higher at TOC") +
                              theme(plot.title = element_blank(),
                                    axis.title.y = element_blank(),
                                    axis.title.x = element_blank())

## weighted pcoa plot color palette and label changes
weighted_pcoa_plot_v2 <- weighted_pcoa_plot +
                           scale_fill_manual(breaks = c(1, 2, 3, 4, 5, 6, 7, 0),
                                             labels = c("Screen", "D1", "D4", "TOC", "D17", "D40", "D70", "HC"),
                                             values = my_viridis_c_pal,
                                             name = "Visit") +
                           labs(title = "Beta Diversity: Sample PCoA Compared to Healthy Controls",
                                subtitle = "Weighted UniFrac")
```

**putting figure together** 
```{r, fig.width=15, fig.height=15}
## creating universal y axis label since the y axis is the same for both plots 
deltas_y_axis_label <- ggdraw() +
  draw_label('Treatment Group',
             size = 20,
             hjust = 0.5,
             y = 1) +
  theme(plot.background = element_rect(fill = "white", color = "white"))

figS5_top_row <- plot_grid(d1_toc_deltas_plot_v2, 
                           d40_toc_deltas_plot_v2,
                           nrow = 1,
                           align = c('hv'),
                           axis = c('tblr'))

figS5_top_row_wLabs <- plot_grid(figS5_top_row,
                                 deltas_y_axis_label,
                                 ncol = 1,
                                 rel_heights = c(1, 0.05),
                                 align = c('hv'),
                                 axis = c('tblr'))

(figS5 <- plot_grid(figS5_top_row_wLabs,
                   weighted_pcoa_plot_v2,
                   wu_treatGroup_strat_heatPlot +
                     labs(title = 'Beta Diversity: Average Distances from Healthy Controls (Weighted UniFrac)') +
                     scale_y_discrete(labels = c('CRS3123\n400mg',
                                                  'Vancomycin\n125mg')),
                   nrow = 3,
                   labels = c('a)', 'b)', 'c)'),
                   label_size = 22,
                   rel_heights = c(1, 0.8, 0.7)))
```

## **Figure S6**
- used to be figure S3

**editing plots for figure**
```{r}
## facet labs 
treat_labs <- c(
  CRS3123_200mg_po_bid = "CRS3123 200mg",
  CRS3123_400mg_po_bid = "CRS3123 400mg",
  Vancomycin_125mg_po_qid = "Vancomycin 125mg"
)

#setting abx_levels and manual color palette for consistency
abx_levels <- c("Screen, D1" = "Pre (Screen, D1)",
                "D4" = "During (D4)", 
                "TOC" = "TOC", 
                "D17, D40, D70" = "Post (D17, D40, D70)")

abx_palette <- viridis(
  n = length(abx_levels),
  option = "C",
  begin = 0,
  end = 0.95
)


wu_biplot_facet_v2 <- wu_biplot_facet +
                        scale_fill_manual(values = abx_palette,
                                          breaks = names(abx_levels),
                                          labels = abx_levels,
                                          name = 'Visits Relative to Treatment') + 
                        facet_wrap(~treatment_group,
                                   nrow = 1, 
                                   labeller = labeller(treatment_group = treat_labs)) +
                        theme_bw(base_size = 20) +
                        theme(legend.position = "top") +
                        guides(fill = guide_legend(ncol = 4))


buty_treatGroup_strat_heatPlot_v2 <- buty_treatGroup_strat_heatPlot +
                      ggtitle('16S Relative Abundance: Butyrate Producing Family Differences') +
                      scale_y_discrete(labels = c('CRS3123\n400mg',
                                                  'Vancomycin\n125mg'))

genusAbun_treatGroup_strat_heatPlot_v2 <- genusAbun_treatGroup_strat_heatPlot +
                      ggtitle('16S Relative Abundance: Pathogenic and Mutualistic Taxa Differences') +
                      scale_y_discrete(labels = c('CRS3123\n400mg',
                                                  'Vancomycin\n125mg'))

buty_abun_plot_v2 <- buty_abun_plot +
                      labs(title = '16S Relative Abundance: Main Butyrate Producing Families',
                           x = 'Visit')
```

**putting figures together**
```{r, fig.height=18, fig.width=30}
figS6_bottom_row <- plot_grid(buty_treatGroup_strat_heatPlot_v2,
                              wu_biplot_facet_v2,
                              nrow = 1,
                              rel_widths = c(0.85, 1),
                              label_size = 22,
                              labels = c('c)', 'd)'))

figS6_top_row <- plot_grid(genusAbun_treatGroup_strat_heatPlot_v2,
                           buty_abun_plot_v2,
                           ncol = 2,
                           label_size = 22,
                           labels = c('a)', 'b)'))

(figS6 <- plot_grid(figS6_top_row,
                    figS6_bottom_row,
                    nrow = 2,
                    rel_heights = c(1, 0.8)))
```

## **Figure S7**
- used to be figure S4
plots:
- weighted unifrac shotgun biplot PC1-PC2
- D1 pathway analysis results 

**editing plots for figure**
```{r}
## shotgun biplot
shotgun_wu_biplot_1to2_v2 <- shotgun_wu_biplot_1to2 +
                               labs(title = 'Metagenomics: Taxa Influencing Distances',
                                    subtitle = 'Weighted UniFrac') +
                               theme(legend.position = "right")


## d1 pathway analysis plot 
d1_gsea_plot_v2 <- d1_gsea_plot +
                      labs(title = 'Baseline (D1): Treatment Group Pathway Representation Differences')
```

**putting figures together**
```{r, fig.width=25, fig.height=18}
(figS7 <- plot_grid(shotgun_wu_biplot_1to2_v2,
                   d1_gsea_plot_v2,
                   nrow = 2, 
                   labels = c('a)', 'b)'),
                   label_size = 22,
                   rel_heights = c(0.8, 1),
                   align = c('hv'),
                   axis = c('tblr')))
```

## **Figure S8**
- used to be figure S5
```{r, fig.width=13, fig.height=10}
(figS8 <- plot_grid(mini_scfaConc_plot +
                      labs(x = 'Visit'),
                    labels = c('a)'),
                    label_size = 22))
```

## **Figure S9**
- used to be figure S7
immune profile biplots by treatment group

plots:
- canberra distance biplots (only significant vectors included)
- stats comparing concentrations of all immune markers at each time point between treatment groups

```{r}
p_pc1_pc2_facet_v2 <- p_pc1_pc2_facet +
                       theme_bw(base_size = 20) +
                       theme(legend.position = "top") +
                       guides(fill = guide_legend(title = 'Visit Groupings'),
                              color = guide_legend(title = 'Visit Groupings'))

immu_visit_strat_heat_plot_v2 <- visit_strat_heat_plot +
                                   theme_bw(base_size = 20) +
                                   theme(strip.text.y.right = element_text(angle = 0),
                                         plot.subtitle = element_text(hjust = 0.5)) +
                                   scale_fill_gradient2(low = "#0D0887FF",
                                             high = "#CC4678FF",
                                             name = 'Log2FC\nof Group 1\nvs. Group 2')
```

```{r, fig.width=18, fig.height=18}
(figS9 <- plot_grid(p_pc1_pc2_facet_v2,
                    immu_visit_strat_heat_plot_v2,
                    nrow = 2,
                    labels = c('a)', 'b)'),
                    label_size = 22) +
   theme(plot.background = element_rect(fill = "white", color = "white")))
```


## **VRE plot edits for Laurie/poster**
```{r, fig.width=15.5, fig.height=6}
fig3b_plot_vLaurie <- fig3b_plot +
                   labs(title = "16S Relative Abundance: Enterococcus Compared to VRE Culture Status") +
                   theme(legend.position = "top",
                         plot.subtitle = element_blank(),
                         legend.text = element_text(size = 20)) +
                   guides(fill = guide_legend(override.aes = list(size = 4)))

fig3b_plot_vLaurie@layers$geom_jitter...4$aes_params[["size"]] <- 4

fig3b_plot_vLaurie
```


## **Saving figure drafts**
```{r}
## fig2
ggsave('fig2.pdf',
       plot = fig2,
       width = 17,
       height = 14)
ggsave('pngs/fig2.png',
       plot = fig2,
       width = 17,
       height = 14)

## fig3
ggsave('fig3.pdf',
       plot = fig3,
       width = 16,
       height = 21)
ggsave('pngs/fig3.png',
       plot = fig3,
       width = 16,
       height = 21)

## fig3 - VRE plot for laurie
ggsave('../madi_crestone/plots/lauries_enteroVRE_plot.pdf',
       plot = fig3b_plot_vLaurie,
       width = 15.5,
       height = 6)
ggsave('../madi_crestone/plots/pngs/lauries_enteroVRE_plot.png',
       plot = fig3b_plot_vLaurie,
       width = 15.5,
       height = 6)

## fig4
ggsave('fig4.pdf',
       plot = fig4,
       width = 27,
       height = 20)
ggsave('pngs/fig4.png',
       plot = fig4,
       width = 27,
       height = 20)

## fig5
ggsave('fig5.pdf',
       plot = fig5,
       width = 15,
       height = 14)
ggsave('pngs/fig5.png',
       plot = fig5,
       width = 15,
       height = 14)

## fig6
ggsave('fig6.pdf',
       plot = fig6,
       width = 15,
       height = 14)
ggsave('pngs/fig6.png',
       plot = fig6,
       width = 15,
       height = 14)

## fig7
ggsave('fig7.pdf',
       plot = fig7,
       width = 17,
       height = 17)
ggsave('pngs/fig7.png',
       plot = fig7,
       width = 17,
       height = 17)

## figS2
ggsave('figS5.pdf',
       plot = figS5,
       width = 15,
       height = 15)
## png version needs to be longer for some reason or the text gets cut off 
ggsave('pngs/figS5.png',
       plot = figS5,
       width = 17,
       height = 15)

## figS3
ggsave('figS6.pdf',
       plot = figS6,
       width = 30,
       height = 18)
ggsave('pngs/figS6.png',
       plot = figS6,
       width = 30,
       height = 18)

## figS4
ggsave('figS7.pdf',
       plot = figS7,
       width = 25,
       height = 18)
ggsave('pngs/figS7.png',
       plot = figS7,
       width = 25,
       height = 18)

## figS5 
ggsave('figS8.pdf',
       plot = figS8,
       width = 13,
       height = 10)
ggsave('pngs/figS8.png',
       plot = figS8,
       width = 13,
       height = 10)

## figS7
ggsave('figS9.pdf',
       plot = figS9,
       width = 18,
       height = 18)
ggsave('pngs/figS9.png',
       plot = figS9,
       width = 20,
       height = 18)
```

## **Figure graveyard**
all figures that I put together that got taken out but keeping just in case we want them again

**16S taxa biplot with fig3**
```{r}
fig3_old <- plot_grid(fig3,
                   wu_biplot_facet_v2,
                   nrow = 1,
                   rel_widths = c(1, 0.8),
                   label_size = 22,
                   labels = c('', 'd)'))
```


**Shotgun taxa biplot**
```{r, fig.width=24, fig.height=17}
shotgun_title <- ggdraw() +
  draw_label('Shotgun Metagenomics: Weighted UniFrac Taxa Biplots',
             size = 28,
             hjust = 1.12) +
  theme(plot.background = element_rect(fill = "white", color = "white"))

full_shotgun_plot <- plot_grid(shotgun_wu_biplot_1to2 +
                     ggtitle('PC1-PC2') +
                     scale_fill_viridis(option = "C", discrete = TRUE,
                                        name = "Visit",
                                        labels = c("D1", "TOC", "Screen")),
                   shotgun_wu_biplot_2to3 +
                     ggtitle('PC2-PC3') +
                     scale_fill_viridis(option = "C", discrete = TRUE,
                                        name = "Visit",
                                        labels = c("D1", "TOC", "Screen")),
                  nrow = 2,
                  align = c('hv'),
                  axis = c('tblr'))

(full_shotgun_plot_wTitle <- plot_grid(shotgun_title,
                   full_shotgun_plot,
                   nrow = 2,
                   align = c('hv'),
                   axis = c('tblr'),
                   rel_heights = c(0.07, 1),
                   labels = c('b)'),
                   label_size = 22))
```

