---
title: "alpha_div"
output: html_document
date: "2025-02-12"
author: "Madi Apgar"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(magrittr)
library(broom)
library(qiime2R)
library(apppleplots)
library(ggpubr)
library(cgwtools)
library(lme4)
```

**Functions**
`source()` pulls in functions placed in other R scripts and saves them to your global environment!
- here i'm pulling in the statistical analysis functions:
  - `test_normality()` - does basic R tests for normality on the data
  - `art_lmer_notStrat()` - runs aligned rank transform lmer on the data in an overall model (not stratified)
  - `art_lmer_strat()` - runs aligned rank transform lmer on the data in a variable stratified model
```{r}
source("../../../art_lmer_functions.R")
source("../../../normality_function.R")
source("../../../test_lmer_vars_function.R")
```

**File paths**
```{r}
faith_pd_fp <- '../data/faith_pd.tsv'
shannon_fp <- '../data/shannon_entropy.tsv'
proc_seq_meta_fp <- '../../../../data_submission/1_16S_rRNA_qiita/CRS3123_fecal_samples/crestone_qiita_sample_information.tsv'

## global plot labels 
treat_group_labs <- c('CRS3123 200mg',
                      'CRS3123 400mg',
                      'Vancomycin 125mg')
names(treat_group_labs) <- c('CRS3123_200mg_po_bid',
                             'CRS3123_400mg_po_bid',
                             'Vancomycin_125mg_po_qid')

visit_x_order <- c('Screen',
                  'D1',
                  'D4',
                  'D10-15/TOC',
                  'D17', 
                  'D40', 
                  'D70')

visit_x_labs <- c('Screen',
                  'D1',
                  'D4',
                  'TOC',
                  'D17', 
                  'D40', 
                  'D70') 
```

**Reading in files**
```{r}
faith_pd <- read_tsv(faith_pd_fp) %>% 
  rename(sampleid = `#SampleID`)

shannon <- read_tsv(shannon_fp) %>% 
  rename(sampleid = `...1`)

proc_seq_meta <- read_tsv(proc_seq_meta_fp) %>%
  select(anonymized_name, host_subject_id, visit, visit_number, treatment_group) %>%
  rename(sampleid = anonymized_name,
         subject_id = host_subject_id,
         new_visit_label = visit,
         new_visit_num = visit_number)
```

**Data wrangled**
```{r}
## combining faith, shannon, and clinical meta w seq metadata
alphaDiv_seq_meta <- proc_seq_meta %>% 
  left_join(faith_pd, by = 'sampleid') %>% 
  left_join(shannon, by = 'sampleid') %>% 
  na.omit() %>% 
  mutate(treat_group_num = case_when(
    treatment_group == "CRS3123_200mg_po_bid" ~ 1,
    treatment_group == "CRS3123_400mg_po_bid" ~ 2,
    treatment_group == "Vancomycin_125mg_po_qid" ~ 3,
  ))
```

## **Stats**

**testing for normality**
- running a two-way repeated measures anova, assuming normality
- want to account for multiple samples taken from same subject at different time points
- this isn't working bc there isn't consistent samples taken at all the different time points per subject so I can't do repeated measures 
```{r}
test_normality(input_table = alphaDiv_seq_meta,
               formula_left = 'faith_pd',
               formula_right = 'new_visit_num * treatment_group',
               group_by_vars = c('treatment_group', 'new_visit_num'),
               facet_by = 'treatment_group')
```

**ART LMER Stats**
another alternative to a two-way non-parametric ANOVA: aligned rank transform (ART)
how to include repeated measures term??:
- can use Error(subject_id) in the formula but this means the model will be stratified by subject_id and implies that the data MUST be balanced, especially if you have more than two strata (which I do)
- can use (1|subject_id) to run a linear mixed effects model while taking into account multiple samples from the same subject 
- partial eta squared tells us how large of an effect the independent variable(s) (RHS) have on the dependent variable (LHS)

contrasts analysis pointers:
- can specify one of the variables to see multiple comparisons (i.e. 'visit_num' or 'treatment_group' individually)
- can also get all comparisons for the combinations of the variables (i.e. 'visit_num:treatment_group')

**Not stratified analysis**
```{r}
factored_alphaDiv <- alphaDiv_seq_meta %>% 
  select(subject_id, new_visit_num, treatment_group, faith_pd) %>%  
  convert_as_factor(subject_id, new_visit_num, treatment_group)

## overall non-stratified model 
faithPD_art_lmer <- art_lmer_notStrat(input_table = factored_alphaDiv,
                                      formula_left = 'faith_pd',
                                      formula_right = 'new_visit_num * treatment_group + (1|subject_id)',
                                      contrast_run_by = 'new_visit_num:treatment_group')

faithPD_art_lmer_notStrat_res <- faithPD_art_lmer$ARTResults
faithPD_art_lmer_notStrat_contrast <- faithPD_art_lmer$ARTContrast
```

**Stratified analysis**
running a separate model for each treatment group - the same as grouping by the treatment group and running stats, overall model results are above 
```{r}
faithPD_art_lmer_strat <- art_lmer_strat(input_table = factored_alphaDiv,
                                         strat_variable = 'treatment_group',
                                         model_formula_left = 'faith_pd',
                                         model_formula_right = 'new_visit_num + (1|subject_id)',
                                         model_contrast_run_by = 'new_visit_num')

faithPD_art_lmerStrat_res <- faithPD_art_lmer_strat$ARTResults
faithPD_art_lmerStrat_contrast <- faithPD_art_lmer_strat$ARTContrast %>% 
  rename(p.adj.signif = signif) %>% 
  mutate(y.position = case_when(
    treatment_group == 'CRS3123_200mg_po_bid' ~ 40,
    treatment_group == 'CRS3123_400mg_po_bid' ~ 50,
    treatment_group == 'Vancomycin_125mg_po_qid' ~ 60
  ))
```

```{r}
## pulling the column name off the group 1/2 comparisons so I can put the p value signif on the plot
faithPD_art_lmerStrat_contrast$group1 <- gsub("new_visit_num*", "", faithPD_art_lmerStrat_contrast$group1)
faithPD_art_lmerStrat_contrast$group2 <- gsub("new_visit_num*", "", faithPD_art_lmerStrat_contrast$group2)
```

between treatment groups at a given day 
```{r}
art_lmer_strat(input_table = factored_alphaDiv,
                                         strat_variable = 'new_visit_num',
                                         model_formula_left = 'faith_pd',
                                         model_formula_right = 'treatment_group',
                                         model_contrast_run_by = 'treatment_group')
```

```{r}
mixed_modeling(predictors = c('treat_group_num', 'new_visit_num'),
               response_vars = c('faith_pd', 'shannon_entropy'),
               random_effect = '(1|subject_id)',
               data = alphaDiv_seq_meta,
               reml = FALSE)
```

## **Plots**
faith pd plot
- create a plot without stats and one with stats
```{r, fig.width=19, fig.height=7}
## plot without stats
## y axis scales fixed 
(faith_plot <- alphaDiv_seq_meta %>% 
  ggplot(aes(x = as.factor(new_visit_num), y = faith_pd)) +
  geom_boxplot(aes(group = new_visit_num), outlier.shape = NA) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.4, size = 2.5) +
  scale_x_discrete(labels = visit_x_labs) +
  theme_bw(base_size = 20) +
  facet_wrap(~treatment_group,
             labeller = labeller(treatment_group = treat_group_labs)) +
  labs(x = 'Visit',
       y = "Faith's PD",
       title = "Alpha Diversity: Faith's PD"))

## plot with stats
## y axis scales free so not all points are smushed near the bottom
(faith_plot_wStats <- faith_plot +
    facet_wrap(~treatment_group,
             labeller = labeller(treatment_group = treat_group_labs),
             scales = "fixed") +
    stat_pvalue_manual(faithPD_art_lmerStrat_contrast,
                       step.group.by = 'treatment_group',
                       step.increase = 0.1,
                       hide.ns = TRUE,
                       label = 'p.adj.signif'))
```

shannon entropy plot
```{r, fig.width=19, fig.height=7}
(shannon_plot <- alphaDiv_seq_meta %>% 
  ggplot(aes(x = as.factor(new_visit_num), y = shannon_entropy)) +
  geom_boxplot(aes(group = new_visit_num), outlier.shape = NA) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.4, size = 2.5) +
  scale_x_discrete(labels = visit_x_labs) +
  theme_bw(base_size = 20) +
  facet_grid(~treatment_group,
             labeller = labeller(treatment_group = treat_group_labs)) +
  labs(x = 'Visit',
       y = "Shannon Entropy",
       title = "Alpha Diversity: Shannon Entropy"))
```

## **Saving my outputs**
```{r}
save(list = c('faith_plot_wStats'),
       file = '../../../../manuscript_figures/ggplots/faith_plot.rdat')
```

```{r}
## pdf version
ggsave('../plots/faith_pd_plot.pdf',
       plot = faith_plot,
       width = 19,
       height = 7)
ggsave('../plots/faith_pd_wStats_plot.pdf',
       plot = faith_plot_wStats,
       width = 19,
       height = 7)
ggsave('../plots/shannon_plot.pdf',
       plot = shannon_plot,
       width = 19,
       height = 7)

## png version
ggsave('../plots/pngs/faith_pd_plot.png',
       plot = faith_plot,
       width = 19,
       height = 7)
ggsave('../plots/pngs/faith_pd_wStats_plot.png',
       plot = faith_plot_wStats,
       width = 19,
       height = 7)
ggsave('../plots/pngs/shannon_plot.png',
       plot = shannon_plot,
       width = 19,
       height = 7)

## stats
write_tsv(faithPD_art_lmerStrat_res,
          '../stats/faithPD_treatGroup_strat.tsv')
write_tsv(faithPD_art_lmerStrat_contrast,
          '../stats/faithPD_treatGroup_strat_postHoc.tsv')
write_tsv(faithPD_art_lmer_notStrat_res,
          '../stats/faithPD_overall.tsv')
write_tsv(faithPD_art_lmer_notStrat_contrast,
          '../stats/faithPD_overall_postHoc.tsv')
```

