---
title: "taxonomy_stats"
output: html_document
date: "2025-02-25"
author: "Madi Apgar"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(magrittr)
library(broom)
library(qiime2R)
library(apppleplots)
library(ggpubr)
library(rstatix)
library(cgwtools)
```

## **Functions**
`source()` pulls in functions placed in other R scripts and saves them to your global environment!
- here i'm pulling in the statistical analysis functions:
  - `test_normality()` - does basic R tests for normality on the data
  - `art_lmer_notStrat()` - runs aligned rank transform lmer on the data in an overall model (not stratified)
  - `art_lmer_strat()` - runs aligned rank transform lmer on the data in a variable stratified model
  - `art_lmer_doubleStrat()` - runs aligned rank transform lmer on the data in a model stratified by two variables
```{r}
source("../../../art_lmer_functions.R")
source("../../../normality_function.R")
source("../../../diff_between_means_calc.R")

abun_file_prep <- function(metadata_fp,
                           tax_fp,
                           otu_table_fp,
                           group_by_cols,
                           tax_levels,
                           filter_tax_by = NULL,
                           filter_tax_col = NULL){
  ## metadata
  metadata_file <- read_tsv(metadata_fp) %>%
    select(anonymized_name, host_subject_id, visit, visit_number, treatment_group) %>%
    rename(sampleid = anonymized_name,
          subject_id = host_subject_id,
          new_visit_label = visit,
          new_visit_num = visit_number)
  ## taxonomy
  taxonomy <- read_qza(tax_fp)$data %>% 
    parse_taxonomy() %>% 
    rownames_to_column('asv')
  ## otu table 
  otu_table <- read_qza(otu_table_fp)$data
  
  otu_table <- otu_table %>% 
    as_tibble(rownames = 'asv') %>% 
    gather(-asv, key = sampleid, value = abun) %>% 
    group_by(sampleid) %>% 
    mutate(rel_abun = abun/sum(abun)) %>% 
    mutate(rel_abun = rel_abun + 0.000001) %>% 
    filter(sampleid %in% metadata_file$sampleid)
  ## joining all tables together 
  wTax_level_groupBy_cols <- append(group_by_cols, tax_levels)
  
  abun_table <- otu_table %>% 
    left_join(metadata_file, by = 'sampleid') %>% 
    left_join(taxonomy, by = 'asv') %>% 
    group_by(across(all_of(wTax_level_groupBy_cols))) %>%
    summarise(rel_abund = sum(rel_abun)) %>% 
    ungroup()
  
  if (is.character(filter_tax_by) & is.character(filter_tax_col)){
    abun_table <- abun_table %>% 
      filter(.data[[filter_tax_col]] %in% filter_tax_by)
  } else {
    abun_table
  }
  
  ## creating a list for my outputs
  my_list <- list(Taxonomy = taxonomy,
                  OTUTable = otu_table,
                  AbundanceTable = abun_table)
  return(my_list)
}

## small function to assign a value to a key in a dataframe column (i.e. numbers to actual categorical variables)
values_to_keys <- function(input_table,
                           key_col_name,
                           value_col_name,
                           key_value_list,
                           col_order_list){
  
  input_table[value_col_name] <- unlist(lapply(unlist(input_table[key_col_name]), 
                                        FUN = function(x){key_value_list[[x]]}))
  
  output_table <- input_table %>%
    select(all_of(col_order_list), everything())
  
  return(output_table)
}
```

## **Reading in files and data wrangling**

**File paths**
```{r}
tax_FP <- '../data/taxonomy.qza'
otu_table_FP <- '../data/taxOnly_otu_table.qza'
proc_seq_meta_fp <- '../../../../data_submission/1_16S_rRNA_qiita/CRS3123_fecal_samples/crestone_qiita_sample_information.tsv'
vre_data_fp <- '../data/vre_data.csv'

## global lists for renaming plot labels
visit_x_labs <- c('1' = 'Screen',
                  '2' = 'D1',
                  '3' = 'D4',
                  '4' = 'TOC',
                  '5' = 'D17', 
                  '6' = 'D40', 
                  '7' = 'D70') 

treat_group_labs <- c('CRS3123 200mg',
                      'CRS3123 400mg',
                      'Vancomycin 125mg')
names(treat_group_labs) <- c('CRS3123_200mg_po_bid',
                             'CRS3123_400mg_po_bid',
                             'Vancomycin_125mg_po_qid')

special_treatGroup_x_labs <- c('CRS3123\n200mg',
                               'CRS3123\n400mg',
                               'Vancomycin\n125mg')

genus_labs <- c('Clostridioides',
                'Escherichia-Shigella',
                'Enterococcus',
                'Bifidobacterium',
                'Bacteroides',
                'Lactobacillus')
names(genus_labs) <- c(1, 2, 3, 4, 5, 6)
```

**Prepping abundance table**
```{r}
metadata <- read_tsv(proc_seq_meta_fp) %>%
  select(anonymized_name, host_subject_id, visit, visit_number, treatment_group) %>%
  rename(sampleid = anonymized_name,
         subject_id = host_subject_id,
         new_visit_label = visit,
         new_visit_num = visit_number)

## only taking VRE data for subjects/visits that have positive reads for culture AND disk diffusion
## 14/22 samples had positives for both culture and disk diffusion
vre_df <- read_csv(vre_data_fp)

vre_df <- vre_df %>% 
  filter(vre_culture_status == 'Positive') %>% 
  select(subject_id, visit_label, vre_culture_status) %>%
  rename(new_visit_label = visit_label)

abun_col_list <- c('sampleid',
                   'subject_id',
                   'new_visit_num',
                   'new_visit_label',
                   'treatment_group')

wanted_family <- c('Enterobacteriaceae')

wanted_genus <- c('Bifidobacterium',
                  'Blautia',
                  'Enterococcus',
                  'Clostridioides',
                  'Lactobacillus',
                  'Bacteroides',
                  'Escherichia-Shigella')

## all genera table
abun_files <- abun_file_prep(metadata_fp = proc_seq_meta_fp,
                             tax_fp = tax_FP,
                             otu_table_fp = otu_table_FP,
                             group_by_cols = abun_col_list,
                             tax_levels = "Genus",
                             filter_tax_by = wanted_genus,
                             filter_tax_col = 'Genus')

proc_abun_table <- abun_files$AbundanceTable

(proc_abun_table <- proc_abun_table %>% 
  na.omit() %>% 
  mutate(new_visit_num = as.numeric(new_visit_num),
         genus_order = case_when(
           Genus == 'Clostridioides' ~ 1,
           Genus == 'Escherichia-Shigella' ~ 2,
           Genus == 'Enterococcus' ~ 3,
           Genus == 'Bifidobacterium' ~ 4,
           Genus == 'Bacteroides' ~ 5,
           Genus == 'Blautia' ~ 6,
           Genus == 'Lactobacillus' ~ 7
         )) %>% 
  left_join(vre_df, by = c('subject_id', 'new_visit_label')) %>% 
  mutate(vre_culture_status = ifelse(is.na(vre_culture_status), "Negative", vre_culture_status)))

## just enterobacteriaceae table
entero_abun_files <- abun_file_prep(metadata_fp = proc_seq_meta_fp,
                                    tax_fp = tax_FP,
                                    otu_table_fp = otu_table_FP,
                                    group_by_cols = abun_col_list,
                                    tax_levels = c("Family", "Genus"),
                                    filter_tax_by = wanted_family,
                                    filter_tax_col = 'Family')

proc_entero_abun_table <- entero_abun_files$AbundanceTable

proc_entero_abun_table <- proc_entero_abun_table %>%  
  na.omit() %>% 
  mutate(new_visit_num = as.numeric(new_visit_num))

## looking at our butyrate producers 
##"Ruminococcaceae", "Oscillospiraceae"
buty_abun_files <- abun_file_prep(metadata_fp = proc_seq_meta_fp,
                                    tax_fp = tax_FP,
                                    otu_table_fp = otu_table_FP,
                                    group_by_cols = abun_col_list,
                                    tax_levels = c("Family"),
                                    filter_tax_by = c("Lachnospiraceae", "Oscillospiraceae", "Ruminococcaceae"),
                                    filter_tax_col = 'Family')

proc_buty_abun_table <- buty_abun_files$AbundanceTable

proc_buty_abun_table <- proc_buty_abun_table %>% 
  na.omit() %>% 
  mutate(new_visit_num = as.numeric(new_visit_num),
         log_rel_abund = log10(rel_abund))


## just ruminococcus gnavus table
crestWanted_abun_files <- abun_file_prep(metadata_fp = proc_seq_meta_fp,
                                    tax_fp = tax_FP,
                                    otu_table_fp = otu_table_FP,
                                    group_by_cols = abun_col_list,
                                    tax_levels = c("Genus"),
                                    filter_tax_by = c("Ruminococcus", "Enterococcus", "Lachnoclostridium"),
                                    filter_tax_col = 'Genus')

crestWanted_abun_table <- crestWanted_abun_files$AbundanceTable

proc_crestWanted_abun_table <- crestWanted_abun_table %>%
  mutate(rel_abund = rel_abund - 0.000001,
         percent_relative_abundance = rel_abund*100) %>% 
  select(sampleid, subject_id, treatment_group, new_visit_label, Genus, rel_abund, percent_relative_abundance) %>% 
  rename(visit = new_visit_label,
         relative_abundance = rel_abund)
```

**calculating alpha diversity rarefaction curve**
```{r}
rare_otu_table <- read_qza(otu_table_FP)$data

## transposing so rows are sampleids 
rare_otu_table <- t(rare_otu_table)
rare_otu_table <- rare_otu_table[rownames(rare_otu_table) %in% metadata$sampleid, ]

vegan::rarecurve(rare_otu_table, step = 1000, label = T)
```


## **Relative abundance over time plots**

**All genera plot**
```{r, fig.width=17, fig.height=15}
(genusAbun_plot <- apppleplots::facet_twice_plots(input_table = proc_abun_table,
                               x_value = 'new_visit_num',
                               y_value = 'rel_abund',
                               x_value_type = 'discrete',
                               y_transform = TRUE,
                               x_labels = visit_x_labs,
                               box_group_by = 'new_visit_num',
                               line_group_by = 'subject_id',
                               point_alpha = 0,
                               facet_rows = 'genus_order',
                               facet_cols = 'treatment_group',
                               row_labs = genus_labs,
                               col_labs = treat_group_labs,
                               x_name = 'Study Day',
                               y_name = 'Relative Abundance (log10)',
                               title_content = 'Select Genera Relative Abundance') +
  scale_x_continuous(breaks = c(1, 2, 3, 4, 5, 6, 7),
                     labels = visit_x_labs) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.4, size = 2.5) +
  theme_bw(base_size = 20) +
  theme(strip.text.y.right = element_text(angle = 0)))
```

**Just Enterobacteriaceae plot**
```{r, fig.width=17, fig.height=10}
(entero_genusAbun_plot <- apppleplots::facet_twice_plots(input_table = proc_entero_abun_table,
                               x_value = 'new_visit_num',
                               y_value = 'rel_abund',
                               x_value_type = 'discrete',
                               y_transform = TRUE,
                               x_labels = visit_x_labs,
                               box_group_by = 'new_visit_num',
                               line_group_by = 'subject_id',
                               point_alpha = 0,
                               facet_rows = 'Genus',
                               facet_cols = 'treatment_group',
                               row_labs = NULL,
                               col_labs = treat_group_labs,
                               x_name = 'Study Day',
                               y_name = 'Relative Abundance (log10)',
                               title_content = 'Enterobacteriaceae Relative Abundance') +
  scale_x_continuous(breaks = c(1, 2, 3, 4, 5, 6, 7),
                     labels = visit_x_labs) +
  # scale_y_continuous(trans = 'log10',
  #                    labels = scales::comma) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.4, size = 2.5) +
  theme_bw(base_size = 20) +
  theme(strip.text.y.right = element_text(angle = 0)))
```

**Just butyrate producers plot**
everyone is losing a lot of lachnoclostridium, lachnospira, and lachnospiraceae groups w abx 
```{r, fig.width=15, fig.height=12}
test_relabun <- proc_buty_abun_table %>% 
  filter(Family == "Lachnospiraceae") %>% 
  ungroup() %>% 
  group_by(new_visit_num, treatment_group, Genus) %>% 
  mutate(mean_rel_abun = mean(rel_abund)) %>% 
  distinct(new_visit_num, treatment_group, Genus, mean_rel_abun, sum_rel_abun)


(av_lachno_genera_plot <- test_relabun %>% 
  ggplot(aes(x = new_visit_num, y = mean_rel_abun)) +
  geom_col(aes(fill = Genus), alpha = 0.8) +
  theme_bw(base_size = 20) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 13),
        legend.title.position = "top") +
  facet_wrap(~treatment_group,
             labeller = labeller(treatment_group = treat_group_labs)) +
  scale_x_continuous(breaks = c(1, 2, 3, 4, 5, 6, 7),
                     labels = visit_x_labs) +
  labs(x = 'Study Day',
       y = 'Average Relative Abundance',
       title = 'Lachnospiraceae Genera'))
```

```{r, fig.width=17, fig.height=14}
## stats
## prep dataframe
statsPrep_butyAbun_table <- proc_buty_abun_table %>% 
    convert_as_factor(subject_id, new_visit_num, treatment_group)
  
buty_art_lmer_strat <- art_lmer_doubleStrat(input_table = statsPrep_butyAbun_table,
                                      strat_variable1 = 'Family',
                                      strat_variable2 = 'treatment_group',
                                      model_formula_left = 'log_rel_abund',
                                      model_formula_right = 'new_visit_num + (1|subject_id)',
                                      model_contrast_run_by = 'new_visit_num')

buty_strat_contrast <- buty_art_lmer_strat$ARTContrast %>% 
     rename(p.adj.signif = signif) %>% 
     mutate(across(all_of(c('group1', 'group2')), ~gsub(., pattern = "new_visit_num*", replacement = "")),
            y.position = case_when(
              Family == 'Lachnospiraceae' ~ as.numeric(1),
              Family == 'Oscillospiraceae' ~ as.numeric(0.02),
              Family == 'Ruminococcaceae' ~ as.numeric(0.3)))

## plot
(buty_abun_plot <- apppleplots::facet_twice_plots(input_table = proc_buty_abun_table,
                               x_value = 'new_visit_num',
                               y_value = 'rel_abund',
                               x_value_type = 'discrete',
                               y_transform = TRUE,
                               x_labels = visit_x_labs,
                               box_group_by = 'new_visit_num',
                               line_group_by = 'subject_id',
                               point_alpha = 0,
                               facet_rows = 'Family',
                               facet_cols = 'treatment_group',
                               row_labs = NULL,
                               col_labs = treat_group_labs,
                               x_name = 'Study Day',
                               y_name = 'Relative Abundance (log10)',
                               title_content = 'Butyrate-Producing Families Relative Abundance') +
  geom_jitter(width = 0.1, height = 0, alpha = 0.4, size = 2.5) +
  facet_grid(rows = vars(Family),
             cols = vars(treatment_group),
             scales = "free_y",
             # space = "free_y",
             labeller = labeller(.cols = treat_group_labs,
                                 .rows = NULL)) +
  theme_bw(base_size = 20) +
  theme(strip.text.y.right = element_text(angle = 0)) +
  stat_pvalue_manual(buty_strat_contrast,
                     step.group.by = c('Family', 'treatment_group'),
                     step.increase = 0.07,
                     bracket.nudge.y = -0.7,
                     tip.length = 0.01,
                     hide.ns = TRUE,
                     label = 'p.adj.signif'))
```

butyrate stats between treat groups at each visit 
```{r}
(buty_treatGroup_strat_contrast <- art_lmer_doubleStrat(input_table = statsPrep_butyAbun_table,
                                      strat_variable1 = 'Family',
                                      strat_variable2 = 'new_visit_num',
                                      model_formula_left = 'log_rel_abund',
                                      model_formula_right = 'treatment_group',
                                      model_contrast_run_by = 'treatment_group')$ARTContrast)
```



**Just ruminococcus_gnavus_group plot**
```{r, fig.width=16, fig.height=5}
(ruminoG_genusAbun_plot <- apppleplots::facet_twice_plots(input_table = proc_ruminoG_abun_table,
                               x_value = 'new_visit_num',
                               y_value = 'rel_abund',
                               x_value_type = 'discrete',
                               y_transform = TRUE,
                               x_labels = visit_x_labs,
                               box_group_by = 'new_visit_num',
                               line_group_by = 'subject_id',
                               point_alpha = 0,
                               facet_rows = 'Genus',
                               facet_cols = 'treatment_group',
                               row_labs = NULL,
                               col_labs = treat_group_labs,
                               x_name = 'Study Day',
                               y_name = 'Relative Abundance (log10)',
                               title_content = 'Ruminococcus gnavus Relative Abundance') +
  scale_x_continuous(breaks = c(1, 2, 3, 4, 5, 6, 7),
                     labels = visit_x_labs) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.4, size = 2.5) +
  theme_bw(base_size = 20) +
  theme(strip.text.y.right = element_text(angle = 0)))
```

## **Individual genus plots (for loop)**
writing a for loop for this bc I'm lazy 
LOG10 TRANSFORMED Y-AXIS PLOTS !!
```{r, fig.width=15, fig.height=5.5}
genus_plots_log10 <- c('Clostridioides',
                        'Bifidobacterium',
                        'Blautia',
                        'Bacteroides',
                        'Enterococcus',
                        'Lactobacillus',
                        'Escherichia-Shigella')
all_genus_stats <- tibble()
for (genus in unique(unlist(genus_plots_log10))) {
  ## filter table to only include wanted genus
  genus_filt_df <- proc_abun_table %>% 
    filter(Genus == genus) %>% 
    mutate(log_rel_abund = log10(rel_abund)) %>% 
    convert_as_factor(subject_id, new_visit_num, treatment_group)
  
  ## running stats on changes in genus rel abun between visits within treatment groups for plots
  perGenus_art_lmer_strat <- art_lmer_strat(input_table = genus_filt_df,
                                           strat_variable = 'treatment_group',
                                           model_formula_left = 'log_rel_abund',
                                           model_formula_right = 'new_visit_num + (1|subject_id)',
                                           model_contrast_run_by = 'new_visit_num')

  perGenus_strat_contrast <- perGenus_art_lmer_strat$ARTContrast %>% 
     rename(p.adj.signif = signif) %>% 
     mutate(across(all_of(c('group1', 'group2')), ~gsub(., pattern = "new_visit_num*", replacement = "")),
            y.position = as.numeric(0.0002),
            Genus = genus)
  
  all_genus_stats <- rbind(all_genus_stats, perGenus_strat_contrast)
  
  ## use genus to put together variable name and plot title so I can save it to that 
  variable_name <- paste(tolower(genus), "only_plot", sep = "_")
  plot_title <- paste(genus, "Relative Abundance", sep = " ")

  assign(variable_name,
         apppleplots::facet_twice_plots(input_table = genus_filt_df,
                               x_value = 'new_visit_num',
                               y_value = 'rel_abund',
                               x_value_type = 'discrete',
                               y_transform = TRUE,
                               x_labels = visit_x_labs,
                               box_group_by = 'new_visit_num',
                               line_group_by = 'subject_id',
                               point_alpha = 0,
                               facet_rows = NULL,
                               facet_cols = 'treatment_group',
                               row_labs = NULL,
                               col_labs = treat_group_labs,
                               x_name = 'Study Day',
                               y_name = 'Relative Abundance (log10)',
                               title_content = plot_title) +
        geom_jitter(width = 0.1, height = 0, alpha = 0.4, size = 2.5) +
        theme_bw(base_size = 20) +
        theme(strip.text.y.right = element_text(angle = 0)) +
        stat_pvalue_manual(perGenus_strat_contrast,
                           step.group.by = c('treatment_group'),
                           step.increase = 0.1,
                           hide.ns = TRUE,
                           label = 'p.adj.signif')
          )
}

bacteroides_only_plot
bifidobacterium_only_plot
blautia_only_plot
clostridioides_only_plot
enterococcus_only_plot
`escherichia-shigella_only_plot`
lactobacillus_only_plot
```

NOT LOG10 TRANSFORMED Y-AXIS PLOT !!
```{r, fig.width=15, fig.height=5}
genus_plots_notLog10 <- c('Bifidobacterium',
                          'Blautia')

for (genus in unique(unlist(genus_plots_notLog10))) {
  ## filter table to only include wanted genus
  genus_filt_df <- proc_abun_table %>% 
    filter(Genus == genus)
  
  ## use genus to put together variable name and plot title so I can save it to that 
  variable_name <- paste(tolower(genus), "notLog10_plot", sep = "_")
  plot_title <- paste(genus, "Relative Abundance", sep = " ")
  
  assign(variable_name,
         apppleplots::facet_twice_plots(input_table = genus_filt_df,
                               x_value = 'new_visit_num',
                               y_value = 'rel_abund',
                               x_value_type = 'discrete',
                               y_transform = FALSE,
                               x_labels = visit_x_labs,
                               box_group_by = 'new_visit_num',
                               line_group_by = 'subject_id',
                               point_alpha = 0,
                               facet_rows = NULL,
                               facet_cols = 'treatment_group',
                               row_labs = NULL,
                               col_labs = treat_group_labs,
                               x_name = 'Study Day',
                               y_name = 'Relative Abundance',
                               title_content = plot_title) +
        scale_x_continuous(breaks = c(1, 2, 3, 4, 5, 6, 7),
                           labels = visit_x_labs) +
        # scale_y_continuous(trans = 'log10',
        #                    labels = scales::comma) +
        geom_jitter(width = 0.1, height = 0, alpha = 0.4, size = 2.5) +
        theme_bw(base_size = 20) +
        theme(strip.text.y.right = element_text(angle = 0))
        )
}

## not log10 transformed plots
bifidobacterium_notLog10_plot
blautia_notLog10_plot
```

## **Select genera plots (for paper) - includes VRE data**
everything but lactobacillus basically 
- take enterococcus out and include separately with VRE data??
- split plots up into "good guys" and "bad guys"

**Bad guys**
```{r, fig.width=17, fig.height=9}
bad_genera <- c('Clostridioides',
                'Escherichia-Shigella')

bad_abun_table <- proc_abun_table %>% 
  filter(Genus %in% bad_genera)

bad_genus_stats <- all_genus_stats %>% 
  filter(Genus %in% bad_genera) %>% 
  mutate(genus_order = case_when(
           Genus == 'Clostridioides' ~ 1,
           Genus == 'Escherichia-Shigella' ~ 2
         ))
 
(fig3a_plot <- apppleplots::facet_twice_plots(input_table = bad_abun_table,
                               x_value = 'new_visit_num',
                               y_value = 'rel_abund',
                               x_value_type = 'discrete',
                               y_transform = TRUE,
                               x_labels = visit_x_labs,
                               box_group_by = 'new_visit_num',
                               line_group_by = 'subject_id',
                               point_alpha = 0,
                               facet_rows = 'genus_order',
                               facet_cols = 'treatment_group',
                               row_labs = genus_labs,
                               col_labs = treat_group_labs,
                               x_name = 'Visit',
                               y_name = 'Relative Abundance (log10)',
                               title_content = 'Relative Abundance of Pathogenic Genera') +
        geom_jitter(width = 0.1, height = 0, alpha = 0.4, size = 2.5) +
        theme_bw(base_size = 20) +
        theme(strip.text.y.right = element_text(angle = 0)) +
        stat_pvalue_manual(bad_genus_stats,
                           step.group.by = c('treatment_group', 'genus_order'),
                           step.increase = 0.06,
                           hide.ns = TRUE,
                           label = 'p.adj.signif'))
```

**VRE data plot (for paper)**
120-2204 - 
  VRE: screen, TOC, D17  
  16S: screen, D1, TOC, D40
128-1002 - 
  VRE: screen, D4  
  16S: screen, D1, D4, D17, D40, D70
116-1006 - 
  VRE: D17    
  16S: screen, D1, TOC, D40
122-1001 - 
  VRE: screen, D1, D4  
  16S: screen, D1, TOC, D17, D40, D70
122-1002 - 
  VRE: screen, D1, D4  
  16S: screen, D1, D4, TOC, D17, D40, D70
130-1001 - 
  VRE: D1, D4, TOC, D17  
  16S: D1, D4, TOC, D17, D40, D70
```{r, fig.width=15.5, fig.height=5.5}
vre_abun_table <- proc_abun_table %>% 
  filter(Genus == 'Enterococcus')

vre_genus_stats <- all_genus_stats %>% 
  filter(Genus == 'Enterococcus')

vre_color_pal <- c("#CC4678FF",
                   "#999999")

(fig3b_plot <- apppleplots::facet_twice_plots(input_table = vre_abun_table,
                               x_value = 'new_visit_num',
                               y_value = 'rel_abund',
                               x_value_type = 'discrete',
                               y_transform = TRUE,
                               x_labels = visit_x_labs,
                               box_group_by = 'new_visit_num',
                               line_group_by = 'subject_id',
                               point_alpha = 0,
                               facet_rows = 'Genus',
                               facet_cols = 'treatment_group',
                               row_labs = NULL,
                               col_labs = treat_group_labs,
                               x_name = 'Visit',
                               y_name = 'Relative Abundance (log10)',
                               title_content = 'Enterococcus') +
        geom_jitter(aes(fill = vre_culture_status, alpha = vre_culture_status), 
                    width = 0.1, height = 0.1, size = 2.5, pch = 21) +
        scale_fill_manual(breaks = c("Positive",
                                     "Negative"),
                          labels = c("Positive",
                                     "Negative"),
                          values = vre_color_pal,
                          name = "VRE Status") +
        scale_alpha_manual(values = c(0.4, 1)) +
        guides(alpha = "none") +
        theme_bw(base_size = 20) +
        theme(strip.text.y.right = element_text(angle = 0)) +
        stat_pvalue_manual(vre_genus_stats,
                           step.group.by = c('treatment_group'),
                           step.increase = 0.06,
                           hide.ns = TRUE,
                           label = 'p.adj.signif') +
        labs(subtitle = 'Subject VRE Presence'))
```

**Good guys**
```{r, fig.width=17, fig.height=9}
good_genera <- c('Bacteroides',
                 'Bifidobacterium')

good_abun_table <- proc_abun_table %>% 
  filter(Genus %in% good_genera)

good_genus_stats <- all_genus_stats %>% 
  filter(Genus %in% good_genera)

(fig3c_plot <- apppleplots::facet_twice_plots(input_table = good_abun_table,
                               x_value = 'new_visit_num',
                               y_value = 'rel_abund',
                               x_value_type = 'discrete',
                               y_transform = TRUE,
                               x_labels = visit_x_labs,
                               box_group_by = 'new_visit_num',
                               line_group_by = 'subject_id',
                               point_alpha = 0,
                               facet_rows = 'Genus',
                               facet_cols = 'treatment_group',
                               row_labs = NULL,
                               col_labs = treat_group_labs,
                               x_name = 'Visit',
                               y_name = 'Relative Abundance (log10)',
                               title_content = 'Relative Abundance of Beneficial Genera') +
        geom_jitter(width = 0.1, height = 0, alpha = 0.4, size = 2.5) +
        theme_bw(base_size = 20) +
        theme(strip.text.y.right = element_text(angle = 0)) +
        stat_pvalue_manual(good_genus_stats,
                           step.group.by = c('treatment_group', 'Genus'),
                           step.increase = 0.06,
                           inherit.aes = FALSE,
                           hide.ns = TRUE,
                           label = 'p.adj.signif') +
    scale_x_discrete(limits = c(1, 2, 3, 4, 5, 6, 7),
                     labels = visit_x_labs))
```

## **ART LMER Stats**

**Not stratified analysis**
```{r}
factored_abun_table <- proc_abun_table %>% 
  select(sampleid, subject_id, new_visit_num, Genus, rel_abund, treatment_group) %>% 
  convert_as_factor(subject_id, new_visit_num, Genus, treatment_group)

genusAbun_art_lmer <- art_lmer_notStrat(input_table = factored_abun_table,
                                        formula_left = 'rel_abund',
                                        formula_right = 'new_visit_num * Genus * treatment_group + (1|subject_id)',
                                        contrast_run_by = 'new_visit_num:Genus:treatment_group')

(genusAbun_notStrat_res <- genusAbun_art_lmer$ARTResults)
(genusAbun_notStrat_contrast <- genusAbun_art_lmer$ARTContrast)
```

**Stratified by treatment group**
```{r}
genusAbun_art_lmer_strat <- art_lmer_strat(input_table = factored_abun_table,
                                           strat_variable = 'treatment_group',
                                           model_formula_left = 'rel_abund',
                                           model_formula_right = 'new_visit_num * Genus + (1|subject_id)',
                                           model_contrast_run_by = 'new_visit_num:Genus')

(genusAbun_strat_res <- genusAbun_art_lmer_strat$ARTResults)
(genusAbun_strat_contrast <- genusAbun_art_lmer_strat$ARTContrast)
```

**Double stratified by genus and visit**
dont need to add subject_id as a random effect to this model since I'm running the stats between relative abundance of the genera individually for each day and treatment group (so you only have one sample per subject bc its not across the whole timeline)
```{r}
genusAbun_art_lmer_doubleStrat <- art_lmer_doubleStrat(input_table = factored_abun_table,
                                                       strat_variable1 = 'Genus',
                                                       strat_variable2 = 'new_visit_num',
                                                       model_formula_left = 'rel_abund',
                                                       model_formula_right = 'treatment_group',
                                                       model_contrast_run_by = 'treatment_group')

(genusAbun_doubleStrat_res <- genusAbun_art_lmer_doubleStrat$ARTResults)
(genusAbun_doubleStrat_contrast <- genusAbun_art_lmer_doubleStrat$ARTContrast)
```

creating a column in the stat results that translates visit number to study day for ease of understanding
```{r}
visit_key_value <- c('1' = 'Screen',
                     '2' = 'D1',
                     '3' = 'D4',
                     '4' = 'D10-15/TOC',
                     '5' = 'D17',
                     '6' = 'D40',
                     '7' = 'D70')
visit_col_order <- c('Genus', 
                     'new_visit_num', 
                     'visit_label')

genusAbun_doubleStrat_res_wValues <- values_to_keys(input_table = genusAbun_doubleStrat_res,
                                                    key_col_name = 'new_visit_num',
                                                    value_col_name = 'visit_label',
                                                    key_value_list = visit_key_value,
                                                    col_order_list = visit_col_order)

genusAbun_doubleStrat_contrast_wValues <- values_to_keys(input_table = genusAbun_doubleStrat_contrast,
                                                         key_col_name = 'new_visit_num',
                                                         value_col_name = 'visit_label',
                                                         key_value_list = visit_key_value,
                                                         col_order_list = visit_col_order)

## making contrast results pretty to give to them
proc_full_genusAbun_doubleStrat_contrast_wValues <- genusAbun_doubleStrat_contrast_wValues %>% 
  select(Genus, visit_label, new_visit_num, group1, group2, p.value, signif) %>% 
  rename(visit = visit_label) 

proc_genusAbun_doubleStrat_contrast_wValues <- proc_full_genusAbun_doubleStrat_contrast_wValues %>% 
  filter(signif != 'ns')
```

**Stat heat plots**
for differences in relative abundances of genera/butyrate-producing families at a given time point between treatment groups 

genera differences 
```{r, fig.width=15, fig.height=8}
## calculating difference bt means to give directionality
genusAbun_treatGroup_strat_diffs <- calc_diff_means(input_table = factored_abun_table,
                                                    first_group = 'treatment_group',
                                                    second_group = c('Genus', 'new_visit_num'),
                                                    mean_value = 'rel_abund',
                                                    post_hoc_test = proc_full_genusAbun_doubleStrat_contrast_wValues) %>% 
  filter(!(Genus %in% c('Blautia', 'Lactobacillus')))

(genusAbun_treatGroup_strat_heatPlot <- apppleplots::stat_heat_plot(input_table = genusAbun_treatGroup_strat_diffs,
                            x_value = 'new_visit_num',
                            y_value = 'group2',
                            x_value_type = 'discrete',
                            x_labels = visit_x_labs,
                            y_value_type = 'discrete',
                            y_labels = treat_group_labs,
                            tile_fill = 'stat_diff_means',
                            text_value = 'signif',
                            legend_name = "Group 1 -\nGroup 2",
                            facet_by = 'Genus',
                            facet_labels = NULL,
                            x_name = 'Visit',
                            y_name = 'Group 2',
                            title_content = 'Treatment Genera 16S Relative Abundance Differences',
                            pairwise = TRUE,
                            low_color = "#0D0887FF",
                            high_color = "#CC4678FF") +
  facet_grid(Genus~group1,
             labeller = labeller(group1 = treat_group_labs)) +
  theme_bw(base_size = 20) +
  theme(strip.text.y.right = element_text(angle = 0),
        plot.subtitle = element_text(hjust = 0.5)))
```

butyrate-producing family differences 
```{r, fig.width=15, fig.height=6}
## calculating difference bt means to give directionality
buty_treatGroup_strat_diffs <- calc_diff_means(input_table = statsPrep_butyAbun_table,
                                               first_group = 'treatment_group',
                                               second_group = c('Family', 'new_visit_num'),
                                               mean_value = 'log_rel_abund',
                                               post_hoc_test = buty_treatGroup_strat_contrast)

## stat heat plot!!
(buty_treatGroup_strat_heatPlot <- apppleplots::stat_heat_plot(input_table = buty_treatGroup_strat_diffs,
                            x_value = 'new_visit_num',
                            y_value = 'group2',
                            x_value_type = 'discrete',
                            x_labels = visit_x_labs,
                            y_value_type = 'discrete',
                            y_labels = treat_group_labs,
                            tile_fill = 'stat_diff_means',
                            text_value = 'signif',
                            legend_name = "Group 1 -\nGroup 2",
                            facet_by = 'Family',
                            facet_labels = NULL,
                            x_name = 'Visit',
                            y_name = 'Group 2',
                            title_content = 'Treatment Butyrate-Producer Family 16S Relative Abundance Differences',
                            pairwise = TRUE,
                            low_color = "#0D0887FF",
                            high_color = "#CC4678FF") +
  facet_grid(Family~group1,
             labeller = labeller(group1 = treat_group_labs)) +
  theme_bw(base_size = 20) +
  theme(strip.text.y.right = element_text(angle = 0),
        plot.subtitle = element_text(hjust = 0.5)))
```

## **Saving my outputs**

**crestone wanted genera**
```{r}
write_tsv(proc_crestWanted_abun_table,
          file = '../data/16s_enteroRuminoLachnoclos_relAbun_data.tsv')

## saving processed butyrate producers table for correlation comparisons 
write_tsv(proc_buty_abun_table,
          file = '../data/16s_butyProd_relAbun.tsv')
```


**saving figures .rdat file**
```{r}
save(list = c('fig3a_plot',
              'fig3b_plot',
              'fig3c_plot',
              'buty_abun_plot',
              'genusAbun_treatGroup_strat_heatPlot',
              'buty_treatGroup_strat_heatPlot'),
      file = "../../../../manuscript_figures/ggplots/taxAbun_plots.rdat")
```

**pdf versions of plots** 
```{r}
ggsave('../plots/genus_relAbun_plot.pdf',
       plot = genusAbun_plot,
       width = 17,
       height = 13)
ggsave('../plots/enterobacteriaceae_relAbun_plot.pdf',
       plot = entero_genusAbun_plot,
       width = 17,
       height = 7)
ggsave('../plots/enterococcusxVRE_relAbun_plot.pdf',
       plot = fig3b_plot,
       width = 15.5,
       height = 5.5)
ggsave('../plots/avLachnoGen_relAbun_plot.pdf',
       plot = av_lachno_genera_plot,
       width = 15,
       height = 12)

## individual genus rel abun plots
ggsave('../plots/bacteroides_relAbun_plot.pdf',
       plot = bacteroides_only_plot,
       width = 15,
       height = 5.5)
ggsave('../plots/bifidobacterium_relAbun_plot.pdf',
       plot = bifidobacterium_only_plot,
       width = 15,
       height = 5.5)
ggsave('../plots/bifido_notLog10_relAbun_plot.pdf',
       plot = bifidobacterium_notLog10_plot,
       width = 15,
       height = 5)
ggsave('../plots/blautia_relAbun_plot.pdf',
       plot = blautia_only_plot,
       width = 15,
       height = 5.5)
ggsave('../plots/blautia_notLog10_relAbun_plot.pdf',
       plot = blautia_notLog10_plot,
       width = 15,
       height = 5)
ggsave('../plots/clostridioides_relAbun_plot.pdf',
       plot = clostridioides_only_plot,
       width = 15,
       height = 5.5)
ggsave('../plots/enterococcus_relAbun_plot.pdf',
       plot = enterococcus_only_plot,
       width = 15,
       height = 5.5)
ggsave('../plots/escherichiaShigella_relAbun_plot.pdf',
       plot = `escherichia-shigella_only_plot`,
       width = 15,
       height = 5.5)
ggsave('../plots/lactobacillus_relAbun_plot.pdf',
       plot = lactobacillus_only_plot,
       width = 15,
       height = 5.5)
```

**png versions of plots** 
```{r}
ggsave('../plots/pngs/genus_relAbun_plot.png',
       plot = genusAbun_plot,
       width = 17,
       height = 13)
ggsave('../plots/pngs/enterobacteriaceae_relAbun_plot.png',
       plot = entero_genusAbun_plot,
       width = 17,
       height = 7)
ggsave('../plots/pngs/enterococcusxVRE_relAbun_plot.png',
       plot = fig3b_plot,
       width = 15.5,
       height = 5.5)

## individual genus rel abun plots
ggsave('../plots/pngs/bacteroides_relAbun_plot.png',
       plot = bacteroides_only_plot,
       width = 15,
       height = 5.5)
ggsave('../plots/pngs/bifidobacterium_relAbun_plot.png',
       plot = bifidobacterium_only_plot,
       width = 15,
       height = 5.5)
ggsave('../plots/pngs/bifido_notLog10_relAbun_plot.png',
       plot = bifidobacterium_notLog10_plot,
       width = 15,
       height = 5)
ggsave('../plots/pngs/blautia_relAbun_plot.png',
       plot = blautia_only_plot,
       width = 15,
       height = 5.5)
ggsave('../plots/pngs/blautia_notLog10_relAbun_plot.png',
       plot = blautia_notLog10_plot,
       width = 15,
       height = 5)
ggsave('../plots/pngs/clostridioides_relAbun_plot.png',
       plot = clostridioides_only_plot,
       width = 15,
       height = 5.5)
ggsave('../plots/pngs/enterococcus_relAbun_plot.png',
       plot = enterococcus_only_plot,
       width = 15,
       height = 5.5)
ggsave('../plots/pngs/escherichiaShigella_relAbun_plot.png',
       plot = `escherichia-shigella_only_plot`,
       width = 15,
       height = 5.5)
ggsave('../plots/pngs/lactobacillus_relAbun_plot.png',
       plot = lactobacillus_only_plot,
       width = 15,
       height = 5.5)
```

**stats**
```{r}
## stats
write_tsv(all_genus_stats,
          '../stats/perGenus_treatGroup_postHoc.tsv')
write_tsv(genusAbun_strat_res,
          '../stats/genusAbun_treatGroup_stratOnly.tsv')
write_tsv(genusAbun_strat_contrast,
          '../stats/genusAbun_treatGroup_stratOnly_postHoc.tsv')
write_tsv(genusAbun_doubleStrat_res_wValues,
          '../stats/genusAbun_treatGroup_visit_strat.tsv')
write_tsv(genusAbun_treatGroup_strat_diffs,
          '../stats/genusAbun_treatGroup_visitStrat_postHoc.tsv')
write_tsv(proc_genusAbun_doubleStrat_contrast_wValues,
          '../stats/signifOnly_genusAbun_treatGroup_visitStrat_postHoc.tsv')
write_tsv(buty_treatGroup_strat_diffs,
          '../stats/buty_familyAbun_treatGroup_visitStrat_postHoc.tsv')
```

